---
title: "Cell line barcode analyses"
format: html
toc-depth: 4
---

## Preliminaries.

Raw barcode count data are listed in Supplementary Tables S2, under the sheet names:

* Barcodes_libr_&_Parental_Palbo (Parental_3_passage_3_Count)
* Barcodes_Parental_Abema
* Barcodes_Palbo_study
* Barcodes_Abema_study
* Barcodes_clones
* Barcodes_tumors_&_recurrences

In the below code, we read in barcode counts and metadata from tables that compile this same data together into a long-format. 

### Load in libraries.

```{r}
library(tidyverse)
library(pheatmap)
library(viridis)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
library(philentropy)
```

### Create output data, plot folders.

```{r}
dir.create("data/raw/01-barcoding", recursive = TRUE) # input files
dir.create("data/processed/01-barcoding", recursive = TRUE) # output data, plots
```

### Sample structure of formatted barcode count, metadata tables.

```{r}
print(str(readr::read_tsv(
  "data/raw/01-barcoding/bct-counts_abema-gradual.tsv"
)))
print(str(readr::read_tsv(
  "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv"
)))

print(str(readr::read_tsv(
  "data/raw/01-barcoding/bct-counts_abema-gradual.tsv"
)))
print(str(readr::read_tsv(
  "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv"
)))

print(str(readr::read_tsv(
  "data/raw/03-in-vivo-barcoding/bct-counts_mice-met-18.tsv"
)))
print(str(readr::read_tsv(
  "data/raw/03-in-vivo-barcoding/bct-metadata_mice-met-18.tsv"
)))
```

### Load in plot color palettes.

```{r}
pal_orig.ident <- c(
  "Mut_AbemaR" = "orange",
  "WT_AbemaR" = "red",
  "Mut_PalboR" = "lightblue",
  "WT_PalboR" = "royalblue",
  "WT_PalboR_Clone" = "purple",
  "Mut_Vehicle" = "gray85",
  "WT_Vehicle" = "gray30",
  "Parental" = "burlywood4"
)

pal_treatment <- c(
  "Abema" = "red",
  "Palbo" = "royalblue",
  "Vehicle" = "orange",
  "Parental" = "burlywood4"
)

pal_esr1_status = c(
  "WT" = "gray30",
  "Y537S" = "gray85",
  "Parental" = "burlywood4"
)

pal_kl_graph_name = c(
  "WT Vehicle CTR of PalboR" = "gray30",
  "Mut Vehicle CTR of PalboR" = "gray85",
  "WT PalboR" = "royalblue",
  "Mut PalboR" = "lightblue",

  "WT Vehicle CTR of AbemaR" = "gray30",
  "Mut Vehicle CTR of AbemaR" = "gray85",
  "WT AbemaR" = "red",
  "Mut AbemaR" = "orange"
)

pal_js_graph_name = c(
  "WT Vehicle" = "gray30",
  "Mut Vehicle" = "gray85",
  "WT Palbo" = "royalblue",
  "Mut Palbo" = "lightblue",
  "WT Abema" = "red",
  "Mut Abema" = "orange"
)
```

## Load in function to plot clustered heatmap of barcoding expression, sample and barcode clusterings.

```{r}
#' Plot a heatmap of barcode counts above a threshold for a given set of samples.
#'
#' Cluster rows and columns using given feature and sample distance matrices.
#'
#' @param infile_counts Path to (long) barcode count matrix.
#' @param infile_feature_dist Path to RDS of feature distances.
#' @param infile_sample_dist Path to RDS of sample distances.
#' @param infile_meta Path to sample metadata.
#' @param dist_type Distance metric used to generate the feature dist and
#' sample dist matrices.
#' @param .condition Vector of Conditions to include in plot. Some of
#' "Parental", "ND", "DOX".
#' @param .treatment Vector of Treatments to include in plot. Some of "DMSO",
#' "Parental", "Palbo", "Abema".
#' Needed as pheatmap does not drop unused factor levels from heatmap legends.
#' One of "Palbo", "Abema", "Palbo Abema Middle", "Palbo Abema",
#' "Palbo Abema Parental".
#' @param freq_thresh Minimum frequency threshold (0-1) to filter out low-frequency barcodes.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param title Plot title.
#' @param lower_thresh Lower threshold to determine color breaks in pheatmap.
#' @param data_dir Processed data directory.
#' @param output_dir Processed output directory.
#'
#' @returns Pheatmap object of plot.
plot_bct_counts_heatmap <- function(
  infile_counts,
  infile_feature_dist,
  infile_sample_dist,
  infile_meta,
  dist_type,
  .condition = c("Parental", "ND", "DOX"),
  .treatment = c("Parental", "DMSO", "Abema", "Palbo"),
  include_clone = FALSE,
  clone_sample_names = NULL,
  freq_thresh = 0.001,
  width = 7,
  height = 10,
  title,
  lower_thresh = -5,
  data_dir = "data/processed/01-barcoding",
  output_dir = "data/processed/01-barcoding"
) {
  counts <- read_tsv(infile_counts, show_col_types = FALSE)
  feature_dist <- readRDS(infile_feature_dist)
  sample_dist <- readRDS(infile_sample_dist)
  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    mutate(
      graph_name = str_c(
        forcats::fct_recode(
          factor(experiment),
          ag = "abema-gradual",
          pg = "palbo-gradual",
          pgc = "palbo-gradual-clones"
        ),
        forcats::fct_recode(factor(condition), WT = "ND", "Mut" = "DOX"),
        forcats::fct_recode(
          factor(treatment),
          AbemaR = "Abema",
          "PalboR" = "Palbo",
          Vehicle = "DMSO"
        ),
        timepoint,
        stringr::str_replace_na(replicate_id),
        sep = " "
      )
    )

  # Use actual sample names for clones to distinguish them properly from each other.
  if (include_clone) {
    meta <- purrr::map_dfr(
      infile_meta,
      readr::read_tsv,
      show_col_types = FALSE
    ) %>%
      mutate(
        treatment = case_when(
          (experiment == "palbo-gradual-clones" & treatment == "PDR") ~ "Palbo", # fix PDR -> Palbo for clones
          .default = treatment
        )
      ) |>
      filter(
        experiment != "palbo-gradual-clones" |
          sample_name %in% clone_sample_names
      ) |>
      mutate(
        graph_name = case_when(
          experiment == "palbo-gradual-clones" ~ gsub(
            "^[0-9]+_{1}",
            "",
            gsub("_CG.*", "", sample_name)
          ),
          .default = str_c(
            forcats::fct_recode(
              factor(experiment),
              ag = "abema-gradual",
              pg = "palbo-gradual",
              pgc = "palbo-gradual-clones"
            ),
            forcats::fct_recode(factor(condition), WT = "ND", "Mut" = "DOX"),
            forcats::fct_recode(
              factor(treatment),
              AbemaR = "Abema",
              "PalboR" = "Palbo",
              Vehicle = "DMSO"
            ),
            timepoint,
            stringr::str_replace_na(replicate_id),
            sep = " "
          )
        )
      )
  }

  file_metadata_dist <- gsub(
    ".*feature-dist_(.*)\\.rds",
    "\\1",
    infile_feature_dist
  )
  outfile <- file.path(
    output_dir,
    paste0("bct_frac_heatmap_", file_metadata_dist, ".pdf")
  )
  print(outfile)

  feature_clust <- hclust(feature_dist)
  sample_clust <- hclust(sample_dist)

  dat_nxp_counts <- counts %>%
    select(sample_name, barcode, count) %>%
    inner_join(meta, by = "sample_name") %>% # implicitly filters to just things that are in meta and the counts matrix
    mutate(
      graph_name = case_when(
        experiment == "palbo-gradual-clones" ~ gsub(
          "^[0-9]+_{1}",
          "",
          gsub("_CG.*", "", sample_name)
        ),
        .default = str_c(
          forcats::fct_recode(
            factor(experiment),
            ag = "abema-gradual",
            pg = "palbo-gradual",
            pgc = "palbo-gradual-clones"
          ),
          forcats::fct_recode(factor(condition), WT = "ND", "Mut" = "DOX"),
          forcats::fct_recode(
            factor(treatment),
            AbemaR = "Abema",
            "PalboR" = "Palbo",
            Vehicle = "DMSO"
          ),
          timepoint,
          stringr::str_replace_na(replicate_id),
          sep = " "
        )
      )
    ) |>
    pivot_wider(
      id_cols = barcode,
      names_from = graph_name,
      values_from = count,
      values_fill = 0
    )
  dat_nxp_counts <- dat_nxp_counts %>%
    column_to_rownames("barcode")
  print(colSums(dat_nxp_counts))

  dat_nxp <- counts %>%
    select(sample_name, barcode, fraction) %>%
    inner_join(meta, by = "sample_name") %>% # implicitly filters to just things that are in meta and the counts matrix
    mutate(
      graph_name = case_when(
        experiment == "palbo-gradual-clones" ~ gsub(
          "^[0-9]+_{1}",
          "",
          gsub("_CG.*", "", sample_name)
        ),
        .default = str_c(
          forcats::fct_recode(
            factor(experiment),
            ag = "abema-gradual",
            pg = "palbo-gradual",
            pgc = "palbo-gradual-clones"
          ),
          forcats::fct_recode(factor(condition), WT = "ND", "Mut" = "DOX"),
          forcats::fct_recode(
            factor(treatment),
            AbemaR = "Abema",
            "PalboR" = "Palbo",
            Vehicle = "DMSO"
          ),
          timepoint,
          stringr::str_replace_na(replicate_id),
          sep = " "
        )
      )
    ) |>
    pivot_wider(
      id_cols = barcode,
      names_from = graph_name,
      values_from = fraction,
      values_fill = 0
    )
  dat_nxp <- dat_nxp %>%
    column_to_rownames("barcode")

  col_anno <- meta %>%
    select(graph_name, condition, treatment, experiment) %>%
    filter(graph_name %in% names(dat_nxp)) %>%
    rename(Treatment = treatment, `ESR1 Status` = condition) %>%
    mutate(
      `ESR1 Status` = fct_recode(
        `ESR1 Status`,
        WT = "ND",
        Y537S = "DOX",
        Parental = "Parental"
      )
    ) %>%
    mutate(`Treatment` = fct_recode(`Treatment`, Vehicle = "DMSO")) %>%
    column_to_rownames("graph_name")

  anno_colors = list()
  # Build up legend element-by-element
  if ("Abema" %in% col_anno$Treatment) {
    anno_colors[["Treatment"]] = c(anno_colors[["Treatment"]], "Abema" = "red")
  }
  if ("Palbo" %in% col_anno$Treatment) {
    anno_colors[["Treatment"]] = c(
      anno_colors[["Treatment"]],
      "Palbo" = "royalblue"
    )
  }
  if ("Vehicle" %in% col_anno$Treatment) {
    anno_colors[["Treatment"]] = c(
      anno_colors[["Treatment"]],
      "Vehicle" = "gold"
    )
  }
  if ("Parental" %in% col_anno$Treatment) {
    anno_colors[["Treatment"]] = c(
      anno_colors[["Treatment"]],
      "Parental" = "burlywood4"
    )
  }

  if ("WT" %in% col_anno$`ESR1 Status`) {
    anno_colors[["ESR1 Status"]] = c(
      anno_colors[["ESR1 Status"]],
      "WT" = "gray30"
    )
  }
  if ("Y537S" %in% col_anno$`ESR1 Status`) {
    anno_colors[["ESR1 Status"]] = c(
      anno_colors[["ESR1 Status"]],
      "Y537S" = "gray85"
    )
  }
  if ("Parental" %in% col_anno$`ESR1 Status`) {
    anno_colors[["ESR1 Status"]] = c(
      anno_colors[["ESR1 Status"]],
      "Parental" = "burlywood4"
    )
  }

  if ("palbo-gradual-clones" %in% col_anno$experiment) {
    col_anno <- col_anno |>
      mutate(
        `Bulk/Clone` = case_when(
          experiment == "palbo-gradual-clones" ~ "Clone",
          .default = "Bulk"
        )
      ) |>
      mutate(
        `Bulk/Clone` = factor(`Bulk/Clone`, levels = c("Bulk", "Clone"))
      ) |>
      relocate(`Bulk/Clone`, .before = `ESR1 Status`)
    anno_colors[["Bulk/Clone"]] = c(
      anno_colors[["Bulk/Clone"]],
      "Bulk" = "burlywood4"
    )
    anno_colors[["Bulk/Clone"]] = c(
      anno_colors[["Bulk/Clone"]],
      "Clone" = "burlywood1"
    )
  }

  # reorder treatment levels
  col_anno <- col_anno |>
    mutate(
      Treatment = forcats::fct_relevel(
        Treatment,
        "Abema",
        "Palbo",
        "Vehicle",
        "Parental"
      )
    )

  # Remove experiment annotation
  col_anno <- col_anno |> select(-experiment)

  # Make sure we're keeping each sample separate, just swapping out some names for
  # more readable version in the graph.
  stopifnot(
    length(unique(rownames(col_anno))) == dim(dat_nxp)[[2]],
    dim(dat_nxp)[[2]] == length(unique(counts$sample_name))
  )

  # Remove any zero-valued counts before logging and plotting (avoids color
  # scale getting altered by log(ct+pseudocount))
  set.seed(238038)
  print(pheatmap::pheatmap(
    log10(
      ((dat_nxp |>
        mutate(across(everything(), \(x) {
          na_if(x, 0)
        }))))
    ),
    color = rev(viridis::magma(150)),
    breaks = seq(lower_thresh, 0, length.out = 151),
    cluster_cols = sample_clust,
    cluster_rows = feature_clust,
    annotation_col = col_anno,
    annotation_colors = anno_colors,
    show_rownames = FALSE,
    show_colnames = TRUE,
    border_color = NA,
    width = width,
    height = height,
    main = title,
    angle_col = 90,
    drop_levels = TRUE,
    na_col = "gray"
  ))
}
```

## Load in function to plot barcode fishplots.

```{r}
#' Plot a fishplot of barcode fractions above a threshold and calculate
#' the normalized KL divergence against a discrete uniform distribution
#' for the End timepoint for each bioreplicate for the 8 clusters.
#'
#' This function gets run indirectly when [get_sample_feature_dist_bct_counts()]
#' is called with parameter plot_fishplot = TRUE. We recommend running in this
#' indirect way to ensure input files are generated correctly with distance
#' matrices.
#'
#' @param infile_full_counts Path to table of unfiltered barcode counts.
#' @param infile_counts Path to table of filtered barcode counts from [get_sample_feature_dist_bct_counts()].
#' @param infile_feature_dist Path to distance matrix of barcodes from [get_sample_feature_dist_bct_counts()].
#' @param infile_sample_dist Path to distance matrix of samples from [get_sample_feature_dist_bct_counts()].
#' @param infile_meta Path to sample metadata.
#' @param nclust Number of clusters to cut in fishplot.
#' @param dist_type Distance metric used to generate the feature dist and
#' sample dist matrices.
#' @param width = Plot width (in).
#' @param height Plot height (in).
#' @param title Plot title.
#' @param subtitle Plot subtitle.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_bct_frac_fishplot <- function(
  infile_full_counts,
  infile_counts,
  infile_feature_dist,
  infile_sample_dist,
  infile_meta,
  nclust,
  dist_type,
  width = 6,
  height = 4,
  title,
  subtitle,
  ddir = "data/processed/01-barcoding",
  odir = "data/processed/01-barcoding"
) {
  if (!dir.exists(file.path(ddir, "fishplots", dist_type))) {
    dir.create(file.path(ddir, "fishplots", dist_type), recursive = TRUE)
  }
  if (!dir.exists(file.path(odir, "fishplots", dist_type))) {
    dir.create(file.path(odir, "fishplots", dist_type), recursive = TRUE)
  }

  full_counts <- read_tsv(infile_full_counts, show_col_types = FALSE)
  counts <- read_tsv(infile_counts, show_col_types = FALSE)
  feature_dist <- readRDS(infile_feature_dist)
  sample_dist <- readRDS(infile_sample_dist)
  meta <- read_tsv(infile_meta) %>%
    mutate(
      condition = factor(condition, levels = c("Parental", "ND", "DOX")),
      treatment = factor(
        treatment,
        levels = c("Parental", "DMSO", "Abema", "Palbo")
      ),
      dose = factor(dose, levels = c("/", "50nM", "200nM", "1uM")),
      timepoint = factor(
        timepoint,
        levels = c("Parental", "Early", "Middle", "End")
      ),
      replicate_id = factor(replicate_id, levels = c("1", "2", "3"))
    )

  # Order graphnames
  ml <- meta %>%
    arrange(condition, treatment, timepoint, dose, replicate_id) %>%
    mutate(
      graph_name = str_c(
        condition,
        treatment,
        dose,
        timepoint,
        replicate_id,
        sep = "_"
      )
    ) %>%
    pull(graph_name)

  # Order filenames
  gl <- meta %>%
    arrange(condition, treatment, timepoint, dose, replicate_id) %>%
    pull(sample_name)

  # Generate output filenames
  file_metadata_dist <- gsub(
    ".*feature-dist_(.*)\\.rds",
    "\\1",
    infile_feature_dist
  )

  # Loop through replicates to make a organize fishplot data for (Parent + (each biorepl))
  # Note, the same parental sample is used for all bioreplicates.
  counts_mat_repl_all_replicates <- purrr::map(1:3, \(i) {
    counts_repl <- counts %>%
      filter(treatment == "Parental" | replicate_id == i) %>%
      filter(!is.na(timepoint)) %>%
      mutate(
        graph_name = str_c(
          condition,
          treatment,
          dose,
          timepoint,
          replicate_id,
          sep = "_"
        )
      ) %>%
      mutate(graph_name = factor(graph_name, levels = ml))

    counts_repl <- counts_repl %>%
      left_join(full_counts, by = c("sample_name", "barcode")) %>%
      mutate(
        fraction = case_when(
          count.x == 0 ~ 0, # need to account for 0 count barcodes
          TRUE ~ fraction.x
        )
      ) |>
      mutate(predicted_cluster = factor(predicted_cluster))

    # Number of unique barcodes per cluster
    cluster_cts <- counts_repl |>
      group_by(predicted_cluster, .drop = FALSE) |>
      mutate(
        predicted_cluster_n = sprintf(
          "Cluster %s (N=%i)",
          predicted_cluster,
          length(unique(barcode))
        )
      ) |>
      distinct(predicted_cluster, predicted_cluster_n)

    counts_mat_repl <- counts_repl |>
      group_by(
        sample_name,
        condition,
        treatment,
        dose,
        timepoint,
        replicate_id,
        predicted_cluster,
        .drop = FALSE
      ) %>%
      summarize(fraction = sum(fraction)) |>
      left_join(cluster_cts, by = "predicted_cluster") |>
      dplyr::rename(`Cluster` = predicted_cluster_n) |>
      dplyr::mutate(
        timepoint = factor(
          timepoint,
          levels = c("Parental", "Early", "Middle", "End")
        )
      )

    cluster_cts <- counts_repl |>
      dplyr::group_by(predicted_cluster, .drop = FALSE) |>
      dplyr::mutate(
        predicted_cluster_n = sprintf(
          "%s (N=%i)",
          predicted_cluster,
          length(unique(barcode))
        )
      ) |>
      dplyr::distinct(predicted_cluster, predicted_cluster_n)

    counts_mat_repl <- counts_repl |>
      dplyr::group_by(
        sample_name,
        condition,
        treatment,
        dose,
        timepoint,
        replicate_id,
        predicted_cluster,
        .drop = FALSE
      ) %>%
      dplyr::summarize(fraction = sum(fraction)) |>
      dplyr::left_join(cluster_cts, by = "predicted_cluster") |>
      dplyr::rename(`Cluster` = predicted_cluster_n) |>
      dplyr::mutate(
        timepoint = factor(
          timepoint,
          levels = c("Parental", "Early", "Middle", "End")
        )
      )

    counts_repl_bioreplicate_id <- counts_repl |>
      dplyr::left_join(cluster_cts, by = "predicted_cluster") |>
      mutate(plot_bioreplicate_id = sprintf("Bioreplicate %i", i))
    return(counts_repl_bioreplicate_id)
  }) |>
    purrr::list_rbind()

  # Format data for plotting.
  counts_mat_repl_all_replicates2 <- counts_mat_repl_all_replicates |>
    dplyr::group_by(
      sample_name,
      condition,
      treatment,
      dose,
      timepoint,
      replicate_id,
      predicted_cluster_n,
      plot_bioreplicate_id,
      .drop = FALSE
    ) %>%
    dplyr::summarize(fraction = sum(fraction)) |>
    dplyr::rename(`Cluster` = "predicted_cluster_n") |>
    dplyr::mutate(
      timepoint = factor(
        timepoint,
        levels = c("Parental", "Early", "Middle", "End")
      )
    )

  # For each (timpoint, Cluster, plot_bioreplicate_id) combination that exists,
  #' impute a zero barcode fraction where that barcode is not seen
  #' (artifact of preprocessing).
  counts_mat_repl_all_replicates2 <- counts_mat_repl_all_replicates2 |>
    ungroup() |>
    tidyr::complete(
      timepoint,
      Cluster,
      plot_bioreplicate_id,
      fill = list(fraction = 0)
    )

  plt <- counts_mat_repl_all_replicates2 |>
    ggplot2::ggplot(aes(
      x = timepoint,
      y = fraction,
      fill = Cluster,
      group = Cluster
    )) +
    ggplot2::geom_area(stat = "identity") +
    ggplot2::labs(title = title, y = "Fraction", x = NULL) +
    ggplot2::lims(y = c(0, 1)) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      text = element_text(size = 8),
      title = element_text(size = 6),
      strip.text = element_text(size = 6),
      legend.key.size = unit(0.3, "cm"),
      legend.title = element_text(size = 6),
      legend.text = element_text(size = 5),
      legend.margin = margin(t = -20, unit = "pt"),
      panel.grid.major = element_line(linewidth = 0.3),
      panel.grid.minor = element_line(linewidth = .2)
    ) +
    ggplot2::scale_x_discrete(expand = expansion(add = .2)) +
    ggplot2::facet_wrap(~plot_bioreplicate_id, nrow = 1) +
    scale_fill_manual(
      values = rev(c(
        "#AFDCE3",
        "#E5C494",
        "#2B388F",
        "#96B53C",
        "#EC1C24",
        "#1B75BB",
        "#F6881E",
        "#259664"
      ))
    )

  outfile <- file.path(
    odir,
    "fishplots",
    dist_type,
    sprintf("fishplot_%s.pdf", file_metadata_dist)
  )
  ggsave(outfile, plt, width = 17, height = 4, units = "cm")

  # Write out the normalized KL divergence against a discrete uniform distribution
  #' for the End timepoint for each bioreplicate for the 8 clusters.
  shannon_results <- counts_mat_repl_all_replicates2 |>
    dplyr::group_by(plot_bioreplicate_id, timepoint) |>
    dplyr::mutate(fraction_rescaled = fraction / sum(fraction)) |>
    dplyr::summarize(
      n_unique_barcodes = dplyr::n(),
      normalized_kl_divergence = philentropy::KL(
        rbind(
          fraction_rescaled,
          (rep(1, length(fraction_rescaled)) / length(fraction_rescaled))
        ),
        unit = "log"
      ) /
        log(length(fraction_rescaled)),
      n_unique_barcodes_check = length(fraction_rescaled)
    ) |>
    dplyr::mutate(metadata = file_metadata_dist)
  saveRDS(
    shannon_results,
    file.path(
      ddir,
      "fishplots",
      dist_type,
      sprintf(
        "shannon-kl-divergence-fishplot-alltimepoints-%s.rds",
        file_metadata_dist
      )
    )
  )

  return(plt)
}
```

## Load function to cluster barcodes and samples.

Note, this function indirectly calls functions to plot heatmaps and fishplots of barcodes.

```{r}
#' Cluster barcodes by their counts according to the given distance metric and
#' write files of distance matrices, corresponding count matrices, and optionally
#' plots a heatmap, fishplot of the log barcode fractions.
#'
#' Only includes barcodes that are above a specified frequency threshold at
#' least once in the dataset.
#'
#' @param infile_meta Filepath to tsv of barcode metadata.
#' @param infile_counts Filepath to tsv of barcode counts data.
#' @param freq_thresh Frequency threshold used to cluster all barcodes with
#' maximum frequency (across samples) less than freq_thresh to their  own
#' separate cluster.
#' @param nclust Number of clusters (besides 0) to cut hclust at.
#' @param .condition Condition to filter to. One of c("DOX", "ND", "Parental").
#' @param .treatment Treatment to filter to. One of c("Parental", "DMSO", "Abema", "Palbo")
#' @param .Timepoint Timepoint to filter to. One of c("Parental", "Early", "Middle", "End")
#' @param dist_metric Distance metric/ preprocessing done on counts matrix to cluster.
#' @param include_palbo_gradual_dmso_only Include Palbo-gradual DMSO only when
#' @param plot_heatmap Plot barcode-sample heatmap?
#' @param plot_fishplot Plot barcode fishplot?
#' @param plt_title Plot title.
#' @param include_clone Logical. Include the WT Palbo clone if the clones infiles are given?
#' @param clone_sample_names Sample names of clones to include in plot.
#' @param clone_suffix Suffix to include in output filepaths to identify which clone samples were included.
#' @param exclude_abema_vehicle Logical. Exclude Abema experiment Vehicle samples?
#' @param outfile_counts Path to write long tibble of barcode counts and fractions. Automatically generated if NULL.
#' @param outfile_counts_nxp Path to write nxp matrix of raw barcode counts. Automatically generated if NULL.
#' @param outfile_sample_dist Path to write sample clustering information. Automatically generated if NULL.
#' @param outfile_feature_dist Path to write feature (barcode) clustering information. Automatically generated if NULL.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
get_sample_feature_dist_bct_counts <- function(
  infile_meta,
  infile_counts,
  freq_thresh = 0.001,
  nclust = 8,
  .condition,
  .treatment,
  .Timepoint = c("Parental", "Early", "Middle", "End"),
  dist_metric,
  include_palbo_gradual_dmso_only = FALSE,
  plot_heatmap = FALSE,
  plot_fishplot = FALSE,
  plt_title = "",
  include_clone = FALSE,
  clone_sample_names = NULL,
  clone_suffix = NULL,
  exclude_abema_vehicle = TRUE,
  outfile_counts = NULL,
  outfile_counts_nxp = NULL,
  outfile_sample_dist = NULL,
  outfile_feature_dist = NULL,
  ddir = "data/processed/01-barcoding",
  odir = "data/processed/01-barcoding"
) {
  dist_metric <- match.arg(
    dist_metric,
    c("column-swept-ppr", "log-fraction-euclidean")
  )

  # Generate output filenames (avoids mistakes with manually specifying filenames)
  experiment <- paste0(
    gsub(".*bct-metadata_(.*)\\.tsv", "\\1", infile_meta),
    collapse = "_"
  )
  filename_metadata <- sprintf(
    "%s_%s_%s_%s_thresh%s",
    paste0(experiment, collapse = "-"),
    paste0(.condition, collapse = "-"),
    paste0(.treatment, collapse = "-"),
    paste0(.Timepoint, collapse = "-"),
    gsub("\\.", "p", as.character(freq_thresh))
  )
  file_metadata_dist <- sprintf(
    "%s_%s_%s_%s_%s_thresh%s",
    paste0(dist_metric, collapse = "-"),
    paste0(experiment, collapse = "-"),
    paste0(.condition, collapse = "-"),
    paste0(.treatment, collapse = "-"),
    paste0(.Timepoint, collapse = "-"),
    gsub("\\.", "p", as.character(freq_thresh))
  )

  # Automatially generate output filenames if not specified.
  if (is.null(outfile_counts)) {
    outfile_counts <- file.path(
      ddir,
      paste0("feature-counts_", file_metadata_dist, ".tsv")
    )
  }
  if (is.null(outfile_counts_nxp)) {
    outfile_counts_nxp <- file.path(
      ddir,
      paste0("feature-counts-nxp_", file_metadata_dist, ".tsv")
    )
  }
  if (is.null(outfile_sample_dist)) {
    outfile_sample_dist <- file.path(
      ddir,
      paste0("sample-dist_", file_metadata_dist, ".rds")
    )
  }
  if (is.null(outfile_feature_dist)) {
    outfile_feature_dist <- file.path(
      ddir,
      paste0("feature-dist_", file_metadata_dist, ".rds")
    )
  }

  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    filter(condition %in% .condition, treatment %in% .treatment) %>%
    mutate(
      condition = factor(condition, levels = c("Parental", "ND", "DOX")),
      treatment = factor(
        treatment,
        levels = c("Parental", "DMSO", "Abema", "Palbo")
      ),
      dose = factor(dose, levels = c("/", "50nM", "200nM", "1uM")),
      timepoint = factor(
        timepoint,
        levels = c("Parental", "Early", "Middle", "End")
      ),
      replicate_id = factor(replicate_id, levels = c("1", "2", "3"))
    ) %>%
    mutate(
      graph_name = str_c(
        condition,
        treatment,
        dose,
        timepoint,
        replicate_id,
        sep = "_"
      )
    )

  # just want to include the WT Palbo Clone (3_15) (same sample as we have scrnaseq on)
  if (include_clone) {
    meta <- purrr::map_dfr(
      infile_meta,
      readr::read_tsv,
      show_col_types = FALSE
    ) %>%
      mutate(
        treatment = case_when(
          (experiment == "palbo-gradual-clones" & treatment == "PDR") ~ "Palbo", # fix PDR -> Palbo for clones
          .default = treatment
        )
      ) |>
      filter(
        experiment != "palbo-gradual-clones" |
          sample_name %in% clone_sample_names
      ) |>
      filter(condition %in% .condition, treatment %in% .treatment) %>%
      mutate(
        condition = factor(condition, levels = c("Parental", "ND", "DOX")),
        treatment = factor(
          treatment,
          levels = c("Parental", "DMSO", "Abema", "Palbo")
        ),
        dose = factor(dose, levels = c("/", "50nM", "200nM", "1uM")),
        timepoint = factor(
          timepoint,
          levels = c("Parental", "Early", "Middle", "End")
        ),
        replicate_id = factor(replicate_id, levels = c("1", "2", "3"))
      ) %>%
      mutate(
        graph_name = str_c(
          condition,
          treatment,
          dose,
          timepoint,
          stringr::str_replace_na(replicate_id),
          sep = "_"
        )
      )

    # Automatially generate output filenames if not specified with a suffix for what clones are included.
    outfile_counts <- file.path(
      ddir,
      sprintf("feature-counts_%s_%s.tsv", filename_metadata, clone_suffix)
    )
    outfile_counts_nxp <- file.path(
      ddir,
      sprintf("feature-counts-nxp_%s_%s.tsv", filename_metadata, clone_suffix)
    )
    outfile_sample_dist <- file.path(
      ddir,
      sprintf("sample-dist_%s_%s.rds", file_metadata_dist, clone_suffix)
    )
    outfile_feature_dist <- file.path(
      ddir,
      sprintf("feature-dist_%s_%s.rds", file_metadata_dist, clone_suffix)
    )
  }

  counts <- purrr::map_dfr(
    infile_counts,
    readr::read_tsv,
    show_col_types = FALSE
  )

  counts_meta <- counts %>%
    left_join(meta, by = "sample_name") %>%
    filter(
      condition %in% .condition,
      treatment %in% .treatment,
      timepoint %in% .Timepoint
    ) %>%
    arrange(barcode, treatment, timepoint, replicate_id) %>%
    select(-fraction) %>% # drop existing fraction and recalculate
    group_by(sample_name) %>%
    mutate(fraction = count / sum(count)) %>% # recalculate fraction per sample -- this is just dividing the count matrix by a scale factor (colsums) row vector (separate for each sample)
    ungroup() %>%
    select(-filename) %>%
    relocate(fraction, .after = count)

  # If specified take Palbo-gradual experiment Vehicle treated samples, not Palbo treated
  if (include_palbo_gradual_dmso_only) {
    counts_meta <- counts_meta |>
      filter(!(experiment == "palbo-gradual" & treatment == "Palbo"))
  }

  # If Palbo, Abema, DMSO present, only take the DMSO from the Palbo sample
  if (
    all(.treatment %in% c("Palbo", "Abema", "DMSO")) &
      (length(unique(.treatment)) == 3) &
      exclude_abema_vehicle
  ) {
    counts_meta <- counts_meta %>%
      filter(!(experiment == "abema-gradual" & treatment == "DMSO"))
  }

  # If Palbo, Abema, DMSO, Parental present, only take the DMSO and Parental from the Palbo sample
  if (
    all(.treatment %in% c("Palbo", "Abema", "DMSO", "Parental")) &
      length(unique(.treatment)) == 4
  ) {
    counts_meta <- counts_meta %>%
      filter(!(experiment == "abema-gradual" & treatment == "DMSO")) %>%
      filter(!(experiment == "abema-gradual" & treatment == "Parental"))
  }

  counts_meta_geq_thresh <- counts_meta %>%
    group_by(barcode) %>% # Filter out any barcodes that are below a certain threshold across all observations
    filter(max(fraction) >= freq_thresh) %>%
    ungroup()

  counts_meta_geq_thresh_nxp <- counts_meta_geq_thresh %>%
    select(sample_name, barcode, count) %>%
    pivot_wider(
      id_cols = barcode,
      names_from = sample_name,
      values_from = count,
      values_fill = 0
    )

  # Write out both long and pivoted (version that gets dist() called on it) to file.
  write_tsv(counts_meta_geq_thresh_nxp, file = outfile_counts_nxp)

  counts_meta_geq_thresh_nxp <- counts_meta_geq_thresh_nxp %>%
    column_to_rownames("barcode")

  # Barcode Fraction clustering.
  if (dist_metric == "log-fraction-euclidean") {
    counts_meta_geq_thresh_nxp <- counts_meta_geq_thresh %>%
      select(sample_name, barcode, fraction) %>%
      pivot_wider(
        id_cols = barcode,
        names_from = sample_name,
        values_from = fraction,
        values_fill = 0
      )
    counts_meta_geq_thresh_nxp <- counts_meta_geq_thresh_nxp %>%
      column_to_rownames("barcode")
    feature_dist <- dist(log(counts_meta_geq_thresh_nxp + 1e-12))
    sample_dist <- dist(t(log(counts_meta_geq_thresh_nxp + 1e-12)))
  }

  # Divide out the column sums to get all samples on the same scale, then
  # calculate genewise poisson pearson residuals, then take Euclidean
  # distances on these.
  if (dist_metric == "column-swept-ppr") {
    #' p = (c - mu)/sqrt(mu)
    counts_meta_geq_thresh_nxp_colswept <- sweep(
      counts_meta_geq_thresh_nxp,
      2,
      Matrix::colSums(counts_meta_geq_thresh_nxp),
      `/`
    )
    pearson_residualized_counts = (counts_meta_geq_thresh_nxp_colswept -
      Matrix::rowMeans(counts_meta_geq_thresh_nxp_colswept)) /
      sqrt(Matrix::rowMeans(counts_meta_geq_thresh_nxp_colswept))
    feature_dist <- dist(pearson_residualized_counts)
    sample_dist <- dist(t(pearson_residualized_counts))
  }

  # Append a column of cluster assignment for each barcode.
  feature_clust <- hclust(feature_dist)
  clusters <- cutree(feature_clust, nclust)
  clusters_tibble <- tibble(
    barcode = names(clusters),
    predicted_cluster = clusters
  )
  counts_meta_geq_thresh <- counts_meta_geq_thresh %>%
    left_join(clusters_tibble, by = "barcode")

  write_tsv(counts_meta_geq_thresh, file = outfile_counts) # contains barcode cluster assignments
  readr::write_rds(feature_dist, outfile_feature_dist)
  readr::write_rds(sample_dist, outfile_sample_dist)

  if (plot_heatmap) {
    plot_bct_counts_heatmap(
      infile_counts = outfile_counts,
      infile_feature_dist = outfile_feature_dist,
      infile_sample_dist = outfile_sample_dist,
      infile_meta = infile_meta,
      dist_type = dist_metric,
      include_clone = include_clone,
      clone_sample_names = clone_sample_names,
      title = plt_title
    )
  }

  if (plot_fishplot) {
    plot_bct_frac_fishplot(
      infile_full_counts = infile_counts,
      infile_counts = outfile_counts,
      infile_feature_dist = outfile_feature_dist,
      infile_sample_dist = outfile_sample_dist,
      infile_meta = infile_meta,
      nclust = nclust,
      dist_type = dist_metric,
      title = plt_title
    )
  }
}
```

### Cluster and plot heatmap of barcodes for Palbo-resistant, Vehicle Endpoint samples (Figure 2C).

```{r}
#| fig-width: 6
#| fig-height: 8
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv",
  freq_thresh = 0.01,
  .condition = c("ND", "DOX"),
  .treatment = c("Palbo", "DMSO"),
  .Timepoint = "End",
  include_clone = FALSE,
  dist_metric = "log-fraction-euclidean",
  plot_heatmap = TRUE,
  plt_title = sprintf("End: PalboR, Vehicle; %s thresh", 0.01)
)
```

### Cluster and plot heatmap of barcodes for Abema-resistant, Vehicle Endpoint samples (Figure 2D).

```{r}
#| fig-width: 6
#| fig-height: 8
get_sample_feature_dist_bct_counts(
  infile_meta = list("data/raw/01-barcoding/bct-metadata_abema-gradual.tsv"),
  infile_counts = list("data/raw/01-barcoding/bct-counts_abema-gradual.tsv"),
  freq_thresh = 0.01,
  .condition = c("ND", "DOX"),
  .treatment = c("Abema", "DMSO"),
  .Timepoint = "End",
  include_clone = FALSE,
  dist_metric = "log-fraction-euclidean",
  plot_heatmap = TRUE,
  plt_title = sprintf("End: AbemaR, Vehicle; %s thresh", 0.01)
)
```

### Cluster and plot heatmap of barcodes for Abema-resistant, Palbo-resistant, and Vehicle Endpoint samples (Figure 4B).

This function call includes Vehicle samples from both Abema and Palbo experiments.

```{r}
#| fig-width: 6
#| fig-height: 8
get_sample_feature_dist_bct_counts(
  infile_meta = list(
    "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
    "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv"
  ),
  infile_counts = list(
    "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
    "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv"
  ),
  freq_thresh = 0.001,
  .condition = c("ND", "DOX"),
  .treatment = c("Palbo", "Abema", "DMSO"),
  .Timepoint = "End",
  include_clone = FALSE,
  dist_metric = "log-fraction-euclidean",
  plot_heatmap = TRUE,
  exclude_abema_vehicle = FALSE,
  plt_title = sprintf("End: PalboR, AbemaR, Vehicle; %s thresh", 0.001)
)
```

## Load in function to plot KL divergences of sample barcode distributions.

```{r}
#' Plot normalized KL divergences of barcode fraction distributions against a
#' discrete uniform reference for the specified cell line samples.
#'
#' Uses normalized natural log KL divergences as D_{KL}(P||Q) / log(|P|) where P
#' is the sample barcode fraction distribution and Q is the discrete uniform
#' distribution over the same number of total barcodes using philentropy::KL,
#' and |P| is the total number of discrete atoms in the discrete distributions P, Q.
#'
#' Note, normalized KL divergences are used here instead of Shannon entropies.
#'
#' @param infile_counts Path to barcode counts.
#' @param infile_meta Path to barcode metadata.
#' @param plt_var_x Variable to run ANOVA/ttests on and plot on x axis.
#' @param plt_t_test_var1 Numerator variable of t test. If left NULL, keeps
#' the overall ANOVA F test among all plotted groups and all pairwise t-tests.
#' @param plt_t_test_var2 Denominator variable of t test. If left NULL, keeps
#' the overall ANOVA F test among all plotted groups and all pairwise t-tests.
#' @param .condition Vector of conditions to filter to.
#' @param .treatment Vector of treatments to filter to.
#' @param .timepoint Vector of timepionts to filter to.
#' @param include_palbo_gradual_dmso_only Include only the DMSO Vehicle sample
#' from palbo gradual?
#' @param padj.signif.stars Plot significance stars for ajusted pvalues of
#' pairwise comparisons?
#' @param freq_thresh Barcode frequency threshold to filter out any barcode
#' that fails to meet threshold over any timepoint for a sample.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param data_dir Processed data directory.
#' @param output_dir Processed output directory.
plot_barcode_fraction_KL_divergences_against_discrete_uniform_cell_line <- function(
  infile_counts,
  infile_meta,
  plt_var_x,
  plt_t_test_var1 = NULL,
  plt_t_test_var2 = NULL,
  .condition = c("Parental", "ND", "DOX"),
  .treatment = c("Parental", "DMSO", "Abema", "Palbo"),
  .timepoint = c("Parental", "Early", "Middle", "End"),
  include_palbo_gradual_dmso_only = FALSE,
  padj.signif.stars = TRUE,
  freq_thresh = NULL,
  width = 5,
  height = 5,
  data_dir = "data/processed/01-barcoding",
  output_dir = "data/processed/01-barcoding"
) {
  counts <- purrr::map_dfr(
    infile_counts,
    readr::read_tsv,
    show_col_types = FALSE
  )
  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    dplyr::mutate(
      condition = factor(condition, levels = c("Parental", "ND", "DOX")),
      treatment = factor(
        treatment,
        levels = c("Parental", "DMSO", "Palbo", "Abema")
      ),
      dose = factor(dose, levels = c("/", "50nM", "200nM", "1uM")),
      timepoint = factor(
        timepoint,
        levels = c("Parental", "Early", "Middle", "End")
      ),
      replicate_id = factor(replicate_id, levels = c("1", "2", "3")),
      experiment = factor(
        experiment,
        levels = c("palbo-gradual", "abema-gradual")
      )
    ) %>%
    dplyr::mutate(
      graph_color = dplyr::case_when(
        treatment == "DMSO" & condition == "ND" ~ "DMSO-ND",
        treatment == "DMSO" & condition == "DOX" ~ "DMSO-DOX",
        treatment == "Palbo" & condition == "ND" ~ "Palbo-ND",
        treatment == "Palbo" & condition == "DOX" ~ "Palbo-DOX",
        treatment == "Abema" & condition == "ND" ~ "Abema-ND",
        treatment == "Abema" & condition == "DOX" ~ "Abema-DOX",
        .default = "Parental"
      )
    ) %>%
    dplyr::mutate(
      graph_color = factor(
        graph_color,
        levels = c(
          "Parental",
          "DMSO-ND",
          "DMSO-DOX",
          "Palbo-ND",
          "Palbo-DOX",
          "Abema-ND",
          "Abema-DOX"
        )
      )
    )

  col_pal <- list(
    "DMSO-ND" = RColorBrewer::brewer.pal(12, "Paired")[[7]],
    "DMSO-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[8]],
    "Palbo-ND" = RColorBrewer::brewer.pal(12, "Paired")[[1]],
    "Palbo-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[2]],
    "Abema-ND" = RColorBrewer::brewer.pal(12, "Paired")[[3]],
    "Abema-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[4]],
    "Parental" = "gray85"
  )

  expt <- gsub(".*bct-metadata_(.*)\\.tsv", "\\1", infile_meta)

  ml <- meta %>%
    dplyr::arrange(
      experiment,
      treatment,
      condition,
      timepoint,
      dose,
      replicate_id
    ) %>%
    dplyr::mutate(
      graph_name = str_c(
        experiment,
        treatment,
        condition,
        timepoint,
        sep = "\n"
      )
    ) %>%
    dplyr::pull(graph_name) %>%
    unique()

  meta <- meta %>%
    dplyr::mutate(
      graph_name = stringr::str_c(
        experiment,
        treatment,
        condition,
        timepoint,
        sep = "\n"
      )
    ) %>%
    dplyr::mutate(graph_name = factor(graph_name, levels = ml)) %>%
    dplyr::mutate(
      graph_name = forcats::fct_recode(
        graph_name,
        `Mut Vehicle CTR of AbemaR` = "abema-gradual\nDMSO\nDOX\nEnd",
        `Mut AbemaR` = "abema-gradual\nAbema\nDOX\nEnd",
        `WT Vehicle CTR of AbemaR` = "abema-gradual\nDMSO\nND\nEnd",
        `WT AbemaR` = "abema-gradual\nAbema\nND\nEnd",

        `Mut Vehicle CTR of PalboR` = "palbo-gradual\nDMSO\nDOX\nEnd",
        `Mut PalboR` = "palbo-gradual\nPalbo\nDOX\nEnd",
        `WT Vehicle CTR of PalboR` = "palbo-gradual\nDMSO\nND\nEnd",
        `WT PalboR` = "palbo-gradual\nPalbo\nND\nEnd"
      )
    ) |>
    dplyr::filter(
      timepoint %in% .timepoint,
      condition %in% .condition,
      treatment %in% .treatment
    ) %>%
    dplyr::mutate(graph_color = forcats::fct_drop(graph_color))

  counts_meta <- counts %>%
    dplyr::inner_join(meta, by = "sample_name") %>%
    dplyr::select(-fraction) %>% # drop existing fraction and recalculate
    dplyr::group_by(sample_name) %>%
    dplyr::mutate(fraction = count / sum(count)) %>%
    dplyr::ungroup()

  counts_meta <- counts_meta %>%
    dplyr::filter(
      timepoint %in% .timepoint,
      condition %in% .condition,
      treatment %in% .treatment
    ) |>
    dplyr::filter(!is.na(timepoint)) # remove unused samples.

  if (!is.null(freq_thresh)) {
    counts_meta <- counts_meta %>%
      dplyr::group_by(barcode) %>% # Filter out any barcodes that are below a certain threshold across all observations
      dplyr::filter(max(fraction) >= freq_thresh) %>%
      dplyr::ungroup()
  }

  # If specified take Palbo-gradual experiment Vehicle treated samples, not Palbo treated
  if (include_palbo_gradual_dmso_only) {
    counts_meta <- counts_meta |>
      dplyr::filter(!(experiment == "palbo-gradual" & treatment == "Palbo"))
  }

  # If Palbo, Abema, DMSO present, only take the DMSO from the Palbo sample
  if (
    all(c("Palbo", "Abema", "DMSO") %in% .treatment) &
      length(unique(.treatment)) == 3
  ) {
    counts_meta <- counts_meta %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "DMSO"))
  }

  # If Palbo, Abema, DMSO, Parental present, only take the DMSO and Parental from the Palbo sample
  if (
    all(c("Palbo", "Abema", "DMSO", "Parental") %in% .treatment) &
      length(unique(.treatment == 4))
  ) {
    counts_meta <- counts_meta %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "DMSO")) %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "Parental"))
  }

  # Test if proper subset of samples is selected
  stopifnot(
    all(counts_meta$timepoint %in% .timepoint),
    length(unique(counts_meta$timepoint)) == length(unique(.timepoint)),
    all(counts_meta$condition %in% .condition),
    length(unique(counts_meta$condition)) == length(unique(.condition)),
    all(counts_meta$treatment %in% .treatment),
    length(unique(counts_meta$treatment)) == length(unique(.treatment))
  )

  counts_meta_entropy <- counts_meta |>
    dplyr::group_by(
      sample_name,
      graph_name,
      graph_color,
      timepoint,
      condition,
      treatment,
      replicate_id
    ) %>%
    dplyr::summarize(
      n_unique_barcodes = dplyr::n(),
      n_count_barcodes = sum(count),
      kl_divergence = philentropy::KL(
        rbind(fraction, (rep(1, length(fraction)) / length(fraction))),
        unit = "log"
      ) /
        log(length(fraction)),
      n_unique_barcodes_check = length(fraction)
    ) |>
    dplyr::rename(Bioreplicate = replicate_id)

  # Plot KL divergences against a discrete normal density
  i <- "kl_divergence"
  yla = "Normalized KL divergence of barcode distribution\nagainst discrete uniform distribution"

  if (include_palbo_gradual_dmso_only) {
    outfile <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-KL-divergences_%s_%s_%s_%s_%s_palbo-gradual-DMSO-only.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-"),
      i
    )
    outfile_ttests <- sprintf(
      file.path(
        data_dir,
        "barcode-fraction-KL-divergences_%s_%s_%s_%s_%s_palbo-gradual-DMSO-only.csv"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-"),
      i
    )
    print(outfile)
  } else {
    outfile <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-KL-divergences_%s_%s_%s_%s_%s.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-"),
      i
    )
    outfile_ttests <- sprintf(
      file.path(data_dir, "barcode-fraction-KL-divergences_%s_%s_%s_%s_%s.csv"),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-"),
      i
    )
    print(outfile)
  }

  # Run an ANOVA across all categories.
  anova_res <- anova(lm(kl_divergence ~ graph_name, data = counts_meta_entropy))
  anova_subtitle <- sprintf(
    "(F(%i,%i)=%.2f; p=%.2e)",
    anova_res$Df[[1]],
    anova_res$Df[[2]],
    anova_res$`F value`[[1]],
    anova_res$`Pr(>F)`[[1]]
  )

  counts_meta_entropy_noparental <- counts_meta_entropy |>
    filter(timepoint != "Parental") # The one obs per group here causes issues with the sd estimation

  res_ttests <- counts_meta_entropy_noparental |>
    ungroup() |> # need to ungroup or else this breaks pairwite_t_test!
    dplyr::mutate(graph_name = forcats::fct_drop(graph_name)) |> # drop unused factor levels or t test tries to test empty groups
    rstatix::pairwise_t_test(
      formula = kl_divergence ~ graph_name,
      detailed = TRUE,
      pool.sd = FALSE,
      var.equal = FALSE
    ) |>
    rstatix::add_y_position()
  readr::write_csv(
    res_ttests |>
      mutate(
        group1 = gsub("\n", " ", group1),
        group2 = gsub("\n", " ", group2)
      ) |>
      as_tibble(),
    outfile_ttests
  )
  print(res_ttests, n = 100)

  set.seed(840)
  plt <- counts_meta_entropy |>
    ggplot2::ggplot(ggplot2::aes(x = graph_name, y = !!sym(i))) +
    ggplot2::geom_point(
      aes(color = graph_name, shape = Bioreplicate), # color shape aes here or else ggpubr::stat_pvalue_manual breaks
      position = ggplot2::position_dodge2(width = .5),
      size = 3
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    ) +
    ggplot2::labs(
      x = NULL,
      y = yla,
      title = "Barcode fraction distribution KL divergences",
      subtitle = anova_subtitle
    ) +
    ggplot2::ylim(0, NA) +
    ggplot2::scale_color_manual(
      values = pal_kl_graph_name,
      drop = TRUE,
      guide = "none"
    )

  if (padj.signif.stars) {
    plt <- plt +
      ggpubr::stat_pvalue_manual(res_ttests, label = "p.adj.signif") +
      ggplot2::scale_y_continuous(
        breaks = c(0, .25, .5, .75, 1, 1.25),
        limits = c(0, 1.35)
      ) # so plots line up nicely
  }

  if (!is.null(plt_t_test_var1) & !is.null(plt_t_test_var2)) {
    a <- counts_meta_entropy |>
      filter(!!sym(plt_var_x) == plt_t_test_var1) |>
      pull(!!sym(i))
    b <- counts_meta_entropy |>
      filter(!!sym(plt_var_x) == plt_t_test_var2) |>
      pull(!!sym(i))

    print(counts_meta_entropy |> filter(!!sym(plt_var_x) == plt_t_test_var1))
    print(counts_meta_entropy |> filter(!!sym(plt_var_x) == plt_t_test_var2))

    ttest_summary <- t.test(a, b)

    outfile <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-KL-divergences_%s_%s_%s_%s_%s_vs_%s_%s.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-"),
      plt_t_test_var1,
      plt_t_test_var2,
      i
    )

    plt <- plt +
      labs(
        subtitle = sprintf(
          "t-test: %s vs %s; pval = %s",
          plt_t_test_var1,
          plt_t_test_var2,
          signif(ttest_summary$p.value, 4)
        )
      )
    # }
  }
  ggplot2::ggsave(outfile, plt, height = height, width = width)
  return(plt)
}
```

### Plot KL divergences for Palbo, Vehicle samples (Figure 2E).

```{r}
plot_barcode_fraction_KL_divergences_against_discrete_uniform_cell_line(
  infile_counts = "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv",
  infile_meta = "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv",
  .condition = c("ND", "DOX"),
  .treatment = c("Palbo", "DMSO"),
  .timepoint = c("End"),
  freq_thresh = NULL
)
```

### Plot KL divergences for Abema, Vehicle samples (Figure 2F).

```{r}
plot_barcode_fraction_KL_divergences_against_discrete_uniform_cell_line(
  infile_counts = "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
  infile_meta = "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
  .condition = c("ND", "DOX"),
  .treatment = c("Abema", "DMSO"),
  .timepoint = c("End"),
  freq_thresh = NULL
)
```

## Load in function to plot sample-sample heatmap of JS divergences of sample barcode distributions.

```{r}
#' Plot sample-sample heatmap of Jensen-Shannon divergences of each sample
#' compared against every other sample for the specified cell line samples.
#'
#' Note, this is a symmetric divergence metric that uses the mean distribution
#' .5(P+Q) of the two input distributions P, Q.
#'
#' @param infile_counts Path to barcode counts.
#' @param infile_meta Path to barcode metadata.
#' @param .condition Vector of conditions to filter to.
#' @param .treatment Vector of treatments to filter to.
#' @param .timepoint Vector of timepionts to filter to.
#' @param include_palbo_gradual_dmso_only Include only the DMSO Vehicle sample
#' from the palbo gradual experiment samples?
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param data_dir Processed data directory.
#' @param output_dir Processed output directory.
plot_barcode_fraction_JS_divergences <- function(
  infile_counts,
  infile_meta,
  .condition = c("Parental", "ND", "DOX"),
  .treatment = c("Parental", "DMSO", "Abema", "Palbo"),
  .timepoint = c("Parental", "Early", "Middle", "End"),
  include_palbo_gradual_dmso_only = FALSE,
  width = 7,
  height = 7,
  data_dir = "data/processed/01-barcoding",
  output_dir = "data/processed/01-barcoding"
) {
  counts <- purrr::map_dfr(
    infile_counts,
    readr::read_tsv,
    show_col_types = FALSE
  )
  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    dplyr::mutate(
      condition = factor(condition, levels = c("Parental", "ND", "DOX")),
      treatment = factor(
        treatment,
        levels = c("Parental", "DMSO", "Palbo", "Abema")
      ),
      dose = factor(dose, levels = c("/", "50nM", "200nM", "1uM")),
      timepoint = factor(
        timepoint,
        levels = c("Parental", "Early", "Middle", "End")
      ),
      replicate_id = factor(replicate_id, levels = c("1", "2", "3")),
      experiment = factor(
        experiment,
        levels = c("palbo-gradual", "abema-gradual")
      )
    ) %>%
    dplyr::mutate(
      graph_color = dplyr::case_when(
        treatment == "DMSO" & condition == "ND" ~ "DMSO-ND",
        treatment == "DMSO" & condition == "DOX" ~ "DMSO-DOX",
        treatment == "Palbo" & condition == "ND" ~ "Palbo-ND",
        treatment == "Palbo" & condition == "DOX" ~ "Palbo-DOX",
        treatment == "Abema" & condition == "ND" ~ "Abema-ND",
        treatment == "Abema" & condition == "DOX" ~ "Abema-DOX",
        .default = "Parental"
      )
    ) %>%
    dplyr::mutate(
      graph_color = factor(
        graph_color,
        levels = c(
          "Parental",
          "DMSO-ND",
          "DMSO-DOX",
          "Palbo-ND",
          "Palbo-DOX",
          "Abema-ND",
          "Abema-DOX"
        )
      )
    )

  col_pal <- list(
    "DMSO-ND" = RColorBrewer::brewer.pal(12, "Paired")[[7]],
    "DMSO-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[8]],
    "Palbo-ND" = RColorBrewer::brewer.pal(12, "Paired")[[1]],
    "Palbo-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[2]],
    "Abema-ND" = RColorBrewer::brewer.pal(12, "Paired")[[3]],
    "Abema-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[4]],
    "Parental" = "gray85"
  )

  expt <- gsub(".*bct-metadata_(.*)\\.tsv", "\\1", infile_meta)

  ml <- meta %>%
    dplyr::arrange(
      experiment,
      treatment,
      condition,
      timepoint,
      dose,
      replicate_id
    ) %>%
    dplyr::mutate(
      graph_name = str_c(experiment, treatment, condition, timepoint, sep = " ")
    ) %>%
    dplyr::pull(graph_name) %>%
    unique()

  meta <- meta %>%
    dplyr::mutate(
      graph_name = stringr::str_c(
        experiment,
        treatment,
        condition,
        timepoint,
        sep = " "
      )
    ) %>%
    dplyr::mutate(graph_name = factor(graph_name, levels = ml)) %>%
    dplyr::mutate(
      graph_name = forcats::fct_recode(
        graph_name,
        `Mut Vehicle CTR of AbemaR` = "abema-gradual DMSO DOX End",
        `Mut AbemaR` = "abema-gradual Abema DOX End",
        `WT Vehicle CTR of AbemaR` = "abema-gradual DMSO ND End",
        `WT AbemaR` = "abema-gradual Abema ND End",

        `Mut Vehicle CTR of PalboR` = "palbo-gradual DMSO DOX End",
        `Mut PalboR` = "palbo-gradual Palbo DOX End",
        `WT Vehicle CTR of PalboR` = "palbo-gradual DMSO ND End",
        `WT PalboR` = "palbo-gradual Palbo ND End"
      )
    ) |>
    dplyr::filter(
      timepoint %in% .timepoint,
      condition %in% .condition,
      treatment %in% .treatment
    ) %>%
    dplyr::mutate(graph_color = forcats::fct_drop(graph_color))

  counts_meta <- counts %>%
    dplyr::inner_join(meta, by = "sample_name") %>%
    dplyr::select(-fraction) %>% # drop existing fraction and recalculate
    dplyr::group_by(sample_name) %>%
    dplyr::mutate(fraction = count / sum(count)) %>%
    dplyr::ungroup()

  counts_meta <- counts_meta %>%
    dplyr::filter(
      timepoint %in% .timepoint,
      condition %in% .condition,
      treatment %in% .treatment
    ) |>
    dplyr::filter(!is.na(timepoint)) # remove unused samples.

  # If specified take Palbo-gradual experiment Vehicle treated samples, not Palbo treated
  if (include_palbo_gradual_dmso_only) {
    counts_meta <- counts_meta |>
      dplyr::filter(!(experiment == "palbo-gradual" & treatment == "Palbo"))
  }

  # If Palbo, Abema, DMSO present, only take the DMSO from the Palbo sample
  if (
    all(c("Palbo", "Abema", "DMSO") %in% .treatment) &
      length(unique(.treatment)) == 3
  ) {
    counts_meta <- counts_meta %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "DMSO"))
  }

  # If Palbo, Abema, DMSO, Parental present, only take the DMSO and Parental from the Palbo sample
  if (
    all(c("Palbo", "Abema", "DMSO", "Parental") %in% .treatment) &
      length(unique(.treatment == 4))
  ) {
    counts_meta <- counts_meta %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "DMSO")) %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "Parental"))
  }

  # Test if proper subset of samples is selected
  stopifnot(
    all(counts_meta$timepoint %in% .timepoint),
    length(unique(counts_meta$timepoint)) == length(unique(.timepoint)),
    all(counts_meta$condition %in% .condition),
    length(unique(counts_meta$condition)) == length(unique(.condition)),
    all(counts_meta$treatment %in% .treatment),
    length(unique(counts_meta$treatment)) == length(unique(.treatment))
  )

  frac_meta_wide <- counts_meta |>
    dplyr::mutate(
      graph_name_replicate_id = stringr::str_c(graph_name, " ", replicate_id)
    ) |>
    tidyr::pivot_wider(
      id_cols = graph_name_replicate_id,
      names_from = barcode,
      values_from = fraction,
      values_fill = 0
    ) |>
    tibble::column_to_rownames("graph_name_replicate_id") |>
    as.matrix()

  res <- philentropy::distance(
    x = frac_meta_wide,
    method = "jensen-shannon",
    test.na = TRUE,
    unit = "log2",
    est.prob = NULL,
    use.row.names = TRUE
  )
  res_dist <- as.dist(res)

  colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

  res_evolutionary_replicates <- res |>
    as.data.frame() |>
    tibble::rownames_to_column("s1") |>
    tibble::as_tibble() |>
    tidyr::pivot_longer(cols = !s1, names_to = "s2", values_to = "js") |>
    dplyr::filter(s1 != s2) |>
    dplyr::mutate(grp_s1 = gsub(" [123]", "", s1)) |>
    dplyr::mutate(grp_s2 = gsub(" [123]", "", s2)) |>
    dplyr::filter(
      (grepl("WT", s1) & grepl("WT", s2)) |
        (grepl("Mut", s1) & grepl("Mut", s2))
    ) |>
    dplyr::filter(
      (grepl("Palbo", grp_s1) & grepl("Abema", grp_s2)) |
        (grepl("Abema", grp_s1) & grepl("Palbo", grp_s2))
    ) |>
    dplyr::group_by(grp_s1, grp_s2) |>
    dplyr::summarize(
      min_js = (min(js)),
      max_js = (max(js)),
      min_js_round = round(min(js), 2),
      max_js_round = round(max(js), 2)
    )

  res_evolutionary_replicates <- res |>
    as.data.frame() |>
    tibble::rownames_to_column("s1") |>
    tibble::as_tibble() |>
    tidyr::pivot_longer(cols = !s1, names_to = "s2", values_to = "js") |>
    dplyr::filter(s1 != s2) |>
    dplyr::mutate(grp_s1 = gsub(" [123]", "", s1)) |>
    dplyr::mutate(grp_s2 = gsub(" [123]", "", s2)) |>
    dplyr::filter(grp_s1 == grp_s2)

  res_evolutionary_replicates |>
    dplyr::arrange(js) |>
    dplyr::mutate(js_round = round(js, 2)) |>
    print(n = 1000)

  res_evolutionary_replicates |>
    dplyr::arrange(js) |>
    dplyr::mutate(js_round = round(js, 2)) |>
    dplyr::group_by(grp_s1, grp_s2) |>
    dplyr::summarize(min_js = round(min(js), 2), max_js = round(max(js), 2)) |>
    print(n = 1000)

  if (include_palbo_gradual_dmso_only) {
    outfile <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-JS-divergences_%s_%s_%s_%s_palbo-gradual-DMSO-only.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-")
    )
    outfile_with_numbers <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-JS-divergences-with-numbers_%s_%s_%s_%s_palbo-gradual-DMSO-only.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-")
    )
    print(outfile)
  } else {
    outfile <- sprintf(
      file.path(output_dir, "barcode-fraction-JS-divergences_%s_%s_%s_%s.pdf"),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-")
    )
    outfile_with_numbers <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-JS-divergences-with-numbers_%s_%s_%s_%s.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(.treatment, collapse = "-")
    )
    print(outfile)
  }

  print(
    pheatmap::pheatmap(
    res,
    clustering_distance_rows = res_dist,
    clustering_distance_cols = res_dist,
    breaks = seq(0, 1, length.out = 256),
    col = colors,
    show_rownames = TRUE,
    show_colnames = TRUE,
    main = "JS Divergence",
    # filename = outfile_with_numbers, # uncomment to save to file
    width = width + 2,
    height = height + 2,
    cellwidth = 20,
    cellheight = 20,
    display_numbers = TRUE
  ))
}
```

### Plot sample-sample heatmap of JS divergences for Palbo, Vehicle samples (Figure 2G).

```{r}
#| fig-width: 8
#| fig-height: 8
plot_barcode_fraction_JS_divergences(
  infile_counts = c("data/raw/01-barcoding/bct-counts_palbo-gradual.tsv"),
  infile_meta = c("data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv"),
  .condition = c("DOX", "ND"),
  .treatment = c("DMSO", "Palbo"),
  .timepoint = c("End"),
  width = 7
)
```

### Plot sample-sample heatmap of JS divergences for Abema, Vehicle samples (Figure 2H).

```{r}
#| fig-width: 8
#| fig-height: 8
plot_barcode_fraction_JS_divergences(
  infile_counts = c("data/raw/01-barcoding/bct-counts_abema-gradual.tsv"),
  infile_meta = c("data/raw/01-barcoding/bct-metadata_abema-gradual.tsv"),
  .condition = c("DOX", "ND"),
  .treatment = c("Abema", "DMSO"),
  .timepoint = c("End"),
  width = 7
)
```

## Load function to plot JS divergences over time.

```{r}
#' Plot Jensen-Shannon divergences of each sample over time, run paired
#' t-tests comparing timepoints.
#'
#' Note, this is a symmetric divergence metric that uses the mean distribution
#' .5(P+Q) of your two input distributions P, Q.
#'
#' @param infile_counts Path to barcode counts.
#' @param infile_meta Path to barcode metadata.
#' @param .condition Vector of conditions to filter to.
#' @param .treatment Vector of treatments to filter to.
#' @param .timepoint Vector of timepionts to filter to.
#' @param include_palbo_gradual_dmso_only Include only the DMSO Vehicle sample
#' from palbo gradual?
#' @param freq_thresh Barcode frequency threshold to filter out any barcode
#' that fails to meet threshold over any timepoint for a sample.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param data_dir Processed data directory.
#' @param output_dir Processed output directory.
plot_barcode_fraction_JS_divergences_scatterplot <- function(
  infile_counts,
  infile_meta,
  .condition = c("Parental", "ND", "DOX"),
  .treatment = c("Parental", "DMSO", "Abema", "Palbo"),
  .timepoint = c("Parental", "Early", "Middle", "End"),
  include_palbo_gradual_dmso_only = FALSE,
  freq_thresh = NULL,
  width = 5,
  height = 4,
  data_dir = "data/processed/01-barcoding",
  output_dir = "data/processed/01-barcoding"
) {
  counts <- purrr::map_dfr(
    infile_counts,
    readr::read_tsv,
    show_col_types = FALSE
  )
  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    dplyr::mutate(
      condition = factor(condition, levels = c("Parental", "ND", "DOX")),
      treatment = factor(
        treatment,
        levels = c("Parental", "DMSO", "Palbo", "Abema")
      ),
      dose = factor(dose, levels = c("/", "50nM", "200nM", "1uM")),
      timepoint = factor(
        timepoint,
        levels = c("Parental", "Early", "Middle", "End")
      ),
      replicate_id = factor(replicate_id, levels = c("1", "2", "3")),
      experiment = factor(
        experiment,
        levels = c("palbo-gradual", "abema-gradual")
      )
    ) %>%
    dplyr::mutate(
      graph_color = dplyr::case_when(
        treatment == "DMSO" &
          condition == "ND" &
          experiment == "palbo-gradual" ~ "DMSO-ND-pg",
        treatment == "DMSO" &
          condition == "DOX" &
          experiment == "palbo-gradual" ~ "DMSO-DOX-pg",
        treatment == "DMSO" &
          condition == "ND" &
          experiment == "abema-gradual" ~ "DMSO-ND-ag",
        treatment == "DMSO" &
          condition == "DOX" &
          experiment == "abema-gradual" ~ "DMSO-DOX-ag",
        treatment == "Palbo" & condition == "ND" ~ "Palbo-ND",
        treatment == "Palbo" & condition == "DOX" ~ "Palbo-DOX",
        treatment == "Abema" & condition == "ND" ~ "Abema-ND",
        treatment == "Abema" & condition == "DOX" ~ "Abema-DOX",
        .default = "Parental"
      )
    ) %>%
    dplyr::mutate(
      graph_color = factor(
        graph_color,
        levels = c(
          "Parental",
          "DMSO-ND-pg",
          "DMSO-DOX-pg",
          "DMSO-ND-ag",
          "DMSO-DOX-ag",
          "Palbo-ND",
          "Palbo-DOX",
          "Abema-ND",
          "Abema-DOX"
        )
      )
    )

  col_pal <- list(
    "DMSO-ND-pg" = RColorBrewer::brewer.pal(12, "Paired")[[7]],
    "DMSO-DOX-pg" = RColorBrewer::brewer.pal(12, "Paired")[[8]],
    "DMSO-ND-ag" = RColorBrewer::brewer.pal(12, "Paired")[[7]],
    "DMSO-DOX-ag" = RColorBrewer::brewer.pal(12, "Paired")[[8]],
    "Palbo-ND" = RColorBrewer::brewer.pal(12, "Paired")[[1]],
    "Palbo-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[2]],
    "Abema-ND" = RColorBrewer::brewer.pal(12, "Paired")[[3]],
    "Abema-DOX" = RColorBrewer::brewer.pal(12, "Paired")[[4]],
    "Parental" = "gray85"
  )

  expt <- gsub(".*bct-metadata_(.*)\\.tsv", "\\1", infile_meta)

  ml <- meta %>%
    dplyr::arrange(
      experiment,
      treatment,
      condition,
      timepoint,
      dose,
      replicate_id
    ) %>%
    dplyr::mutate(
      graph_name = str_c(experiment, treatment, condition, timepoint, sep = " ")
    ) %>%
    dplyr::pull(graph_name) %>%
    unique()

  meta <- meta %>%
    dplyr::mutate(
      graph_name = stringr::str_c(
        experiment,
        treatment,
        condition,
        timepoint,
        sep = " "
      )
    ) %>%
    dplyr::mutate(graph_name = factor(graph_name, levels = ml)) %>%
    dplyr::mutate(
      graph_name = forcats::fct_recode(
        graph_name,
        `Mut Vehicle CTR of AbemaR` = "abema-gradual DMSO DOX End",
        `Mut AbemaR` = "abema-gradual Abema DOX End",
        `WT Vehicle CTR of AbemaR` = "abema-gradual DMSO ND End",
        `WT AbemaR` = "abema-gradual Abema ND End",

        `Mut Vehicle CTR of PalboR` = "palbo-gradual DMSO DOX End",
        `Mut PalboR` = "palbo-gradual Palbo DOX End",
        `WT Vehicle CTR of PalboR` = "palbo-gradual DMSO ND End",
        `WT PalboR` = "palbo-gradual Palbo ND End"
      )
    ) |>
    dplyr::filter(
      timepoint %in% .timepoint,
      condition %in% .condition,
      treatment %in% .treatment
    ) %>%
    dplyr::mutate(graph_color = forcats::fct_drop(graph_color))

  counts_meta <- counts %>%
    dplyr::inner_join(meta, by = "sample_name") %>%
    dplyr::select(-fraction) %>% # drop existing fraction and recalculate
    dplyr::group_by(sample_name) %>%
    dplyr::mutate(fraction = count / sum(count)) %>%
    dplyr::ungroup()

  counts_meta <- counts_meta %>%
    dplyr::filter(
      timepoint %in% .timepoint,
      condition %in% .condition,
      treatment %in% .treatment
    ) |>
    dplyr::filter(!is.na(timepoint)) # remove unused samples.

  # If specified take Palbo-gradual experiment Vehicle treated samples, not Palbo treated
  if (include_palbo_gradual_dmso_only) {
    counts_meta <- counts_meta |>
      dplyr::filter(!(experiment == "palbo-gradual" & treatment == "Palbo"))
  }

  # If Palbo, Abema, DMSO present, only take the DMSO from the Palbo sample
  if (
    all(c("Palbo", "Abema", "DMSO") %in% .treatment) &
      length(unique(.treatment)) == 3
  ) {
    counts_meta <- counts_meta %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "DMSO"))
  }

  # If Palbo, Abema, DMSO, Parental present, only take the DMSO and Parental from the Palbo sample
  if (
    all(c("Palbo", "Abema", "DMSO", "Parental") %in% .treatment) &
      length(unique(.treatment == 4))
  ) {
    counts_meta <- counts_meta %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "DMSO")) %>%
      dplyr::filter(!(experiment == "abema-gradual" & treatment == "Parental"))
  }

  # Test if proper subset of samples is selected
  stopifnot(
    all(counts_meta$timepoint %in% .timepoint),
    length(unique(counts_meta$timepoint)) == length(unique(.timepoint)),
    all(counts_meta$condition %in% .condition),
    length(unique(counts_meta$condition)) == length(unique(.condition)),
    all(counts_meta$treatment %in% .treatment),
    length(unique(counts_meta$treatment)) == length(unique(.treatment))
  )

  frac_meta_wide <- counts_meta |>
    dplyr::mutate(
      graph_name_replicate_id = stringr::str_c(graph_name, " ", replicate_id)
    ) |>
    tidyr::pivot_wider(
      id_cols = sample_name,
      names_from = barcode,
      values_from = fraction,
      values_fill = 0
    ) |>
    tibble::column_to_rownames("sample_name") |>
    as.matrix()

  res <- philentropy::distance(
    x = frac_meta_wide,
    method = "jensen-shannon",
    test.na = TRUE,
    unit = "log2",
    est.prob = NULL,
    use.row.names = TRUE
  )
  res_dist <- as.dist(res)

  colors <- colorRampPalette(rev(brewer.pal(9, "Blues")))(255)

  res_df <- res |>
    as.data.frame() |>
    tibble::rownames_to_column("s1") |>
    tibble::as_tibble() |>
    tidyr::pivot_longer(cols = !s1, names_to = "s2", values_to = "js") |>
    dplyr::filter(
      s1 %in%
        c(
          "20171215_MCYS_D17_350_3_CG5109_S3_R1_001",
          "20200226_Basal_P10_CG7902_S1_R1_001"
        )
    ) |> # The parental sample for either Palbo or Abema experiment
    dplyr::filter(
      !(s2 %in%
        c(
          "20171215_MCYS_D17_350_3_CG5109_S3_R1_001",
          "20200226_Basal_P10_CG7902_S1_R1_001"
        ))
    ) |>
    dplyr::inner_join(meta, by = c("s2" = "sample_name")) |>
    dplyr::mutate(
      condition = forcats::fct_recode(
        condition,
        `WT` = "ND",
        `Mut` = "DOX"
      )
    ) |>
    dplyr::mutate(
      treatment = forcats::fct_recode(
        treatment,
        `Vehicle` = "DMSO"
      )
    ) |>
    dplyr::mutate(graph_grp = stringr::str_c(condition, " ", treatment)) |>
    dplyr::mutate(
      graph_grp = forcats::fct_relevel(
        graph_grp,
        "WT Palbo",
        "Mut Palbo",
        "WT Abema",
        "Mut Abema",
        "WT Vehicle",
        "Mut Vehicle"
      )
    ) |>
    dplyr::rename(Replicate = "replicate_id")

  res_df_wide <- res_df |>
    dplyr::arrange(condition, timepoint, desc(timepoint), Replicate) |>
    dplyr::mutate(timepoint = as.character(timepoint)) |>
    dplyr::select(-passage, -dose, -graph_name, -s1, -s2) |>
    tidyr::pivot_wider(names_from = timepoint, values_from = js)

  # Run 'pairwise' (manually specified timepoint comparisons) **paired** ttests
  out_res <- res_df_wide |>
    dplyr::group_split(condition, treatment) |>
    purrr::map(\(df) {
      list(c("End", "Middle"), c("End", "Early"), c("Middle", "Early")) |>
        purrr::map(\(tm) {
          t.test(
            x = (df |> dplyr::pull(!!sym(tm[[1]]))),
            y = (df |> dplyr::pull(!!sym(tm[[2]]))),
            paired = TRUE
          ) |>
            broom::tidy() |>
            dplyr::mutate(group1 = tm[[1]], group2 = tm[[2]]) |>
            dplyr::relocate(group1, group2)
        }) |>
        purrr::list_rbind() |>
        dplyr::mutate(
          condition = unique(df$condition),
          treatment = unique(df$treatment)
        ) |>
        dplyr::relocate(condition, treatment)
    }) |>
    purrr::list_rbind() |>
    dplyr::group_by(condition, treatment) |>
    dplyr::mutate(p.adj.holm = p.adjust(p.value, method = "holm")) |>
    dplyr::ungroup() |>
    rstatix::add_significance(p.col = "p.adj.holm") # use default cutpoints, for graphing

  outfile_paired_ttests <- sprintf(
    file.path(
      data_dir,
      "barcode-fraction-JS-divergences-paired-ttests_%s_%s_%s_%s.csv"
    ),
    paste0(expt, collapse = "-"),
    paste0(.timepoint, collapse = "-"),
    paste0(.condition, collapse = "-"),
    paste0(.treatment, collapse = "-")
  )

  readr::write_csv(out_res, outfile_paired_ttests)

  (plt <- res_df |>
    ggplot2::ggplot(aes(
      y = js,
      x = timepoint,
      color = graph_grp,
      shape = Replicate
    )) +
    ggplot2::geom_point(
      size = 2,
      position = ggplot2::position_dodge(width = .5)
    ) +
    ggplot2::scale_y_continuous(breaks = seq(0, 1.25, .25)) +
    ggplot2::theme_bw() +
    ggplot2::labs(
      y = "JS divergence against Parental",
      x = NULL,
      color = "Group"
    ) +
    ggplot2::scale_color_manual(values = pal_js_graph_name) +
    ggplot2::coord_cartesian(ylim = c(0, NA)))

  if (nrow(out_res) == 3) {
    plt <- plt +
      ggpubr::stat_pvalue_manual(
        out_res |>
          dplyr::mutate(graph_grp = stringr::str_c(condition, " ", treatment)),
        group = "graph_grp",
        y.position = 1.1,
        label = "p.adj.holm.signif",
        step.increase = .1,
        shape = NA,
        legend.show = FALSE
      )
  }

  (plt_faceted <- res_df |>
    ggplot2::ggplot(aes(
      y = js,
      x = timepoint,
      color = graph_grp,
      shape = Replicate
    )) +
    ggplot2::geom_point(
      size = 2,
      position = ggplot2::position_dodge(width = .5)
    ) +
    ggplot2::scale_y_continuous(breaks = seq(0, 1.5, .25)) +
    ggplot2::theme_bw() +
    ggplot2::labs(
      y = "JS divergence against Parental",
      x = NULL,
      color = "Group"
    ) +
    ggplot2::facet_wrap(~graph_grp) +
    ggplot2::scale_color_manual(values = pal_js_graph_name) +
    ggplot2::coord_cartesian(ylim = c(0, 1.5)) +
    ggpubr::stat_pvalue_manual(
      out_res |>
        dplyr::mutate(graph_grp = stringr::str_c(condition, " ", treatment)) |>
        dplyr::mutate(
          graph_grp = forcats::fct_relevel(
            graph_grp,
            "WT Palbo",
            "Mut Palbo",
            "WT Abema",
            "Mut Abema",
            "WT Vehicle",
            "Mut Vehicle"
          )
        ) |>
        dplyr::group_by(graph_grp) |>
        dplyr::mutate(ypos = 1.1 + (row_number() - 1) * 0.15) |> # manually make spacing for test sigstar brackets
        dplyr::ungroup(),
      group = "graph_grp",
      y.position = "ypos",
      label = "p.adj.holm.signif",
      shape = NA,
      legend.show = FALSE
    ))

  outfile <- sprintf(
    file.path(
      output_dir,
      "barcode-fraction-JS-divergences-scatterplot_%s_%s_%s_%s.pdf"
    ),
    paste0(expt, collapse = "-"),
    paste0(.timepoint, collapse = "-"),
    paste0(.condition, collapse = "-"),
    paste0(.treatment, collapse = "-")
  )

  outfile_faceted <- sprintf(
    file.path(
      output_dir,
      "barcode-fraction-JS-divergences-scatterplot_%s_%s_%s_%s-faceted.pdf"
    ),
    paste0(expt, collapse = "-"),
    paste0(.timepoint, collapse = "-"),
    paste0(.condition, collapse = "-"),
    paste0(.treatment, collapse = "-")
  )

  ggplot2::ggsave(outfile, plt, height = height, width = width)
  ggplot2::ggsave(outfile_faceted, plt_faceted, height = 5, width = 6)
  ggplot2::ggsave(
    gsub("pdf$", "png", outfile),
    plt,
    height = height,
    width = width
  )

  print(outfile)

  return(plt_faceted)
}
```

### Plot JS divergences over time for Palbo, Vehicle samples (Supplementary Figure S2G).

```{r}
plot_barcode_fraction_JS_divergences_scatterplot(
  infile_counts = c("data/raw/01-barcoding/bct-counts_palbo-gradual.tsv"),
  infile_meta = c("data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv"),
  .condition = c("DOX", "ND", "Parental"),
  .treatment = c("DMSO", "Palbo", "Parental")
)
```

### Plot JS divergences over time for Abema, Vehicle samples (Supplementary Figure S2H).

```{r}
plot_barcode_fraction_JS_divergences_scatterplot(
  infile_counts = c("data/raw/01-barcoding/bct-counts_abema-gradual.tsv"),
  infile_meta = c("data/raw/01-barcoding/bct-metadata_abema-gradual.tsv"),
  .condition = c("DOX", "ND", "Parental"),
  .treatment = c("DMSO", "Abema", "Parental")
)
```

### Plot sample-sample heatmap of JS divergences for Palbo, Abema samples (Figure 3A).

```{r}
#| fig-width: 8
#| fig-height: 8
plot_barcode_fraction_JS_divergences(
  infile_counts = c(
    "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
    "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv"
  ),
  infile_meta = c(
    "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
    "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv"
  ),
  .condition = c("DOX", "ND"),
  .treatment = c("Abema", "Palbo"),
  .timepoint = c("End"),
  width = 7
)
```

## Cluster and plot fishplots of clonal prevalence for Parental and PalboR and AbemaR evolutionary replicates. 

### WT PalboR (Figure 3D).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "WT PalboR escalating",
  .condition = c("Parental", "ND"),
  .treatment = c("Parental", "Palbo"),
  dist_metric = "column-swept-ppr"
)
```

### Mut PalboR (Figure 3E).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "Mut PalboR escalating",
  .condition = c("Parental", "DOX"),
  .treatment = c("Parental", "Palbo"),
  dist_metric = "column-swept-ppr"
)
```

### WT AbemaR (Figure 3F).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "WT AbemaR",
  .condition = c("Parental", "ND"),
  .treatment = c("Parental", "Abema"),
  dist_metric = "column-swept-ppr"
)
```

### Mut AbemaR (Figure 3G).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "Mut AbemaR",
  .condition = c("Parental", "DOX"),
  .treatment = c("Parental", "Abema"),
  dist_metric = "column-swept-ppr"
)
```

## Load in function to plot KL divergences for clusters.

```{r}
#' Plot scatterplot of Shannon entropies of the 8 clusters per each given experimental
#' condition, run ANOVA/ttest across plotted groups.
#'
#' Used to quantify how different drug treatments lead to different distributions
#' of the End timepoint cluster cumulative fractions.
#'
#' @param infiles Vector of path to KL divergences of fishplot clusters
#' generated by [plot_bct_frac_fishplot()].
#' @param suffix Suffix to add to output filepath.
#' @param filter_timepoint Vector of timepoints to filter to. Keeps all if NULL.
#' @param padj.signif.stars Plot significance stars?
#' @param dist_type Distance metric used to generate the feature dist and
#' sample dist matrices.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_kl_divergences_clusters <- function(
  infiles,
  suffix,
  filter_timepoint = NULL,
  padj.signif.stars = TRUE,
  dist_type = "column-swept-ppr",
  ddir = "data/processed/01-barcoding",
  odir = "data/processed/01-barcoding"
) {
  if (!dir.exists(file.path(ddir, "entropy-plots", dist_type))) {
    dir.create(file.path(ddir, "entropy-plots", dist_type), recursive = TRUE)
  }
  if (!dir.exists(file.path(odir, "entropy-plots", dist_type))) {
    dir.create(file.path(odir, "entropy-plots", dist_type), recursive = TRUE)
  }

  infiles |>
    purrr::map_dfr(readRDS) |>
    dplyr::distinct(metadata)

  d_plt <- infiles |>
    purrr::map_dfr(readRDS) |>
    dplyr::mutate(
      metadata_plt = forcats::fct_recode(
        metadata,
        `WT PalboR` = "column-swept-ppr_palbo-gradual_Parental-ND_Parental-Palbo_Parental-Early-Middle-End_thresh0p001",
        `Mut PalboR` = "column-swept-ppr_palbo-gradual_Parental-DOX_Parental-Palbo_Parental-Early-Middle-End_thresh0p001",
        `WT AbemaR` = "column-swept-ppr_abema-gradual_Parental-ND_Parental-Abema_Parental-Early-Middle-End_thresh0p001",
        `Mut AbemaR` = "column-swept-ppr_abema-gradual_Parental-DOX_Parental-Abema_Parental-Early-Middle-End_thresh0p001",

        `WT Vehicle CTR of PalboR` = "column-swept-ppr_palbo-gradual_Parental-ND_Parental-DMSO_Parental-Early-Middle-End_thresh0p001",
        `Mut Vehicle CTR of PalboR` = "column-swept-ppr_palbo-gradual_Parental-DOX_Parental-DMSO_Parental-Early-Middle-End_thresh0p001",
        `WT Vehicle CTR of AbemaR` = "column-swept-ppr_abema-gradual_Parental-ND_Parental-DMSO_Parental-Early-Middle-End_thresh0p001",
        `Mut Vehicle CTR of AbemaR` = "column-swept-ppr_abema-gradual_Parental-DOX_Parental-DMSO_Parental-Early-Middle-End_thresh0p001",
      )
    ) |>
    dplyr::mutate(
      metadata_plt = forcats::fct_relevel(
        metadata_plt,

        "WT Vehicle CTR of PalboR",
        "Mut Vehicle CTR of PalboR",
        "WT PalboR",
        "Mut PalboR",

        "WT Vehicle CTR of AbemaR",
        "Mut Vehicle CTR of AbemaR",
        "WT AbemaR",
        "Mut AbemaR"
      )
    ) |>
    dplyr::distinct(
      metadata_plt,
      plot_bioreplicate_id,
      timepoint,
      normalized_kl_divergence
    ) |>
    dplyr::mutate(
      plot_bioreplicate_id = gsub("Bioreplicate ", "", plot_bioreplicate_id)
    ) |>
    dplyr::rename(`Bioreplicate` = plot_bioreplicate_id)

  if (!is.null(filter_timepoint)) {
    d_plt <- d_plt |> dplyr::filter(timepoint %in% filter_timepoint)
  }

  plt <- d_plt |>
    ggplot(ggplot2::aes(
      x = metadata_plt,
      y = normalized_kl_divergence,
      color = Bioreplicate
    )) +
    geom_point(position = position_dodge2(.4)) +
    ylim(0, 1.2) +
    labs(
      title = "Clonal prevalence across resistant clusters",
      subtitle = paste0(filter_timepoint, collapse = "-"),
      x = NULL,
      y = "Normalized KL divergence of cluster distribution\nagainst discrete uniform distribution"
    ) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

  # -------------------------------------------------------------------------

  # Run an ANOVA across all categories.
  anova_res <- anova(lm(normalized_kl_divergence ~ metadata_plt, data = d_plt))
  anova_subtitle <- sprintf(
    "(F(%i,%i)=%.2f; p=%.2e)",
    anova_res$Df[[1]],
    anova_res$Df[[2]],
    anova_res$`F value`[[1]],
    anova_res$`Pr(>F)`[[1]]
  )

  res_ttests <- d_plt |>
    ungroup() |> # ungroup or else this breaks pairwise_t_test
    rstatix::pairwise_t_test(
      formula = normalized_kl_divergence ~ metadata_plt,
      detailed = TRUE,
      pool.sd = FALSE,
      var.equal = FALSE
    ) |>
    rstatix::add_y_position()
  readr::write_csv(
    res_ttests |> as_tibble(),
    file.path(
      ddir,
      "entropy-plots",
      dist_type,
      sprintf(
        "kl-divergence-cluster-ttest-%s-%s.csv",
        paste0(filter_timepoint, collapse = "-"),
        suffix
      )
    )
  )
  print(res_ttests, n = 100)

  res_ttests_detailed <- res_ttests |>
    dplyr::mutate(se_calculated = estimate / statistic) |>
    dplyr::relocate(se_calculated) |>
    dplyr::mutate(
      text_results = sprintf(
        "(DM=%.2f, 95%% UMCI=[%.2f,%.2f], t(%.2f)=%.2f, padj=%s)", # 95% unadjusted marginal confidence interval for the diff in means
        estimate,
        conf.low,
        conf.high,
        df,
        statistic,
        format.pval(p.adj, digits = 2, eps = 1e-6, scientific = TRUE)
      )
    ) |> # caps pvalues at 1e-6
    dplyr::relocate(text_results) |>
    dplyr::mutate(
      conf.low.check = estimate - se_calculated * qt(1 - (0.05 / 2), df = df),
      .after = conf.low
    ) |>
    dplyr::mutate(
      conf.high.check = estimate + se_calculated * qt(1 - (0.05 / 2), df = df),
      .after = conf.high
    ) |>
    dplyr::mutate(
      pcheck = 2 * pt(abs(statistic), df = df, lower.tail = FALSE)
    ) |> # two-sided unadjusted pvalue
    dplyr::mutate(
      p.adj.check = p.adjust(pcheck, method = "holm"),
      .after = "p.adj"
    ) |>
    dplyr::relocate(pcheck, .after = p)
  readr::write_csv(
    res_ttests_detailed |> as_tibble(),
    file.path(
      ddir,
      "entropy-plots",
      dist_type,
      sprintf(
        "kl-divergence-cluster-ttest-%s-%s-detailed.csv",
        paste0(filter_timepoint, collapse = "-"),
        suffix
      )
    )
  )
  print(file.path(
    ddir,
    "entropy-plots",
    dist_type,
    sprintf(
      "kl-divergence-cluster-ttest-%s-%s-detailed.csv",
      paste0(filter_timepoint, collapse = "-"),
      suffix
    )
  ))

  set.seed(840)

  # Plot significance stars? When just one tested comparison here, adjusted pvalue same as unadjusted.
  if (padj.signif.stars) {
    plt <- plt + ggpubr::stat_pvalue_manual(res_ttests, label = "p.adj.signif")
  }

  ggsave(
    file.path(
      odir,
      "entropy-plots",
      dist_type,
      sprintf(
        "kl-divergence-cluster-%s-%s.pdf",
        paste0(filter_timepoint, collapse = "-"),
        suffix
      )
    ),
    plt,
    width = 5,
    height = 5
  )
  return(plt)
}
```

### Plot normalized KL divergences for PalboR clusters (Figure 3H).

```{r}
plot_kl_divergences_clusters(
  infiles = c(
    "data/processed/01-barcoding/fishplots/column-swept-ppr/shannon-kl-divergence-fishplot-alltimepoints-column-swept-ppr_palbo-gradual_Parental-DOX_Parental-Palbo_Parental-Early-Middle-End_thresh0p001.rds",
    "data/processed/01-barcoding/fishplots/column-swept-ppr/shannon-kl-divergence-fishplot-alltimepoints-column-swept-ppr_palbo-gradual_Parental-ND_Parental-Palbo_Parental-Early-Middle-End_thresh0p001.rds"
  ),
  filter_timepoint = "End",
  suffix = "palbo-gradual-PalboR-samples"
)
```

### Plot normalized KL divergences for AbemaR clusters (Figure 3I).

```{r}
plot_kl_divergences_clusters(
  infiles = c(
    "data/processed/01-barcoding/fishplots/column-swept-ppr/shannon-kl-divergence-fishplot-alltimepoints-column-swept-ppr_abema-gradual_Parental-DOX_Parental-Abema_Parental-Early-Middle-End_thresh0p001.rds",
    "data/processed/01-barcoding/fishplots/column-swept-ppr/shannon-kl-divergence-fishplot-alltimepoints-column-swept-ppr_abema-gradual_Parental-ND_Parental-Abema_Parental-Early-Middle-End_thresh0p001.rds"
  ),
  filter_timepoint = "End",
  suffix = "abema-gradual-AbemaR-samples"
)
```

## Load in function to plot barcode overlaps of sample clusters.

```{r}
#' Plot contingency table of barcode cluster overlaps between two conditions,
#' including cases where barcodes were filtered out in one of the two samples.
#'
#' Clusters correspond exactly to clusters generated in [get_sample_feature_dist_bct_counts()]
#' and plotted in fishplots from [plot_bct_frac_fishplot()].
#'
#' @param infile_counts1 Path to filtered barcode counts for a condition generated from [get_sample_feature_dist_bct_counts()].
#' @param infile_counts2 Path to filtered barcode counts for a condition generated from [get_sample_feature_dist_bct_counts()].
#' @param label1 Label for x-axis corresponding to infile_counts1.
#' @param label2 Label for y-axis corresponding to infile_counts2.
#' @param dist_type Distance metric used to generate the feature dist and
#' sample dist matrices from [get_sample_feature_dist_bct_counts()].
#' @param output_width Plot width (in).
#' @param output_height Plot height (in).
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_barcode_comparison_scatter <- function(
  infile_counts1,
  infile_counts2,
  label1,
  label2,
  dist_type,
  output_width = 6,
  output_height = 6,
  ddir = "data/processed/01-barcoding",
  odir = "data/processed/01-barcoding"
) {
  clustdir <- file.path(odir, "cluster-overlaps")
  if (!dir.exists(clustdir)) {
    dir.create(clustdir)
  }
  outfile_pdf <- sprintf(
    file.path(clustdir, dist_type, "cluster-overlaps_%s_vs_%s.pdf"),
    gsub(".*feature-counts_(.*)\\.tsv", "\\1", infile_counts1),
    gsub(".*feature-counts_(.*)\\.tsv", "\\1", infile_counts2)
  )

  counts1 <- readr::read_tsv(infile_counts1, show_col_types = FALSE) |>
    dplyr::distinct(barcode, predicted_cluster) |>
    dplyr::group_by(predicted_cluster) |>
    dplyr::add_count() |>
    dplyr::mutate(Cluster = sprintf("Cluster %i (N=%i)", predicted_cluster, n))

  counts2 <- readr::read_tsv(infile_counts2, show_col_types = FALSE) |>
    dplyr::distinct(barcode, predicted_cluster) |>
    dplyr::group_by(predicted_cluster) |>
    dplyr::add_count() |>
    dplyr::mutate(Cluster = sprintf("Cluster %i (N=%i)", predicted_cluster, n))

  dat <- counts1 |>
    dplyr::full_join(counts2, by = "barcode") |>
    dplyr::mutate(n.x = case_when(is.na(n.x) ~ 0, .default = n.x)) |>
    dplyr::mutate(n.y = case_when(is.na(n.y) ~ 0, .default = n.y)) |>
    dplyr::mutate(
      Cluster.x = case_when(
        is.na(Cluster.x) ~ "Filtered out",
        .default = Cluster.x
      )
    ) |>
    dplyr::mutate(
      Cluster.y = case_when(
        is.na(Cluster.y) ~ "Filtered out",
        .default = Cluster.y
      )
    ) |>
    dplyr::group_by(Cluster.x, Cluster.y, .drop = FALSE) |>
    dplyr::count() |>
    dplyr::ungroup() |>
    tidyr::complete(Cluster.x, Cluster.y, fill = list(n = 0)) # fill in zeros where no barcodes overlap

  (plt <- dat |>
    ggplot2::ggplot(ggplot2::aes(
      x = Cluster.x,
      y = Cluster.y,
      label = n,
      fill = (Cluster.x == "Filtered out" | Cluster.y == "Filtered out")
    )) +
    ggplot2::geom_tile(color = "black") +
    ggplot2::geom_text() +
    ggplot2::labs(title = "Number of shared barcodes", x = label1, y = label2) +
    ggplot2::scale_y_discrete(limits = rev) +
    ggplot2::theme_bw() +
    ggplot2::theme(
      legend.position = "none",
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
    ) +
    ggplot2::coord_fixed(ratio = 1) +
    ggplot2::scale_fill_manual(values = c("white", "lightgray")))

  ggsave(
    outfile_pdf,
    plt,
    height = output_height,
    width = output_width,
    create.dir = TRUE
  )
  return(plt)
}
```

### Mut Vehicle CTR of PalboR and WT Vehicle CTR of PalboR (Supplementary Figure S3A).

```{r}
plot_barcode_comparison_scatter(
  infile_counts1 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_palbo-gradual_Parental-DOX_Parental-DMSO_Parental-Early-Middle-End_thresh0p001.tsv",
  infile_counts2 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_palbo-gradual_Parental-ND_Parental-DMSO_Parental-Early-Middle-End_thresh0p001.tsv",
  dist_type = "column-swept-ppr",
  label1 = "Mut Vehicle CTR of PalboR",
  label2 = "WT Vehicle CTR of PalboR"
)
```

### Mut Vehicle CTR of AbemaR and WT Vehicle CTR of AbemaR (Supplementary Figure S3B).

```{r}

plot_barcode_comparison_scatter(
  infile_counts1 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_abema-gradual_Parental-DOX_Parental-DMSO_Parental-Early-Middle-End_thresh0p001.tsv",
  infile_counts2 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_abema-gradual_Parental-ND_Parental-DMSO_Parental-Early-Middle-End_thresh0p001.tsv",
  dist_type = "column-swept-ppr",
  label1 = "Mut Vehicle CTR of AbemaR",
  label2 = "WT Vehicle CTR of AbemaR"
)
```

### Mut PalboR and WT PalboR (Supplementary Figure S3C).

```{r}

plot_barcode_comparison_scatter(
  infile_counts1 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_palbo-gradual_Parental-DOX_Parental-Palbo_Parental-Early-Middle-End_thresh0p001.tsv",
  infile_counts2 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_palbo-gradual_Parental-ND_Parental-Palbo_Parental-Early-Middle-End_thresh0p001.tsv",
  dist_type = "column-swept-ppr",
  label1 = "Mut PalboR escalating",
  label2 = "WT PalboR escalating"
)
```

### Mut AbemaR and WT AbemaR (Supplementary Figure S3D).

```{r}

plot_barcode_comparison_scatter(
  infile_counts1 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_abema-gradual_Parental-DOX_Parental-Abema_Parental-Early-Middle-End_thresh0p001.tsv",
  infile_counts2 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_abema-gradual_Parental-ND_Parental-Abema_Parental-Early-Middle-End_thresh0p001.tsv",
  dist_type = "column-swept-ppr",
  label1 = "Mut AbemaR",
  label2 = "WT AbemaR"
)
```

### WT PalboR and WT AbemaR (Supplementary Figure S3E).

```{r}
plot_barcode_comparison_scatter(
  infile_counts1 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_palbo-gradual_Parental-ND_Parental-Palbo_Parental-Early-Middle-End_thresh0p001.tsv",
  infile_counts2 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_abema-gradual_Parental-ND_Parental-Abema_Parental-Early-Middle-End_thresh0p001.tsv",
  dist_type = "column-swept-ppr",
  label1 = "WT PalboR escalating",
  label2 = "WT AbemaR"
)
```

### Mut PalboR and Mut AbemaR (Supplementary Figure S3F).

```{r}
plot_barcode_comparison_scatter(
  infile_counts1 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_palbo-gradual_Parental-DOX_Parental-Palbo_Parental-Early-Middle-End_thresh0p001.tsv",
  infile_counts2 = "data/processed/01-barcoding/feature-counts_column-swept-ppr_abema-gradual_Parental-DOX_Parental-Abema_Parental-Early-Middle-End_thresh0p001.tsv",
  dist_type = "column-swept-ppr",
  label1 = "Mut PalboR escalating",
  label2 = "Mut AbemaR"
)
```

## Cluster and plot fishplots of clonal prevalence for Parental, Vehicle controls of and Palbo gradual and AbemaR evolutionary replicates. 

### WT Vehicle CTR of PalboR (Supplementary Figure S3G).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "WT Vehicle CTR of PalboR",
  .condition = c("Parental", "ND"),
  .treatment = c("Parental", "DMSO"),
  dist_metric = "column-swept-ppr"
)
```

### Mut Vehicle CTR of PalboR (Supplementary Figure S3H).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "Mut Vehicle CTR of PalboR",
  .condition = c("Parental", "DOX"),
  .treatment = c("Parental", "DMSO"),
  dist_metric = "column-swept-ppr"
)
```

### WT Vehicle CTR of AbemaR (Supplementary Figure S3I).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "WT Vehicle CTR of AbemaR",
  .condition = c("Parental", "ND"),
  .treatment = c("Parental", "DMSO"),
  dist_metric = "column-swept-ppr"
)
```

### Mut Vehicle CTR of AbemaR (Supplementary Figure S3J).

```{r}
#| fig-width: 6
#| fig-height: 3
get_sample_feature_dist_bct_counts(
  infile_meta = "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
  infile_counts = "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
  freq_thresh = 0.001,
  plot_fishplot = TRUE,
  plt_title = "Mut Vehicle CTR of AbemaR",
  .condition = c("Parental", "DOX"),
  .treatment = c("Parental", "DMSO"),
  dist_metric = "column-swept-ppr"
)
```

## Print session info.

```{r}
sessionInfo()
```
