---
title: "In-vivo barcode analyses"
format: html
toc-depth: 4
---

## Preliminaries

Raw barcode count data are listed in Supplementary Tables S2, under the sheet names:

* Barcodes_libr_&_Parental_Palbo (Parental_3_passage_3_Count)
* Barcodes_Parental_Abema
* Barcodes_Palbo_study
* Barcodes_Abema_study
* Barcodes_clones
* Barcodes_tumors_&_recurrences

In the below code, we read in barcode counts and metadata from tables that compile this same data together into a long-format. 

### Load in libraries.

```{r}
library(tidyverse)
library(viridis)
library(RColorBrewer)
library(ggpubr)
library(rstatix)
library(philentropy)
```

### Create output data, plot folders.

```{r}
dir.create("data/raw/03-in-vivo-barcoding", recursive = TRUE) # input files
dir.create("data/processed/03-in-vivo-barcoding", recursive = TRUE) # output data, plots
```

### Sample structure of formatted barcode count, metadata tables.

```{r}
print(str(readr::read_tsv("data/raw/01-barcoding/bct-counts_abema-gradual.tsv")))
print(str(readr::read_tsv("data/raw/01-barcoding/bct-metadata_abema-gradual.tsv")))

print(str(readr::read_tsv("data/raw/01-barcoding/bct-counts_abema-gradual.tsv")))
print(str(readr::read_tsv("data/raw/01-barcoding/bct-metadata_abema-gradual.tsv")))

print(str(readr::read_tsv("data/raw/03-in-vivo-barcoding/bct-counts_mice-met-18.tsv")))
print(str(readr::read_tsv("data/raw/03-in-vivo-barcoding/bct-metadata_mice-met-18.tsv")))
```

## Load in function to calculate in-vivo barcode entropies

```{r}
#' Plot normalized KL divergences of barcode fraction distributions against a
#' discrete uniform normal for the specified mouse samples separately.
#'
#' Uses normalized natural log KL divergences as D_{KL}(P||Q) / log(|P|) where P
#' is the sample barcode fraction distribution and Q is the discrete uniform
#' distribution over the same number of total barcodes using philentropy::KL,
#' and |P| is the total number of discrete atoms in the discrete distributions P, Q.
#'
#' Note, normalized KL divergences are used here instead of Shannon entropies.
#'
#' @param infile_counts Path to tsv of long-format barcode counts.
#' @param infile_meta Path to tsv of long-format barcode metadata.
#' @param filter_treatment Vector of treatments to filter samples to. Some of 
#' c("primary-early", "primary-late", "recurrent", "metastasis-lung",
#' "metastasis-spleen", "metastasis-liver"). Keeps all if NULL.
#' @param filter_mouse_id Vector of mouse IDs to filter to. Keeps all if NULL.
#' @param include_palbo_gradual_dmso_only Logical. Include the Vehicle sample 
#' from the palbo-gradual experiment only?
#' @param padj.signif.stars Logical. Plot significance stars?
#' @param freq_thresh Minimum frequency threshold (0-1) to filter out low-frequency barcodes.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param data_dir Processed data directory.
#' @param output_dir Processed output directory.
plot_barcode_fraction_KL_divergences_against_discrete_uniform_mouse_met_18_separate_samples <- function(
  infile_counts,
  infile_meta,
  filter_treatment = c(
    "primary-early",
    "primary-late",
    "recurrent",
    "metastasis-lung",
    "metastasis-spleen",
    "metastasis-liver"
  ),
  filter_mouse_id = NULL,
  include_palbo_gradual_dmso_only = FALSE,
  # padj.signif.stars = FALSE,
  freq_thresh = NULL,
  width = 7,
  height = 5,
  data_dir = "data/processed/03-in-vivo-barcoding",
  output_dir = "data/processed/03-in-vivo-barcoding"
) {
  counts <- purrr::map_dfr(
    infile_counts,
    readr::read_tsv,
    show_col_types = FALSE
  )
  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    dplyr::mutate(
      treatment = dplyr::case_when(
        timepoint == "early" & tumor_type == "primary" ~ "primary-early",
        timepoint == "late" & tumor_type == "primary" ~ "primary-late",
        tumor_type == "recurrent" ~ "recurrent",
        tumor_type == "metastasis" &
          grepl("lung", biorepl_id) ~ "metastasis-lung",
        tumor_type == "metastasis" &
          grepl("spleen", biorepl_id) ~ "metastasis-spleen",
        tumor_type == "metastasis" &
          grepl("liver", biorepl_id) ~ "metastasis-liver"
      )
    ) |>
    dplyr::mutate(
      treatment = factor(
        treatment,
        levels = c(
          "primary-early",
          "primary-late",
          "recurrent",
          "metastasis-lung",
          "metastasis-spleen",
          "metastasis-liver"
        )
      )
    ) |>
    dplyr::mutate(graph_name = treatment) |>
    dplyr::mutate(graph_color = treatment)
  expt <- gsub(".*bct-metadata_(.*)\\.tsv", "\\1", infile_meta)

  meta <- meta %>%
    dplyr::mutate(graph_color = fct_drop(graph_color))

  if (!is.null(filter_treatment)) {
    meta <- meta |> dplyr::filter(treatment %in% filter_treatment)
  }
  if (!is.null(filter_mouse_id)) {
    meta <- meta |> dplyr::filter(mouse_id %in% filter_mouse_id)
  }

  counts_meta <- counts %>%
    dplyr::inner_join(meta, by = "sample_name") %>%
    dplyr::select(-fraction) %>% # drop existing fraction and recalculate
    dplyr::group_by(sample_name) %>%
    dplyr::mutate(fraction = count / sum(count)) %>%
    dplyr::ungroup()

  if (!is.null(filter_treatment)) {
    counts_meta <- counts_meta |> dplyr::filter(treatment %in% filter_treatment)
  }
  if (!is.null(filter_mouse_id)) {
    counts_meta <- counts_meta |> dplyr::filter(mouse_id %in% filter_mouse_id)
  }

  if (!is.null(freq_thresh)) {
    counts_meta <- counts_meta %>%
      dplyr::group_by(barcode) %>% # Filter out any barcodes that are below a certain threshold across all observations
      dplyr::filter(max(fraction) >= freq_thresh) %>%
      dplyr::ungroup()
  }

  # Test if proper subset of samples is selected
  if (!is.null(filter_treatment)) {
    stopifnot(all(counts_meta$treatment %in% filter_treatment))
  }
  if (!is.null(filter_mouse_id)) {
    stopifnot(all(counts_meta$mouse_id %in% filter_mouse_id))
  }

  counts_meta_entropy <- counts_meta %>%
    dplyr::group_by(
      sample_name,
      graph_name,
      graph_color,
      biorepl_id,
      mouse_id,
      treatment
    ) %>%
    dplyr::summarize(
      n_unique_barcodes = dplyr::n(),
      n_count_barcodes = sum(count),
      js_divergence = philentropy::JSD(rbind(
        fraction,
        (rep(1, length(fraction)) / length(fraction))
      )),
      kl_divergence = philentropy::KL(
        rbind(fraction, (rep(1, length(fraction)) / length(fraction))),
        unit = "log"
      ) /
        log(length(fraction)),
      n_unique_barcodes_check = length(fraction)
    )

  i <- "kl_divergence"
  yla = "Normalized KL divergence of barcode distribution\nagainst discrete uniform distribution"

  outfile <- sprintf(
    file.path(
      output_dir,
      "barcode-fraction-KL-divergences-scatterplot-by-sample_%s_%s_%s_%s.pdf"
    ),
    paste0(expt, collapse = "-"),
    paste0(filter_treatment, collapse = "-"),
    paste0(filter_mouse_id, collapse = "-"),
    i
  )
  outfile_jsd <- sprintf(
    file.path(
      output_dir,
      "barcode-fraction-JS-divergences-scatterplot-by-sample_%s_%s_%s_%s.pdf"
    ),
    paste0(expt, collapse = "-"),
    paste0(filter_treatment, collapse = "-"),
    paste0(filter_mouse_id, collapse = "-"),
    "js_divergence"
  )

  outfile_csv <- sprintf(
    file.path(
      data_dir,
      "barcode-fraction-KL-divergences-scatterplot-by-sample_%s_%s_%s_%s.csv"
    ),
    paste0(expt, collapse = "-"),
    paste0(filter_treatment, collapse = "-"),
    paste0(filter_mouse_id, collapse = "-"),
    i
  )

  # Write out the table of the normalized KL divergences
  readr::write_csv(counts_meta_entropy, outfile_csv)
  outfile_ttests <- sprintf(
    file.path(
      data_dir,
      "barcode-fraction-KL-divergences-scatterplot-by-sample_%s_%s_%s_%s.csv"
    ),
    paste0(expt, collapse = "-"),
    paste0(filter_treatment, collapse = "-"),
    paste0(filter_mouse_id, collapse = "-"),
    i
  )
  print(outfile)

  set.seed(840)
  plt <- counts_meta_entropy %>%
    dplyr::ungroup() |>
    dplyr::mutate(
      biorepl_id = forcats::fct_recode(
        biorepl_id,
        `816_Lung` = "816_lung_met",
        `817_Spleen` = "817_spleen"
      )
    ) |>
    dplyr::mutate(
      biorepl_id = forcats::fct_relevel(
        biorepl_id,
        "816_P-E_1",
        "816_P-E_2",
        "816_Rec_1",
        "816_Rec_2",
        "816_Lung",

        "817_P-L",
        "817_regr_1",
        "817_regr_2",
        "817_Spleen",

        "818_P-E_1",
        "818_P-E_2",
        "818_rec",

        "821_P-L",
        "821_regr_1",
        "821_regr_2"
      )
    ) |>

    dplyr::mutate(`Mouse ID` = factor(mouse_id)) |>
    dplyr::rename(Type = "graph_color") |>
    ggplot2::ggplot(ggplot2::aes(x = biorepl_id, y = !!sym(i))) +
    ggplot2::geom_point(
      ggplot2::aes(color = `Mouse ID`, shape = Type),
      size = 3
    ) +
    ggplot2::theme_bw() +
    ggplot2::ylim(0, 1) +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 45, vjust = 1, hjust = 1)
    ) +
    ggplot2::labs(
      x = NULL,
      y = yla,
      title = sprintf(
        "Barcode fraction KL divergences: %s",
        paste0(filter_mouse_id, collapse = "-")
      )
    )

  ggplot2::ggsave(outfile, plt, height = height, width = width)

  return(plt)
}
```

## Plot barcode entropies for in-vivo local recurrences and respective primary-early or primery-late tumors (Figure 7H).

```{r}
plot_barcode_fraction_KL_divergences_against_discrete_uniform_mouse_met_18_separate_samples(
  infile_counts = c("data/raw/03-in-vivo-barcoding/bct-counts_mice-met-18.tsv"),
  infile_meta = c("data/raw/03-in-vivo-barcoding/bct-metadata_mice-met-18.tsv"),
  filter_mouse_id = c(816, 817, 818, 821),
  filter_treatment = c("primary-early", "primary-late", "recurrent"),
  width = 6,
  freq_thresh = NULL
)
```

## Load in function to calculate and compare barcode entropies across in-vivo samples.

```{r}
#' Plot normalized KL divergences of barcode fraction distributions against a
#' discrete uniform normal for the specified mouse met 18 samples, run one way 
#' ANOVA and pairwise t-tests comparing sample groups.
#'
#' Uses normalized natural log KL divergences as D_{KL}(P||Q) / log(|P|) where P
#' is the sample barcode fraction distribution and Q is the discrete uniform
#' distribution over the same number of total barcodes using philentropy::KL,
#' and |P| is the total number of discrete atoms in the discrete distributions P, Q.
#'
#' Note, normalized KL divergences are used here instead of Shannon entropies.
#'
#' @param infile_counts Path to barcode counts.
#' @param infile_meta Path to barcode metadata.
#' @param plt_var_x Variable to plot on x axis and perform t tests on.
#' @param plt_t_test_var1 Variable level for t test numerator. Tests all pairwise comparisons if NULL.
#' @param plt_t_test_var2 Variable level for t test denominator. Tests all pairwise comparisons if NULL.
#' @param filter_treatment Vector of treatments to filter samples to. Some of 
#' c("primary-early", "primary-late", "recurrent", "metastasis-lung",
#' "metastasis-spleen", "metastasis-liver"). Keeps all if NULL.
#' @param filter_mouse_id Vector of mouse IDs to filter to. Keeps all if NULL.
#' @param include_palbo_gradual_dmso_only Logical. Include the Vehicle sample 
#' from the palbo-gradual experiment only?
#' @param padj.signif.stars Logical. Plot significance stars?
#' @param freq_thresh Minimum frequency threshold (0-1) to filter out low-frequency barcodes.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param data_dir Processed data directory.
#' @param output_dir Processed output directory
plot_barcode_fraction_KL_divergences_against_discrete_uniform_mouse_met_18 <- function(
  infile_counts,
  infile_meta,
  plt_var_x,
  plt_t_test_var1 = NULL,
  plt_t_test_var2 = NULL,
  filter_treatment = c(
    "primary-early",
    "primary-late",
    "recurrent",
    "metastasis-lung",
    "metastasis-spleen",
    "metastasis-liver"
  ),
  filter_mouse_id = NULL,
  include_palbo_gradual_dmso_only = FALSE,
  padj.signif.stars = TRUE,
  freq_thresh = NULL,
  width = 7,
  height = 5,
  data_dir = "data/processed/03-in-vivo-barcoding",
  output_dir = "data/processed/03-in-vivo-barcoding"
) {
  counts <- purrr::map_dfr(
    infile_counts,
    readr::read_tsv,
    show_col_types = FALSE
  )
  meta <- purrr::map_dfr(
    infile_meta,
    readr::read_tsv,
    show_col_types = FALSE
  ) %>%
    mutate(
      treatment = case_when(
        timepoint == "early" & tumor_type == "primary" ~ "primary-early",
        timepoint == "late" & tumor_type == "primary" ~ "primary-late",
        tumor_type == "recurrent" ~ "recurrent",
        tumor_type == "metastasis" &
          grepl("lung", biorepl_id) ~ "metastasis-lung",
        tumor_type == "metastasis" &
          grepl("spleen", biorepl_id) ~ "metastasis-spleen",
        tumor_type == "metastasis" &
          grepl("liver", biorepl_id) ~ "metastasis-liver"
      )
    ) |>
    mutate(
      treatment = factor(
        treatment,
        levels = c(
          "primary-early",
          "primary-late",
          "recurrent",
          "metastasis-lung",
          "metastasis-spleen",
          "metastasis-liver"
        )
      )
    ) |>
    mutate(graph_name = treatment) |>
    mutate(graph_color = treatment)
  expt <- gsub(".*bct-metadata_(.*)\\.tsv", "\\1", infile_meta)

  meta <- meta %>%
    mutate(graph_color = fct_drop(graph_color))

  if (!is.null(filter_treatment)) {
    meta <- meta |> dplyr::filter(treatment %in% filter_treatment)
  }
  if (!is.null(filter_mouse_id)) {
    meta <- meta |> dplyr::filter(mouse_id %in% filter_mouse_id)
  }

  counts_meta <- counts %>%
    dplyr::inner_join(meta, by = "sample_name") %>%
    dplyr::select(-fraction) %>% # drop existing fraction and recalculate
    dplyr::group_by(sample_name) %>%
    dplyr::mutate(fraction = count / sum(count)) %>%
    dplyr::ungroup()

  if (!is.null(filter_treatment)) {
    counts_meta <- counts_meta |> dplyr::filter(treatment %in% filter_treatment)
  }
  if (!is.null(filter_mouse_id)) {
    counts_meta <- counts_meta |> dplyr::filter(mouse_id %in% filter_mouse_id)
  }

  if (!is.null(freq_thresh)) {
    counts_meta <- counts_meta %>%
      dplyr::group_by(barcode) %>% # Filter out any barcodes that are below a certain threshold across all observations
      dplyr::filter(max(fraction) >= freq_thresh) %>%
      dplyr::ungroup()
  }

  # Test if proper subset of samples is selected
  if (!is.null(filter_treatment)) {
    stopifnot(all(counts_meta$treatment %in% filter_treatment))
  }
  if (!is.null(filter_mouse_id)) {
    stopifnot(all(counts_meta$mouse_id %in% filter_mouse_id))
  }

  counts_meta_entropy <- counts_meta %>%
    dplyr::group_by(sample_name, graph_name, graph_color, treatment) %>%
    dplyr::summarize(
      n_unique_barcodes = dplyr::n(),
      n_count_barcodes = sum(count),
      kl_divergence = philentropy::KL(
        rbind(fraction, (rep(1, length(fraction)) / length(fraction))),
        unit = "log"
      ) /
        log(length(fraction)),
      n_unique_barcodes_check = length(fraction)
    )

  i <- "kl_divergence"
  yla = "Normalized KL divergence of barcode distribution\nagainst discrete uniform distribution"

  outfile <- sprintf(
    file.path(output_dir, "barcode-fraction-KL-divergences_%s_%s_%s_%s.pdf"),
    paste0(expt, collapse = "-"),
    paste0(filter_treatment, collapse = "-"),
    paste0(filter_mouse_id, collapse = "-"),
    i
  )
  outfile_ttests <- sprintf(
    file.path(data_dir, "barcode-fraction-KL-divergences_%s_%s_%s_%s.csv"),
    paste0(expt, collapse = "-"),
    paste0(filter_treatment, collapse = "-"),
    paste0(filter_mouse_id, collapse = "-"),
    i
  )
  print(outfile)

  # Run an ANOVA across all categories.
  # Note, should really filter out any n=1 groups before we do this.
  anova_res <- anova(lm(kl_divergence ~ graph_name, data = counts_meta_entropy))
  anova_subtitle <- sprintf(
    "(F(%i,%i)=%.2f; p=%.2e)",
    anova_res$Df[[1]],
    anova_res$Df[[2]],
    anova_res$`F value`[[1]],
    anova_res$`Pr(>F)`[[1]]
  )

  # Avoids cases where we don't have enough samples to run a test.
  if (nrow(counts_meta_entropy) < 4) {
    return()
  }

  res_ttests <- counts_meta_entropy |>
    ungroup() |> # need to ungroup!
    filter(treatment != "metastasis-lung") |> # only one sample in this group
    dplyr::mutate(graph_name = forcats::fct_drop(graph_name)) |> # drop unused factor levels or t test tries to test empty groups
    dplyr::group_by(graph_name) |>
    dplyr::filter(n() >= 2) |> # drop groups with only one observation as we can't test this.
    dplyr::ungroup() |>
    dplyr::mutate(graph_name = forcats::fct_drop(graph_name)) |> # drop unused factor levels or t test tries to test empty groups
    rstatix::pairwise_t_test(
      formula = kl_divergence ~ graph_name,
      detailed = TRUE,
      pool.sd = FALSE,
      var.equal = FALSE
    ) |>
    rstatix::add_y_position()

  readr::write_csv(
    res_ttests |>
      mutate(
        group1 = gsub("\n", " ", group1),
        group2 = gsub("\n", " ", group2)
      ) |>
      as_tibble(),
    outfile_ttests
  )
  print(res_ttests, n = 100)

  # Include a format string to write results intext.
  res_ttests_detailed <- res_ttests |>
    dplyr::mutate(se_calculated = estimate / statistic) |> # back calculate what SE of the diff in means is from the t statistic
    dplyr::relocate(se_calculated) |>
    dplyr::mutate(
      text_results = sprintf(
        "(DM=%.2f, 95%% UMCI=[%.2f,%.2f], t(%.2f)=%.2f, padj=%s)", # 95% unadjusted marginal confidence interval for the diff in means
        estimate,
        conf.low,
        conf.high,
        df,
        statistic,
        format.pval(p.adj, digits = 2, eps = 1e-6, scientific = TRUE)
      )
    ) |> # caps pvalues at 1e-6
    dplyr::relocate(text_results) |>
    dplyr::mutate(
      conf.low.check = estimate - se_calculated * qt(1 - (0.05 / 2), df = df),
      .after = conf.low
    ) |>
    dplyr::mutate(
      conf.high.check = estimate + se_calculated * qt(1 - (0.05 / 2), df = df),
      .after = conf.high
    ) |>
    dplyr::mutate(
      pcheck = 2 * pt(abs(statistic), df = df, lower.tail = FALSE)
    ) |> # two-sided unadjusted pvalue
    dplyr::mutate(
      p.adj.check = p.adjust(pcheck, method = "holm"),
      .after = "p.adj"
    ) |>
    dplyr::relocate(pcheck, .after = p)

  readr::write_csv(
    res_ttests_detailed |>
      mutate(
        group1 = gsub("\n", " ", group1),
        group2 = gsub("\n", " ", group2)
      ) |>
      as_tibble(),
    gsub("\\.csv", "-detailed.csv", outfile_ttests)
  )

  set.seed(840)
  plt <- counts_meta_entropy %>%
    ggplot2::ggplot(ggplot2::aes(x = graph_name, y = !!sym(i))) +
    geom_boxplot(outlier.shape = NA) +
    ggplot2::geom_point(
      ggplot2::aes(col = graph_color),
      position = ggplot2::position_dodge2(width = .5)
    ) +
    ggplot2::theme_bw() +
    ggplot2::theme(legend.position = "none") +
    ggplot2::labs(
      x = NULL,
      y = yla,
      title = "Barcode fraction distribution KL divergences",
      subtitle = anova_subtitle
    ) +
    ggplot2::ylim(0, NA)

  if (padj.signif.stars) {
    plt <- plt + ggpubr::stat_pvalue_manual(res_ttests, label = "p.adj.signif")
  }

  # Plot and show just a single t test results if we specify so.
  if (!is.null(plt_t_test_var1) & !is.null(plt_t_test_var2)) {
    a <- counts_meta_entropy %>%
      dplyr::filter(!!sym(plt_var_x) == plt_t_test_var1) %>%
      dplyr::pull(!!sym(i))
    b <- counts_meta_entropy %>%
      dplyr::filter(!!sym(plt_var_x) == plt_t_test_var2) %>%
      dplyr::pull(!!sym(i))

    print(
      counts_meta_entropy %>% dplyr::filter(!!sym(plt_var_x) == plt_t_test_var1)
    )
    print(
      counts_meta_entropy %>% dplyr::filter(!!sym(plt_var_x) == plt_t_test_var2)
    )

    ttest_summary <- t.test(a, b)

    outfile <- sprintf(
      file.path(
        output_dir,
        "barcode-fraction-KL-divergences_%s_%s_%s_%s_%s_%s_vs_%s_%s.pdf"
      ),
      paste0(expt, collapse = "-"),
      paste0(.timepoint, collapse = "-"),
      paste0(.condition, collapse = "-"),
      paste0(filter_treatment, collapse = "-"),
      paste0(filter_mouse_id, collapse = "-"),
      plt_t_test_var1,
      plt_t_test_var2,
      i
    )

    plt <- plt +
      ggplot2::labs(
        subtitle = sprintf(
          "t-test: %s vs %s; pval = %s",
          plt_t_test_var1,
          plt_t_test_var2,
          signif(ttest_summary$p.value, 4)
        )
      )
  }

  ggplot2::ggsave(outfile, plt, height = height, width = width)
  return(plt)
}
```

## Plot barcode entropies across in-vivo samples (Figure 7K).

```{r}
plot_barcode_fraction_KL_divergences_against_discrete_uniform_mouse_met_18(
  infile_counts = c("data/raw/03-in-vivo-barcoding/bct-counts_mice-met-18.tsv"),
  infile_meta = c("data/raw/03-in-vivo-barcoding/bct-metadata_mice-met-18.tsv"),
  plt_var_x = "treatment",
  padj.signif.stars = TRUE,
  width = 8,
  freq_thresh = NULL
)
```

## Load in function to calculate and plot barcode overlaps between in-vivo and cell line samples.

```{r}
#' Plot contingency table plot of shared barcodes above some fraction across
#' specified samples and conditions from mouse and cell line samples.
#'
#' Also write out versions that are just mets (rows) x cell line (column).
#'
#' @param infile_meta_mm18 Path to tsv of mouse barcode metadata.
#' @param infile_meta_cellline = Vector of paths to tsvs of cell line barcode data for Abema and Palbo.
#' @param infile_counts Vector of paths to tsvs of mouse and cell line barcode counts.
#' @param filter_mouse_id Vector of mouse IDs to filter to. Keeps all if NULL.
#' @param filter_condition Vector of mouse conditions to filter to. Keeps all if left as NULL.
#' @param filter_treatment Vector of mouse treatments to filter to. Keeps all if left as NULL.
#' @param filter_timepoint Vector of mouse timepoints to filter to. Keeps all if left as NULL.
#' @param filter_experiment Vector of mouse experiments to filter to. Keeps all if left as NULL.
#' @param filter_biorepl_id Vector of mouse bioreplicate ids to filter to. Keeps all if left as NULL.
#' @param filter_exclude_sid Vector of mouse sids (sample ids) to exclude from analysis. Keeps all if left as NULL.
#' 
#' @param filter_condition_cellline Vector of cell line conditions to filter to. Keeps all if left as NULL.
#' @param filter_treatment_cellline Vector of cell line treatments to filter to. Keeps all if left as NULL.
#' @param filter_timepoint_cellline Vector of cell line timepoints to filter to. Keeps all if left as NULL.
#' @param filter_experiment_cellline Vector of cell line experiments to filter to. Keeps all if left as NULL.
#' @param filter_exclude_sid_cellline Vector of cell line sids (sample ids) to exclude from analysis. Keeps all if left as NULL.
#' 
#' @param freq_thresh Minimum frequency threshold (0-1) to filter out low-frequency barcodes.
#' @param include_palbo_gradual_dmso_only Logical. Include the Vehicle sample 
#' from the palbo-gradual experiment only?
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param ddir Processed data directory.
#' @param odir Processed output directory
plot_bct_barcode_overlaps_cellline_mouse_met_18 <- function(
  infile_meta_mm18 = "data/raw/03-in-vivo-barcoding/bct-metadata_mice-met-18.tsv",
  infile_meta_cellline = c(
    "data/raw/01-barcoding/bct-metadata_abema-gradual.tsv",
    "data/raw/01-barcoding/bct-metadata_palbo-gradual.tsv"
  ),
  infile_counts = c(
    "data/raw/03-in-vivo-barcoding/bct-counts_mice-met-18.tsv",
    "data/raw/01-barcoding/bct-counts_abema-gradual.tsv",
    "data/raw/01-barcoding/bct-counts_palbo-gradual.tsv"
  ),
  filter_mouse_id = NULL,
  filter_condition = NULL,
  filter_treatment = NULL,
  filter_timepoint = NULL,
  filter_experiment = NULL,
  filter_biorepl_id = NULL,
  filter_exclude_sid = NULL,

  filter_condition_cellline = NULL,
  filter_treatment_cellline = NULL,
  filter_timepoint_cellline = NULL,
  filter_experiment_cellline = NULL,
  filter_exclude_sid_cellline = NULL,

  freq_thresh = 0.01,
  include_palbo_gradual_dmso_only,
  width = 10,
  height = 6,
  ddir = "data/processed/03-in-vivo-barcoding",
  odir = "data/processed/03-in-vivo-barcoding"
) {
  meta_mm18 <- readr::read_tsv(infile_meta_mm18) |>
    dplyr::arrange(mouse_id)

  meta_cellline <- readr::read_tsv(infile_meta_cellline) |>
    dplyr::mutate(
      condition = forcats::fct_recode(condition, Mut = "DOX", WT = "ND")
    ) |>
    dplyr::mutate(
      treatment = forcats::fct_recode(
        treatment,
        PalboR = "Palbo",
        AbemaR = "Abema",
        Vehicle = "DMSO"
      )
    ) |>
    dplyr::mutate(
      experiment = forcats::fct_recode(
        experiment,
        ag = "abema-gradual",
        pg = "palbo-gradual"
      )
    ) |>
    dplyr::mutate(condition = forcats::fct_relevel(condition, "WT", "Mut")) |>
    dplyr::mutate(
      treatment = forcats::fct_relevel(
        treatment,
        "Parental",
        "PalboR",
        "AbemaR",
        "Vehicle"
      )
    ) |>
    dplyr::mutate(experiment = forcats::fct_relevel(experiment, "pg", "ag")) |>
    dplyr::mutate(
      timepoint = forcats::fct_relevel(
        timepoint,
        "Parental",
        "Early",
        "Middle",
        "End"
      )
    ) |>
    dplyr::arrange(experiment, treatment, condition, timepoint, replicate_id) |> # orders the samples correctly
    dplyr::mutate(
      biorepl_id = sprintf(
        "%s_%s_%s_%s_%s",
        condition,
        treatment,
        timepoint,
        replicate_id,
        experiment
      )
    )

  counts <- readr::read_tsv(infile_counts) |>
    group_by(sample_name) |>
    mutate(fraction = count / sum(count)) |> # recalculate fraction per sample based on counts
    ungroup()

  # filter_mouse_id = NULL,
  if (!is.null(filter_mouse_id)) {
    meta_mm18 <- meta_mm18 |> dplyr::filter(mouse_id %in% filter_mouse_id)
  }
  if (!is.null(filter_condition)) {
    meta_mm18 <- meta_mm18 |> dplyr::filter(condition %in% filter_condition)
  }
  if (!is.null(filter_treatment)) {
    meta_mm18 <- meta_mm18 |> dplyr::filter(treatment %in% filter_treatment)
  }
  if (!is.null(filter_timepoint)) {
    meta_mm18 <- meta_mm18 |> dplyr::filter(timepoint %in% filter_timepoint)
  }
  if (!is.null(filter_experiment)) {
    meta_mm18 <- meta_mm18 |> dplyr::filter(experiment %in% filter_experiment)
  }
  if (!is.null(filter_biorepl_id)) {
    meta_mm18 <- meta_mm18 |> dplyr::filter(biorepl_id %in% filter_biorepl_id)
  }
  if (!is.null(filter_exclude_sid)) {
    meta_mm18 <- meta_mm18 |>
      dplyr::filter(!(sample_name %in% filter_exclude_sid))
  }

  # Check that filtering was indeed performed.
  if (!is.null(filter_mouse_id)) {
    stopifnot(all(meta_mm18$mouse_id %in% filter_mouse_id))
  }
  if (!is.null(filter_condition)) {
    stopifnot(all(meta_mm18$condition %in% filter_condition))
  }
  if (!is.null(filter_treatment)) {
    stopifnot(all(meta_mm18$treatment %in% filter_treatment))
  }
  if (!is.null(filter_timepoint)) {
    stopifnot(all(meta_mm18$timepoint %in% filter_timepoint))
  }
  if (!is.null(filter_experiment)) {
    stopifnot(all(meta_mm18$experiment %in% filter_experiment))
  }
  if (!is.null(filter_exclude_sid)) {
    stopifnot(all(!(meta_mm18$sample_name %in% filter_exclude_sid)))
  }

  if (!is.null(filter_condition_cellline)) {
    meta_cellline <- meta_cellline |>
      dplyr::filter(condition %in% filter_condition_cellline)
  }
  if (!is.null(filter_treatment_cellline)) {
    meta_cellline <- meta_cellline |>
      dplyr::filter(treatment %in% filter_treatment_cellline)
  }
  if (!is.null(filter_timepoint_cellline)) {
    meta_cellline <- meta_cellline |>
      dplyr::filter(timepoint %in% filter_timepoint_cellline)
  }
  if (!is.null(filter_experiment_cellline)) {
    meta_cellline <- meta_cellline |>
      dplyr::filter(experiment %in% filter_experiment_cellline)
  }
  if (!is.null(filter_exclude_sid_cellline)) {
    meta_cellline <- meta_cellline |>
      dplyr::filter(!(sample_name %in% filter_exclude_sid_cellline))
  }

  # Check that filtering was indeed performed.
  if (!is.null(filter_condition)) {
    stopifnot(all(meta_cellline$condition %in% filter_condition))
  }
  if (!is.null(filter_treatment)) {
    stopifnot(all(meta_cellline$treatment %in% filter_treatment))
  }
  if (!is.null(filter_timepoint)) {
    stopifnot(all(meta_cellline$timepoint %in% filter_timepoint))
  }
  if (!is.null(filter_experiment)) {
    stopifnot(all(meta_cellline$experiment %in% filter_experiment))
  }
  if (!is.null(filter_exclude_sid)) {
    stopifnot(all(!(meta_cellline$sample_name %in% filter_exclude_sid)))
  }

  outfile <- file.path(
    odir,
    sprintf(
      "contingency-plot-cellline-mouse-met-18-freqthresh%0.2e-%s-%s-%s-%s-%s-%s-%s-%s-%s-%s.pdf",
      freq_thresh,
      paste0(filter_mouse_id, collapse = "-"),
      paste0(filter_condition, collapse = "-"),
      paste0(filter_treatment, collapse = "-"),
      paste0(filter_timepoint, collapse = "-"),
      paste0(filter_experiment, collapse = "-"),
      paste0(filter_biorepl_id, collapse = "-"),
      paste0(filter_condition_cellline, collapse = "-"),
      paste0(filter_treatment_cellline, collapse = "-"),
      paste0(filter_timepoint_cellline, collapse = "-"),
      paste0(filter_experiment_cellline, collapse = "-")
    )
  )

  outfile_subsetted_col_cumfrac <- file.path(
    odir,
    sprintf(
      "contingency-plot-subsetted-column-cumfrac-cellline-mouse-met-18-freqthresh%0.2e-%s-%s-%s-%s-%s-%s-%s-%s-%s-%s.pdf",
      freq_thresh,
      paste0(filter_mouse_id, collapse = "-"),
      paste0(filter_condition, collapse = "-"),
      paste0(filter_treatment, collapse = "-"),
      paste0(filter_timepoint, collapse = "-"),
      paste0(filter_experiment, collapse = "-"),
      paste0(filter_biorepl_id, collapse = "-"),
      paste0(filter_condition_cellline, collapse = "-"),
      paste0(filter_treatment_cellline, collapse = "-"),
      paste0(filter_timepoint_cellline, collapse = "-"),
      paste0(filter_experiment_cellline, collapse = "-")
    )
  )

  outfile_subsetted_row_cumfrac <- file.path(
    odir,
    sprintf(
      "contingency-plot-subsetted-row-cumfrac-cellline-mouse-met-18-freqthresh%0.2e-%s-%s-%s-%s-%s-%s-%s-%s-%s-%s.pdf",
      freq_thresh,
      paste0(filter_mouse_id, collapse = "-"),
      paste0(filter_condition, collapse = "-"),
      paste0(filter_treatment, collapse = "-"),
      paste0(filter_timepoint, collapse = "-"),
      paste0(filter_experiment, collapse = "-"),
      paste0(filter_biorepl_id, collapse = "-"),
      paste0(filter_condition_cellline, collapse = "-"),
      paste0(filter_treatment_cellline, collapse = "-"),
      paste0(filter_timepoint_cellline, collapse = "-"),
      paste0(filter_experiment_cellline, collapse = "-")
    )
  )

  meta <- rbind(
    meta_cellline |> dplyr::select(sample_name, biorepl_id),
    meta_mm18 |> dplyr::select(sample_name, biorepl_id)
  )

  counts <- counts |> # Filter out any barcodes that are below a certain threshold across all observations
    dplyr::filter(fraction >= freq_thresh) %>%
    dplyr::filter(sample_name %in% meta$sample_name)

  stopifnot(all(counts$sample_name %in% meta$sample_name))

  meta <- meta |>
    dplyr::mutate(graph_name = biorepl_id) |>
    dplyr::mutate(
      graph_name = forcats::fct_relevel(
        graph_name,
        meta_mm18$biorepl_id,
        meta_cellline$biorepl_id
      )
    )

  (counts_ch <- counts |> dplyr::inner_join(meta, by = "sample_name"))

  ct <- counts_ch |>
    dplyr::select(biorepl_id, barcode, fraction) |>
    tidyr::pivot_wider(
      id_cols = barcode,
      names_from = biorepl_id,
      values_from = fraction,
      values_fill = 0
    ) |>
    tibble::column_to_rownames("barcode")

  d_overlaps <- tidyr::expand_grid(s1 = colnames(ct), s2 = colnames(ct)) |>
    purrr::pmap_dfr(\(s1, s2) {
      data.frame(
        s1 = s1,
        s2 = s2,
        n_shared_barcodes = sum((ct[, s1] > 0) & (ct[, s2] > 0)),
        cum_frac_s1 = sum(ct[, s1] * ((ct[, s1] > 0) & (ct[, s2] > 0))),
        cum_frac_s2 = sum(ct[, s2] * ((ct[, s1] > 0) & (ct[, s2] > 0)))
      )
    }) |>
    dplyr::mutate(
      s1 = forcats::fct_relevel(
        s1,
        meta_mm18$biorepl_id,
        meta_cellline$biorepl_id
      )
    ) |> # need to relevel after pivoting
    dplyr::mutate(
      s2 = forcats::fct_relevel(
        s2,
        meta_mm18$biorepl_id,
        meta_cellline$biorepl_id
      )
    ) # need to relevel after pivoting

  plt <- d_overlaps |>
    ggplot2::ggplot(ggplot2::aes(
      x = s1,
      y = s2,
      fill = cum_frac_s1,
      label = ifelse(n_shared_barcodes > 0, n_shared_barcodes, NA)
    )) +
    ggplot2::geom_tile(color = "black") +
    ggplot2::geom_text(
      ggplot2::aes(color = cum_frac_s1 > 0.5),
      size = 2,
      show.legend = FALSE
    ) +
    ggplot2::theme_bw() +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      fill = "Cumulative fraction\nof shared barcodes\nwithin column sample"
    ) +
    ggplot2::scale_y_discrete(limits = rev) +
    ggplot2::theme(
      text = ggplot2::element_text(size = 7),
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
    ) +
    ggplot2::scale_fill_gradient(
      low = "white",
      high = "blue",
      limits = c(0, 1)
    ) +
    ggplot2::scale_color_manual(values = c("black", "white")) +
    ggplot2::coord_fixed(1)

  if (!is.null(filter_biorepl_id)) {
    # mets x cell line, fill is column cumulative fraction
    plt_col_met <- d_overlaps |>
      dplyr::filter(s2 %in% filter_biorepl_id) |>
      dplyr::filter(!(s1 %in% filter_biorepl_id)) |>
      ggplot2::ggplot(ggplot2::aes(
        x = s1,
        y = s2,
        fill = cum_frac_s1,
        label = ifelse(n_shared_barcodes > 0, n_shared_barcodes, NA)
      )) +
      ggplot2::geom_tile(color = "black") +
      ggplot2::geom_text(
        ggplot2::aes(color = cum_frac_s1 > 0.5),
        size = 2,
        show.legend = FALSE
      ) +
      ggplot2::theme_bw() +
      ggplot2::labs(
        x = NULL,
        y = NULL,
        fill = "Cumulative fraction\nof shared barcodes\nwithin column sample"
      ) +
      ggplot2::scale_y_discrete(limits = rev) +
      ggplot2::theme(
        text = ggplot2::element_text(size = 7),
        axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
      ) +
      ggplot2::scale_fill_gradient(
        low = "white",
        high = "blue",
        limits = c(0, 1)
      ) +
      ggplot2::scale_color_manual(values = c("black", "white")) +
      ggplot2::coord_fixed(1)

    ggplot2::ggsave(
      outfile_subsetted_col_cumfrac,
      plt_col_met,
      width = width,
      height = height
    )
  }

  ggplot2::ggsave(outfile, plt, width = width, height = height)

  return(list(plt_col_met, plt))
}
```

## Plot barcode overlaps between in-vivo and cell line samples (Figure 7L, Supplementary Figure S9E).

```{r}
plot_bct_barcode_overlaps_cellline_mouse_met_18(
  filter_biorepl_id = c("814_spleen_met", "816_lung_met", "817_spleen", "817_regr_2"),
  filter_timepoint_cellline = "End",
  freq_thresh = 0.01,
  include_palbo_gradual_dmso_only = FALSE,
  width = 6,
  height = 6)
```

## Print session info.

```{r}
sessionInfo()
```
