---
title: "scRNA-seq analyses"
format: html
toc-depth: 4
---

## Preliminaries

Raw and processed scRNAseq files are deposited on GEO under accession number [GSE326012](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE326012). Raw fastqs are used as inputs to nf-core/scrnaseq v2.7.1. We include as processed data R object files (rds) of the following three types:

1. Each individual sample's unfiltered cells. Filename format: scrnaseq-seurat-object-unfiltered-<library name>.rds

2. Each individual sample's QC-metric-filtered cells. Filename format: scrnaseq-seurat-object-qcfiltered-processed-<library name>.rds. Contains predicted cell cycle scores, phase, GLM-PCA embeddings, UMAP-GLM-PCA embeddings, and kmeans clustering assignments in object slots for each sample.

3. All individual sample's QC-metric-filtered cells merged together into a single object. Filename: scrnaseq-seurat-object-qcfiltered-processed-all-samples.rds. Contains predicted cell cycle scores, phase, GLM-PCA embeddings, UMAP-GLM-PCA embeddings, and kmeans clustering on all cells in object slots. The 'orig.ident' column of the 'meta.data' object slot identifies the sample of origin of a cell and matches the "library name" sample-level metadata field. Dimensionality reductions and clustering were done on all merged cells together.

We recommend using pre-computed files from (2) and (3) above for downstream analyses to ensure consistency with paper figures. In cases where preprocessing steps may be computationally intensive or introduce non-deterministic outputs, we load in pre-computed files.

### Load in libraries.

```{r}
library(Seurat)
library(tidyverse)
library(ggrepel)
library(ggridges)
library(clustree)
library(cluster)
library(fishpond)
library(Matrix)
library(SingleCellExperiment)
library(scSHC)
library(org.Hs.eg.db)
library(msigdbr)
library(limma)
library(MatrixGenerics)
library(data.table)
# library(infercnv) # needs JAGS to be loaded to run
```

### Create output data, plot folders.

```{r}
dir.create("data/raw/02-scrnaseq", recursive = TRUE) # input files
dir.create("data/processed/02-scrnaseq", recursive = TRUE) # output data, plots
```

### Load in plot color palettes.

```{r}
pal_orig.ident <- c(
  "Mut_AbemaR" = "orange",
  "WT_AbemaR" = "red",
  "Mut_PalboR" = "lightblue",
  "WT_PalboR" = "royalblue",
  "WT_PalboR_Clone" = "purple",
  "Mut_Vehicle" = "gray85",
  "WT_Vehicle" = "gray30",
  "Parental" = "burlywood4"
)

col_low <- palette.colors(palette = "Okabe-Ito", names = TRUE)[["blue"]]
col_high <- palette.colors(palette = "Okabe-Ito", names = TRUE)[["orange"]]
```

## Sample alignment

### Sample SLURM submission script for nf-core/scrnaseq v2.7.1.

```{bash}
#| eval: false

#!/bin/bash
#SBATCH --job-name=12-nfcore-scrnaseq
#SBATCH --mem=30G
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=1
#SBATCH --ntasks-per-node=1

module load singularity
nextflow run nf-core/scrnaseq \
  -r 2.7.1 \
  -profile singularity \
  -c nextflow.config \
  --input data/processed/12-nfcore-scrnaseq/cdk46i-scrnaseq-metadata.csv \
  --fasta hg38.fa \
  --gtf hg38.gtf.gz \
  --protocol 10XV3 \
  --save_reference \
  --aligner alevin \
  --skip_emptydrops \
  -work-dir /path/to/work/directory \
  --outdir /path/to/output/directory
```

## Create R objects from alignment outputs, QC, cluster

### Load in function to read in nf-core/scrnaseq outputs into Seurat objects without filtering out any cells.

```{r}
#' Collect outputs of nf-core scrnaseq alevin-fry via simpleaf quant (sq), load
#' into a gene x cell matrix, then collapse gene IDs to
#' gene symbols (summing), following Jeremy Simon's tutorial:
#'
#' https://jeremymsimon.github.io/HSPH_scRNA_20241023/HSPH_scRNA_20241023.html#import-alevin-fry-counts-into-r-with-fishpondloadfry
#'
#' Note, this keeps all cells without removing any emptyDroplets cells. This
#' corresponds to individual processed data files per sample of the naming
#' scheme
#'
#' "<GEO-sample-accession-number>_scrnaseq-seurat-object-unfiltered-<sample-name>.rds"
#'
#' deposited in GEO.
#'
#' @param sample_name One of c(Parental, WT_Vehicle, Mut_Vehicle, WT_PalboR,
#' WT_PalboR_Clone, Mut_PalboR, WT_AbemaR, Mut_AbemaR)
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Seurat object of unfiltered counts.
create_count_mtx_from_nfcore_scrnaseq_unfiltered <- function(
  sample_name,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  idir = sprintf(
    "/rafalab/drusso/rjlab/cdk46i/data/processed/12-nfcore-scrnaseq/alevin/%s_alevin_results/af_quant",
    sample_name
  )

  sample_rna <- loadFry(idir, outputFormat = "snRNA")

  tx2gene <- read.table(
    "/rafalab/drusso/rjlab/cdk46i/data/processed/12-nfcore-scrnaseq/alevin/salmon/ref/gene_id_to_name.tsv",
    header = FALSE,
    sep = "\t",
    col.names = c("tx", "gene")
  )

  # Use matrix multiplication to group and sum counts by gene symbol
  exp.txId <- rownames(counts(sample_rna))
  exp.geneId <- as.vector(tx2gene$gene[match(exp.txId, tx2gene$tx)])
  exp.tx.grp <- t(sparse.model.matrix(~ 0 + exp.geneId))

  sample_rna.summarized <- exp.tx.grp %*% counts(sample_rna)
  rownames(sample_rna.summarized) <- rownames(sample_rna.summarized) |>
    str_replace_all(".+.geneId", "")

  out <- CreateSeuratObject(counts = sample_rna.summarized)
  out$orig.ident <- sample_name # for easy merging of all samples later on into a single Seurat object

  saveRDS(
    out,
    file.path(
      ddir,
      sprintf("scrnaseq-seurat-object-unfiltered-%s.rds", sample_name)
    )
  )
  return(out)
}
```

### Read in samples into unfiltered Seurat Objects.

#### Parental 
```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "Parental")
```

#### WT PalboR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "WT_PalboR")
```

#### Mut PalboR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "Mut_PalboR")
```

#### WT PalboR Clone

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(
  sample_name = "WT_PalboR_Clone"
)
```

#### WT AbemaR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "WT_AbemaR")
```

#### Mut AbemaR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "Mut_AbemaR")
```

#### WT Vehicle

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "WT_Vehicle")
```

#### Mut Vehicle

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq_unfiltered(sample_name = "Mut_Vehicle")
```

### Load in function to read in nf-core/scrnaseq outputs into Seurat objects.

```{r}
#' Collect outputs of nf-core scrnaseq alevin-fry via simpleaf quant (sq), load
#' into a gene x cell matrix, remove emptyDroplets, then collapse gene IDs to
#' gene symbols (summing), following Jeremy Simon's tutorial:
#'
#' https://jeremymsimon.github.io/HSPH_scRNA_20241023/HSPH_scRNA_20241023.html#import-alevin-fry-counts-into-r-with-fishpondloadfry
#'
#' Note, paths will need to be changed to match the location of the
#' nf-core/scrnaseq run outputs.
#'
#' @param sample_name One of c(Parental, WT_Vehicle, Mut_Vehicle, WT_PalboR,
#' WT_PalboR_Clone, Mut_PalboR, WT_AbemaR, Mut_AbemaR)
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Seurat object of emptyDrops-filtered counts.
create_count_mtx_from_nfcore_scrnaseq <- function(
  sample_name,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  # Change prefix of path to alevin output directory of nf-core/scrnaseq output
  idir = sprintf(
    "/rafalab/drusso/rjlab/cdk46i/data/processed/12-nfcore-scrnaseq/alevin/%s_alevin_results/af_quant",
    sample_name
  )

  sample_rna <- loadFry(idir, outputFormat = "snRNA")

  e.out <- DropletUtils::emptyDrops(counts(sample_rna))
  is.cell <- e.out$FDR <= 0.01
  is.cell[is.na(is.cell)] = FALSE
  table(is.cell)

  # Remove empty droplets
  gex_q.filt <- sample_rna[, is.cell]

  tx2gene <- read.table(
    "/rafalab/drusso/rjlab/cdk46i/data/processed/12-nfcore-scrnaseq/alevin/salmon/ref/gene_id_to_name.tsv",
    header = FALSE,
    sep = "\t",
    col.names = c("tx", "gene")
  )

  # Use matrix multiplication to group and sum counts by gene symbol
  exp.txId <- rownames(counts(gex_q.filt))
  exp.geneId <- as.vector(tx2gene$gene[match(exp.txId, tx2gene$tx)])
  exp.tx.grp <- t(sparse.model.matrix(~ 0 + exp.geneId))

  gex_q.filt.summarized <- exp.tx.grp %*% counts(gex_q.filt)
  rownames(gex_q.filt.summarized) <- rownames(gex_q.filt.summarized) |>
    str_replace_all(".+.geneId", "")

  out <- CreateSeuratObject(counts = gex_q.filt.summarized)
  out$orig.ident <- sample_name # for easy merging of all samples later on into a single Seurat object

  saveRDS(
    e.out,
    file.path(ddir, sprintf("simpleaf-quant-emptyDrops-%s.rds", sample_name))
  )
  saveRDS(
    out,
    file.path(ddir, sprintf("simpleaf-quant-dds-%s.rds", sample_name))
  )
  return(out)
}
```

#### Parental 

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "Parental")
```

#### WT PalboR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "WT_PalboR")
```

#### Mut PalboR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "Mut_PalboR")
```

#### WT PalboR Clone

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "WT_PalboR_Clone")
```

#### WT AbemaR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "WT_AbemaR")
```

#### Mut AbemaR

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "Mut_AbemaR")
```

#### WT Vehicle

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "WT_Vehicle")
```

#### Mut Vehicle

```{r}
#| eval: false
create_count_mtx_from_nfcore_scrnaseq(sample_name = "Mut_Vehicle")
```

### Load in function to merge filtered Seurat objects per sample into a single object.

```{r}
#' Merge Seurat objects of multiple samples, add in percent mitochondrial
#' genes.
#'
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns SeuratObject of merged samples.
merge_seurat_simpleaf <- function(
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  idir = "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-eda"
  Parental <- readRDS(file.path(idir, "simpleaf-quant-dds-Parental.rds"))
  WT_Vehicle <- readRDS(file.path(idir, "simpleaf-quant-dds-WT_Vehicle.rds"))
  Mut_Vehicle <- readRDS(file.path(idir, "simpleaf-quant-dds-Mut_Vehicle.rds"))
  WT_PalboR <- readRDS(file.path(idir, "simpleaf-quant-dds-WT_PalboR.rds"))
  WT_PalboR_Clone <- readRDS(file.path(
    idir,
    "simpleaf-quant-dds-WT_PalboR_Clone.rds"
  ))
  Mut_PalboR <- readRDS(file.path(idir, "simpleaf-quant-dds-Mut_PalboR.rds"))
  WT_AbemaR <- readRDS(file.path(idir, "simpleaf-quant-dds-WT_AbemaR.rds"))
  Mut_AbemaR <- readRDS(file.path(idir, "simpleaf-quant-dds-Mut_AbemaR.rds"))

  sce <- merge(
    Parental,
    y = c(
      WT_Vehicle,
      Mut_Vehicle,
      WT_PalboR,
      WT_PalboR_Clone,
      Mut_PalboR,
      WT_AbemaR,
      Mut_AbemaR
    ),
    add.cell.ids = c(
      "Parental",
      "WT_Vehicle",
      "Mut_Vehicle",
      "WT_PalboR",
      "WT_PalboR_Clone",
      "Mut_PalboR",
      "WT_AbemaR",
      "Mut_AbemaR"
    ),
    project = "cdk46i"
  )
  sce <- PercentageFeatureSet(
    sce,
    pattern = "^MT-",
    col.name = "percent.mt"
  )
  # Ordering for plot panels, legends, etc.
  sce$orig.ident <- factor(
    sce$orig.ident,
    levels = c(
      "Mut_AbemaR",
      "WT_AbemaR",
      "Mut_PalboR",
      "WT_PalboR",
      "WT_PalboR_Clone",
      "Mut_Vehicle",
      "WT_Vehicle",
      "Parental"
    )
  )

  saveRDS(sce, file.path(ddir, sprintf("simpleaf-quant-dds-allsamples.rds")))
  return(sce)
}
```

### Merge filtered Seurat objects per sample into a single object.

```{r}
#| eval: false
merge_seurat_simpleaf()
```

### Load in function to add QC information to Seurat object of all samples.

```{r}
#' Plot QC metrics of scRNAseq cells, filter dataset according to given fixed
#' thresholds.
#'
#' Uses fixed thresholds as opposed to something like miseq, as the distribution
#' of the QC metrics, makes the mixture model miseq uses a poor fit.
#'
#' @param thresh_max_percent.mt Maximum percent.mt for which cells are kept.
#' @param thresh_min_nFeature_RNA Min nFeature_RNA for which cells are kept.
#' @param thresh_min_nCount_RNA Min nCount_RNA for which cells are kept.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Seurat object of filtered data.
filter_eda_qc <- function(
  thresh_max_percent.mt = 25,
  thresh_min_nFeature_RNA = 100,
  thresh_min_nCount_RNA = 1e4,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds <- readRDS("data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples.rds")

  # Filter the Seurat object, and write to file.
  dds <- dds[, dds$percent.mt <= thresh_max_percent.mt]
  dds <- dds[, dds$nFeature_RNA >= thresh_min_nFeature_RNA]
  dds <- dds[, dds$nCount_RNA >= thresh_min_nCount_RNA]

  dds <- JoinLayers(dds)
  dds <- dds[rowSums(dds@assays$RNA@layers$counts) > 0, ] # filter out genes with zero counts across all cells

  saveRDS(
    dds,
    file.path(ddir, sprintf("simpleaf-quant-dds-allsamples_filterqc.rds"))
  )
  return(dds)
}
```

### Add QC information to Seurat object of all samples.

```{r}
#| eval: false
filter_eda_qc()
```

### Calculate GLM-PCA embeddings.

#### R wrapper script to run GLM-PCA step as SLURM arrayed job: R/07-sc-simpleaf-fastglmpca-kraken.R.

This wrapper R script should be put in a file R/07-sc-simpleaf-fastglmpca-kraken.R.

```{r}
#| eval: false

#' Run scry-glmpca on the merged scrnaseq data separately for each sample.
#'
#' This function is computationally intensive, so we recommend running in an
#' HPC envionment using the sample wrapper R and slurm scripts below.
#'
#' @param infile_seurat_rds Path to rds of Seurat object of all samples.
#' @param sids Vector of sample ids to filter object to. Some of
#' c(Parental, WT_Vehicle, Mut_Vehicle, WT_PalboR, WT_PalboR_Clone,
#' Mut_PalboR, WT_AbemaR, Mut_AbemaR)
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
simpleaf_scry_glmpca <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc.rds",
  sids,
  ddir = "data/processed/07-sc-simpleaf-fastglmpca-kraken",
  odir = "output/07-sc-simpleaf-fastglmpca-kraken"
) {
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  dds <- readRDS(infile_seurat_rds)

  dds_sid <- dds[, dds$orig.ident %in% sids]
  keep <- rowSums(dds_sid@assays$RNA@layers$counts) > 0
  dds_sid <- dds_sid[keep, ]
  cts <- LayerData(dds_sid, assay = "RNA", layer = "counts")

  cat(sids, sep = "\n")
  cat(
    "Calculating top deviance features usng scry::devianceFeatureSelection",
    sep = "\n"
  )
  set.seed(3857)
  sce <- scry::devianceFeatureSelection(cts)
  sce <- sort(sce, decreasing = TRUE)
  saveRDS(
    sce,
    file.path(
      ddir,
      sprintf(
        "scrygenesdeviance_%s_%s.rds",
        prefix,
        paste0(sids, collapse = "-")
      )
    )
  )
  sce_top <- names(sce[1:2500])

  plot(
    log(sce),
    type = "l",
    xlab = "ranked genes",
    ylab = "binomial deviance",
    main = "Feature Selection with Deviance"
  )
  abline(v = 2500, lty = 2, col = "red")

  cat("Running fastglmpca on top deviance features.", sep = "\n")
  fastglmpca::set_fastglmpca_threads(16) # change depending on computing resources
  fit0 <- fastglmpca::init_glmpca_pois(cts[sce_top, ], K = 50)
  fit <- fastglmpca::fit_glmpca_pois(cts[sce_top, ], fit0 = fit0)

  # Shorten output name if all 8 samples are being run.
  if (length(unique(sids)) == 8) {
    saveRDS(
      fit,
      file.path(ddir, sprintf("fastglmpca-K50-scrytop2500bydev_%s.rds", prefix))
    )
    saveRDS(
      cts[sce_top, ],
      file.path(ddir, sprintf("counts-scrytop2500bydev_%s.rds", prefix))
    )
  } else {
    saveRDS(
      fit,
      file.path(
        ddir,
        sprintf(
          "fastglmpca-K50-scrytop2500bydev_%s_%s.rds",
          prefix,
          paste0(sids, collapse = "-")
        )
      )
    )
    saveRDS(
      cts[sce_top, ],
      file.path(
        ddir,
        sprintf(
          "counts-scrytop2500bydev_%s_%s.rds",
          prefix,
          paste0(sids, collapse = "-")
        )
      )
    )
  }
}

args <- commandArgs(trailingOnly = TRUE)
lst = list(
  c(
    "Mut_AbemaR",
    "WT_AbemaR",
    "Mut_PalboR",
    "WT_PalboR",
    "WT_PalboR_Clone",
    "Mut_Vehicle",
    "WT_Vehicle",
    "Parental"
  ), # all samples together
  "Mut_AbemaR",
  "WT_AbemaR",
  "Mut_PalboR",
  "WT_PalboR",
  "WT_PalboR_Clone",
  "Mut_Vehicle",
  "WT_Vehicle",
  "Parental"
)
simpleaf_scry_glmpca(sids = lst[[as.numeric(args[[1]])]]) # uses SLURM array index to choose samples
```

#### Sample SLURM arrayed submission script.

```{bash}
#| eval: false

#!/bin/bash
#SBATCH --job-name=07-sc-simpleaf-fastglmpca-kraken
#SBATCH --time=4:00:00
#SBATCH --mem=72G
#SBATCH --ntasks=1
#SBATCH --array=8
#SBATCH --cpus-per-task=16
#SBATCH --ntasks-per-node=1
#SBATCH --error=logs/07-sc-simpleaf-fastglmpca-kraken_%A-%a.err
#SBATCH --output=logs/07-sc-simpleaf-fastglmpca-kraken_%A-%a.out

module load R/4.3.2

Rscript R/07-sc-simpleaf-fastglmpca-kraken.R $SLURM_ARRAY_TASK_ID
```

### Load in function to add GLM-PCA embeddings to Seurat Objects, calculate UMAP embeddings

We use precomputed outputs of the GLM-PCA computation in this step.

```{r}
#' Merge glmpca embeddings as a slot in a Seurat object of just those cells,
#' calculate cellcycle phases, umap on the glmpca imputs, and kmeans
#' clustering on the data.
#'
#' @param infile_seurat_rds Path to RDS of Seurat object.
#' @param infile_seurat_rds Path to rds of GLM-PCA output.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
merge_seurat_glmpca_umap_kmeans_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc.rds",
  # Paths need to be changed to point to fastglmpca output files.
  infile_fastglmpca = c(
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle.rds",
    "/rafalab/drusso/rjlab/cdk46i/data/processed/07-sc-simpleaf-fastglmpca-kraken/fastglmpca-K50-scrytop2500bydev_simpleaf-quant-dds-allsamples_filterqc_Parental.rds"
  ),
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds_full <- readRDS(infile_seurat_rds)
  dds_full <- NormalizeData(dds_full) # only for calculating Seurat's cell cycle scoring, not for any other analyses.
  s.genes <- base::intersect(cc.genes$s.genes, rownames(dds_full))
  g2m.genes <- base::intersect(cc.genes$g2m.genes, rownames(dds_full))
  dds_full <- CellCycleScoring(
    dds_full,
    s.features = s.genes,
    g2m.features = g2m.genes,
    set.ident = FALSE
  )

  # Loop through GLM-PCA fits, process each separately.
  for (f in infile_fastglmpca) {
    prefix <- sprintf(
      ifelse(
        grepl("excludecc", f),
        "%s_excludecc_glmpca-umap-kmeans",
        "%s_glmpca-umap-kmeans"
      ),
      gsub(
        ".rds",
        "",
        gsub("^fastglmpca-K50.*scrytop2500bydev_", "", basename(f))
      )
    )

    outfile_rds <- file.path(ddir, sprintf("%s.rds", prefix))
    outfile_kmeans_output_rds <- file.path(
      ddir,
      sprintf("%s_kmeans-output.rds", gsub("_glmpca-umap-kmeans", "", prefix))
    )

    outfile_plt_inertia_pdf <- file.path(
      odir,
      sprintf("%s_kmeans-inertia.pdf", prefix)
    )
    outfile_plt_silhouette_pdf <- file.path(
      odir,
      sprintf("%s_kmeans-silhouette.pdf", prefix)
    )
    outfile_plt_inertia_png <- file.path(
      odir,
      sprintf("%s_kmeans-inertia.png", prefix)
    )
    outfile_plt_silhouette_png <- file.path(
      odir,
      sprintf("%s_kmeans-silhouette.png", prefix)
    )

    print(outfile_rds)
    fit <- readRDS(f)
    dds <- dds_full[, rownames(fit$V)] # only take cells in glm-pca fit from entire dataset.

    # Add the glmpca gene embeddings to the Seurat object
    dds[["fastglmpca"]] <- CreateDimReducObject(
      embeddings = fit$V,
      key = "k_",
      assay = DefaultAssay(dds)
    )
    set.seed(101341)

    set.seed(385)
    umap.fit <- uwot::umap(fit$V)
    dds[["umap_fastglmpca"]] <- CreateDimReducObject(
      embeddings = umap.fit,
      key = "",
      assay = DefaultAssay(dds)
    )

    # kmeans clustering
    dist.matrix <- dist(Embeddings(dds, reduction = "fastglmpca"))

    lst = purrr::map(2:10, \(k) {
      print(k)

      # kmeans cluster the glm-pca embeddings
      clst.kmeans <- kmeans(
        Embeddings(dds, reduction = "fastglmpca"),
        k,
        iter.max = 25,
        nstart = 25
      )
      dds@meta.data <<- dds@meta.data |> # global assignment
        mutate(!!sym(sprintf("kmeans%s", k)) := clst.kmeans$cluster)

      centers <- clst.kmeans$centers
      clusters <- clst.kmeans$cluster

      # Calculate inertia
      # sum of square distances of each point to its closest (assigned) cluster centroid
      inertia <- sum(
        (Embeddings(dds, reduction = "fastglmpca") - centers[clusters, ])^2
      )
      clst.kmeans$inertia <- inertia

      # Calculate average silhouette score
      sil <- cluster::silhouette(x = clst.kmeans$cluster, dist = dist.matrix)
      meansil <- mean(sil[, 3]) # average silhouette width over all cells.
      clst.kmeans$silhouette <- meansil

      return(clst.kmeans)
    })

    st = unique(dds$orig.ident)
    # Inertia and Silhouette scores for choosing a suitable k
    kmeans_metrics <- tibble(
      k = 2:10,
      inertia = purrr::map_dbl(lst, \(x) x$inertia),
      silhouette = purrr::map_dbl(lst, \(x) x$silhouette)
    )

    (plt_inertia <- kmeans_metrics |>
      ggplot(aes(x = k, y = inertia)) +
      geom_point() +
      labs(
        title = "Inertia of kmeans clusterings",
        subtitle = paste0(st, collapse = " "),
        x = "Number of kmeans clusters",
        y = "Inertia (sum of squares)"
      ) +
      scale_x_continuous(breaks = 2:10) +
      theme_bw())

    (plt_silhouette <- kmeans_metrics |>
      ggplot(aes(x = k, y = silhouette)) +
      geom_point() +
      labs(
        title = "Average silhouette scores of kmeans clusterings",
        subtitle = paste0(st, collapse = " "),
        x = "Number of kmeans clusters",
        y = "Average silhouette score"
      ) +
      scale_x_continuous(breaks = 2:10) +
      theme_bw())

    # Save plots of inertia and silhouette scores
    ggsave(outfile_plt_inertia_pdf, plt_inertia, width = 6, height = 6)
    ggsave(outfile_plt_silhouette_pdf, plt_silhouette, width = 6, height = 6)
    ggsave(
      outfile_plt_inertia_png,
      plt_inertia,
      width = 6,
      height = 6,
      units = "in"
    )
    ggsave(
      outfile_plt_silhouette_png,
      plt_silhouette,
      width = 6,
      height = 6,
      units = "in"
    )

    # Save seurat objects with kmeans clustering information.
    saveRDS(lst, outfile_kmeans_output_rds)
    saveRDS(dds, outfile_rds)
  }
}
```

### Add GLM-PCA embeddings to Seurat Objects, calculate UMAP embeddings.

```{r}
#| eval: false
merge_seurat_glmpca_umap_kmeans_simpleaf()
```

## Plotting downstream analyses

### Load in function to plot GLM-PCA and UMAP-GLM-PCA embeddings for calculated clusters.

```{r}
#' Plot GLM-PCA and UMAPs of data, and kmeans clusterings.
#'
#' @param infile_seurat_rds Path to rds of Seurat object with embeddings, kmeans clusterings.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_seurat_glmpca_umap_cluster_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds <- readRDS(infile_seurat_rds)
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))

  outfile_pdf <- file.path(
    ddir,
    sprintf(
      "%s_glmpca-umap-kmeans.rds",
      gsub(".rds", "", basename(infile_seurat_rds))
    )
  )
  st = unique(dds$orig.ident)

  meta_glmpca <- Embeddings(dds, reduction = "fastglmpca") |>
    as.data.frame() |>
    tibble::rownames_to_column("sid") |>
    inner_join(
      dds@meta.data |> as.data.frame() |> tibble::rownames_to_column("sid"),
      by = "sid"
    )

  meta_umapglmpca <- Embeddings(dds, reduction = "umap_fastglmpca") |>
    as.data.frame() |>
    tibble::rownames_to_column("sid") |>
    inner_join(
      dds@meta.data |> as.data.frame() |> tibble::rownames_to_column("sid"),
      by = "sid"
    )

  meta_glmpca$zero_fraction = colSums(
    GetAssayData(dds, layer = "counts") == 0
  ) /
    nrow(dds)
  meta_umapglmpca$zero_fraction = colSums(
    GetAssayData(dds, layer = "counts") == 0
  ) /
    nrow(dds)

  plt_glmpca <- meta_glmpca |>
    ggplot(aes(x = k_1, y = k_2, col = orig.ident)) +
    geom_point(alpha = .5, size = .25) +
    theme_bw() +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + # so legend is readable
    labs(
      title = "GLM-PCA cell embeddings",
      subtitle = paste0(st, collapse = " "),
      x = "GLM-PCA Embedding 1",
      y = "GLM-PCA Embedding 2"
    ) +
    scale_color_manual(values = pal_orig.ident, name = "Condition")
  ggsave(
    file.path(odir, sprintf("%s_glmpca-condition.pdf", prefix)),
    plt_glmpca,
    width = 7,
    height = 6
  )
  ggsave(
    file.path(odir, sprintf("%s_glmpca-condition.png", prefix)),
    plt_glmpca,
    width = 7,
    height = 6,
    units = "in"
  )

  plt_glmpca_phase <- meta_glmpca |>
    ggplot(aes(x = k_1, y = k_2, col = Phase)) +
    geom_point(alpha = .5, size = .25) +
    theme_bw() +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + # so legend is readable
    labs(
      title = "GLM-PCA cell embeddings",
      subtitle = paste0(st, collapse = " "),
      x = "GLM-PCA Embedding 1",
      y = "GLM-PCA Embedding 2"
    )
  ggsave(
    file.path(odir, sprintf("%s_glmpca-phase.pdf", prefix)),
    plt_glmpca_phase,
    width = 7,
    height = 6
  )
  ggsave(
    file.path(odir, sprintf("%s_glmpca-phase.png", prefix)),
    plt_glmpca_phase,
    width = 7,
    height = 6,
    units = "in"
  )

  (plt_umapglmpca <- meta_umapglmpca |>
    ggplot(aes(
      x = !!sym(colnames(Embeddings(dds, reduction = "umap_fastglmpca"))[[1]]),
      y = !!sym(colnames(Embeddings(dds, reduction = "umap_fastglmpca"))[[2]]),
      col = orig.ident
    )) +
    geom_point(alpha = .5, size = .25) +
    theme_bw() +
    guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + # so legend is readable
    labs(
      title = "UMAP on GLM-PCA cell embeddings",
      subtitle = paste0(st, collapse = " "),
      x = "UMAP-GLM-PCA Embedding 1",
      y = "UMAP-GLM-PCA Embedding 2"
    ) +
    scale_color_manual(values = pal_orig.ident, name = "Condition"))
  print(plt_umapglmpca)
  ggsave(
    file.path(odir, sprintf("%s_umap-condition.pdf", prefix)),
    plt_umapglmpca,
    width = 7,
    height = 6
  )
  ggsave(
    file.path(odir, sprintf("%s_umap-condition.png", prefix)),
    plt_umapglmpca,
    width = 7,
    height = 6,
    units = "in"
  )

  print(
    plt_umapglmpca_phase <- meta_umapglmpca |>
      ggplot(aes(
        x = !!sym(colnames(Embeddings(dds, reduction = "umap_fastglmpca"))[[
          1
        ]]),
        y = !!sym(colnames(Embeddings(dds, reduction = "umap_fastglmpca"))[[
          2
        ]]),
        col = Phase
      )) +
      geom_point(alpha = .5, size = .25) +
      theme_bw() +
      guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) + # so legend is readable
      labs(
        title = "UMAP on GLM-PCA cell embeddings",
        subtitle = paste0(st, collapse = " "),
        x = "UMAP-GLM-PCA Embedding 1",
        y = "UMAP-GLM-PCA Embedding 2"
      )
  )
  ggsave(
    file.path(odir, sprintf("%s_umap-phase.pdf", prefix)),
    plt_umapglmpca_phase,
    width = 7,
    height = 6
  )
  ggsave(
    file.path(odir, sprintf("%s_umap-phase.png", prefix)),
    plt_umapglmpca_phase,
    width = 7,
    height = 6,
    units = "in"
  )

  # Clustree diagram of kmeans clustering
  plt_clustree_kmeans <- clustree::clustree(dds, prefix = "kmeans") +
    labs(
      title = sprintf(
        "kmeans clustering Resolutions %s",
        paste0(gsub(" ", "-", st), collapse = "-")
      )
    )
  print(plt_clustree_kmeans)
  ggsave(
    file.path(odir, sprintf("%s_kmeans-clustree.pdf", prefix)),
    plt_clustree_kmeans,
    width = 8,
    height = 12
  )
  ggsave(
    file.path(odir, sprintf("%s_kmeans-clustree.png", prefix)),
    plt_clustree_kmeans,
    width = 8,
    height = 12,
    units = "in"
  )

  # Make plots for all kmeans, adding in cluster size information.
  for (i in stringr::str_c("kmeans", 2:10)) {
    plt_pca_kmeans_cluster <- # PC1 against PC2
      meta_glmpca |>
      group_by(!!sym(i)) |>
      add_count() |>
      mutate(!!sym(i) := str_c(!!sym(i), " (N = ", n, ")")) |>
      ggplot(aes(x = k_1, y = k_2, col = factor(!!sym(i)))) +
      geom_point(alpha = .5, size = .5) +
      theme_bw() +
      guides(color = guide_legend(override.aes = list(size = 2))) +
      labs(
        title = "GLM-PCA cell embeddings",
        subtitle = sprintf(
          "%s; kmeans clustering %s",
          paste0(st, collapse = " "),
          i
        ),
        x = "GLM-PCA Embedding 1",
        y = "GLM-PCA Embedding 2"
      )

    plt_umap_kmeans_clusters <- meta_umapglmpca |>
      group_by(!!sym(i)) |>
      add_count() |>
      mutate(!!sym(i) := str_c(!!sym(i), " (N = ", n, ")")) |>
      ggplot(aes(
        x = !!sym(colnames(Embeddings(dds, reduction = "umap_fastglmpca"))[[
          1
        ]]),
        y = !!sym(colnames(Embeddings(dds, reduction = "umap_fastglmpca"))[[
          2
        ]]),
        col = factor(!!sym(i))
      )) +
      geom_point(alpha = .5, size = .5) +
      theme_bw() +
      guides(color = guide_legend(override.aes = list(size = 2))) +
      labs(
        title = "UMAP on GLM-PCA cell embeddings",
        subtitle = sprintf(
          "%s; kmeans clustering %s",
          paste0(st, collapse = " "),
          i
        ),
        x = "UMAP-GLM-PCA Embedding 1",
        y = "UMAP-GLM-PCA Embedding 2"
      )
    print(plt_umap_kmeans_clusters)

    ggsave(
      file.path(odir, sprintf("%s_glmpca_%s.pdf", prefix, i)),
      plt_pca_kmeans_cluster,
      width = 7,
      height = 6
    )
    ggsave(
      file.path(odir, sprintf("%s_umap_%s.pdf", prefix, i)),
      plt_umap_kmeans_clusters,
      width = 7,
      height = 6
    )

    ggsave(
      file.path(odir, sprintf("%s_glmpca_%s.png", prefix, i)),
      plt_pca_kmeans_cluster,
      width = 7,
      height = 6,
      units = "in"
    )
    ggsave(
      file.path(odir, sprintf("%s_umap_%s.png", prefix, i)),
      plt_umap_kmeans_clusters,
      width = 7,
      height = 6,
      units = "in"
    )
  }
}
```

### Plot GLM-PCA and UMAP-GLM-PCA embeddings for calculated clusters, coloring clusters by sample name or kmeans cluster. 

These function calls plot GLM-PCA, UMAP-GLM-PCA embeddings of cells for specified samples, as well as the clustree diagram of kmeans clustering values, silhouette score plots.

#### All samples (Figures 6A, 6B).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds"
)
```

#### Parental (Figure 6E).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds"
)
```

#### WT PalboR (Figure 6F).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds"
)
```

#### Mut PalboR (Figure 6G).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds"
)
```

#### WT PalboR Clone (Figure 6H).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds"
)
```

#### WT AbemaR (Figure 6I).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds"
)
```

#### Mut AbemaR (Figure 6J).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds"
)
```

#### WT Vehicle (Supplementary Figure S7D).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds"
)
```

#### Mut Vehicle (Supplementary Figure S7E).

```{r}
plot_seurat_glmpca_umap_cluster_simpleaf(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds"
)
```

### Load in function to plot barplot of proportions of cell cycle phases per sample.

```{r}
#' Plot a barplot of the proportion of cells in each predicted Seurat cell
#' cycle phase per sample.
#'
#' @param infile_seurat_rds Path to rds of Seurat object of all samples.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns ggplot2 object of the cell cycle phase barplot.
plot_cell_cycle_barplot_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  dds <- readRDS(infile_seurat_rds)

  (plt <- dds@meta.data |>
    ggplot(aes(x = orig.ident, fill = Phase)) +
    geom_bar(position = "fill") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(
      x = NULL,
      y = "Proportion of cells",
      title = "Seurat cell cycle scoring predicted class counts"
    ))

  ggsave(
    file.path(
      odir,
      sprintf("%s_seurat-cell-cycle-scoring_barplot.pdf", prefix)
    ),
    plt,
    height = 5,
    width = 5
  )
  return(plt)
}
```

### Plot cell cycle phase proportions per sample (Figure 6C).

```{r}
plot_cell_cycle_barplot_simpleaf()
```

### Load in function to plot cell-cell distances within each sample.

```{r}
#' Per sample, calculate matrices of cell-cell Euclidean distances on GLM-PCA
#' cell embeddings for 1000 sampled cells.
#'
#' @param infile_seurat_rds Path to rds of Seurat object of all samples.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
get_cell_cell_dists_within_sample_fastglmpca_embeddings_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  dds <- readRDS(infile_seurat_rds)
  Idents(dds) <- "orig.ident"

  gene.embedding.tbl <- Embeddings(dds, "fastglmpca") |>
    as.data.frame() |>
    tibble::rownames_to_column("barcode_id") |>
    tibble::as_tibble()

  gene.embedding.tbl.meta <- dds@meta.data |>
    tibble::rownames_to_column("barcode_id") |>
    inner_join(gene.embedding.tbl, by = "barcode_id") |>
    group_by(orig.ident)

  set.seed(214)
  dists <- group_split(gene.embedding.tbl.meta) |>
    purrr::map(\(df) {
      df |>
        tibble::column_to_rownames("barcode_id") |>
        dplyr::select(k_1:k_50) |>
        dplyr::slice_sample(n = 1000) |>
        dist(method = "euclidean")
    })

  dists.tbl <- dists |>
    purrr::map(\(df) {
      ds <- df |> as.matrix()
      ds[upper.tri(ds, diag = TRUE)] <- NA
      ds |>
        as.data.frame() |>
        tibble::rownames_to_column("barcode_id_1") |>
        pivot_longer(
          !barcode_id_1,
          names_to = "barcode_id_2",
          values_to = "dist"
        ) |>
        dplyr::filter(!is.na(dist))
    })

  dist.tbl.meta <- dists.tbl |>
    purrr::list_rbind() |>
    left_join(
      dds@meta.data |>
        tibble::rownames_to_column("barcode_id") |>
        tibble::as_tibble(),
      by = c("barcode_id_1" = "barcode_id")
    )

  boxplt <- dist.tbl.meta |>
    ggplot(aes(x = orig.ident, y = dist, fill = orig.ident)) +
    geom_boxplot() +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      legend.position = 'none'
    ) +
    labs(
      title = "Cell-cell Euclidean distances on GLM-PCA embeddings",
      y = "Euclidean distance",
      subtitle = "1000 random cells per sample",
      x = NULL
    ) +
    scale_fill_manual(values = pal_orig.ident)

  ggsave(
    file.path(
      odir,
      sprintf("%s_cell-cell-dists-boxplot-1000sampled.pdf", prefix)
    ),
    boxplt,
    height = 6,
    width = 6
  )
  ggsave(
    file.path(
      odir,
      sprintf("%s_cell-cell-dists-boxplot-1000sampled.png", prefix)
    ),
    boxplt,
    height = 6,
    width = 6
  )

  print(boxplt)
}
```

### Plot cell-cell distances within each sample (Supplementary Figure S7A).

```{r}
get_cell_cell_dists_within_sample_fastglmpca_embeddings_simpleaf()
```

### Load in function to plot dotplot of expression of specific genes per sample.

```{r}
#' Plot dotplot of specific CNV related genes for all samples.
#'
#' @param infile_seurat_rds Path to rds of Seurat object of all samples.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_markers_cnv_genes_dotplots_per_sample_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  # Contains genes (gene_id), annotations (Function) listed in Supplementary Figure S7B
  cnvgenes <- readxl::read_xlsx(
    "/rafalab/drusso/rjlab/cdk46i/data/raw/CNV_ordered_gene_list_19-11-24.xlsx",
    col_names = FALSE
  )
  names(cnvgenes) <- c("Function", "gene_id")
  cnvgenes <- cnvgenes |>
    fill(Function, .direction = "down")

  cnvgenes |> print(n = 100)

  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  dds <- readRDS(infile_seurat_rds)

  Idents(dds) <- "orig.ident"
  (plt <- DotPlot(dds, cnvgenes$gene_id)$data |>
    inner_join(cnvgenes, by = c("features.plot" = "gene_id")) |>
    tibble::rownames_to_column("gid") |>
    dplyr::mutate(
      orig.ident = forcats::fct_relevel(
        id,
        "Mut_AbemaR",
        "WT_AbemaR",
        "Mut_PalboR",
        "WT_PalboR",
        "WT_PalboR_Clone",
        "Mut_Vehicle",
        "WT_Vehicle",
        "Parental"
      )
    ) |>
    ggplot(aes(
      x = orig.ident,
      y = features.plot,
      col = avg.exp.scaled,
      size = pct.exp
    )) +
    geom_point() +
    theme_bw() +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      strip.text.y = element_text(angle = 0),
      axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
    ) +
    scale_color_gradient(
      low = "lightgray",
      high = "blue",
      name = "Average Expression"
    ) +
    scale_radius(
      range = c(0, 3),
      limits = c(NA, NA),
      name = "Percent Expressed"
    ) +
    labs(
      title = "Marker expressions by sample",
      x = NULL,
      y = NULL
    ) +
    facet_grid(rows = vars(Function), scales = "free", space = "free_y"))

  ggsave(
    file.path(odir, sprintf("%s_dotplot_cnvgenes.pdf", prefix)),
    plt,
    width = 7,
    height = 12
  )

  return(plt)
}
```

### Plot dotplot of expression of specific genes per sample (Supplementary Figure S7B).

```{r}
#| fig-width: 8
#| fig-height: 11
plot_markers_cnv_genes_dotplots_per_sample_simpleaf()
```

## Gene set testing 

### Estimate genewise dispersions and expression rates per sample.

```{r}
#' Calculate genewise dispersions (phi) per each of the 8 samples.
#'
#' These dispersions get used in calculating the Wald statistic of the difference
#' in average expression between groups when doing differential expression and
#' gene set testing.
#'
#' We allow each sample to have its own genewise dispersions, as these may
#' differ among the 8 samples.
#'
#' @param infile_seurat_rds Path to Seurat object rds.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of overdispersion and rate (lambda) terms per gene, per sample.
get_genewise_rates_and_dispersions_per_sample <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds <- readRDS(infile_seurat_rds)
  cts <- SeuratObject::GetAssayData(dds, layer = "counts")
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))

  phi_df <- c(
    "Mut_AbemaR",
    "WT_AbemaR",
    "Mut_PalboR",
    "WT_PalboR",
    "WT_PalboR_Clone",
    "Mut_Vehicle",
    "WT_Vehicle",
    "Parental"
  ) |>
    purrr::map_dfr(\(b) {
      cts <- SeuratObject::GetAssayData(
        dds[, dds$orig.ident == b],
        layer = "counts"
      ) # gene x cells orientation
      n <- Matrix::colSums(cts)
      N <- sum(n)
      lam <- Matrix::rowSums(cts) / N # gene-specific rates
      rates <- outer(lam, n)
      phi <- Matrix::rowSums((cts - rates)^2 / rates) / (ncol(cts) - 1) # estimated dispersion
      stat <- lam / sqrt(pmin(phi, 1) * lam)

      phi |>
        as.data.frame() |>
        cbind(lam) |>
        cbind(stat) |>
        tibble::rownames_to_column("gene") |>
        dplyr::mutate(N_reads = N) |>
        dplyr::mutate(N_cells = ncol(cts)) |>
        dplyr::mutate(orig.ident = b)
    })

  readr::write_csv(
    phi_df,
    file.path(ddir, sprintf("genewise-overdispersions_%s.csv", prefix))
  )
  saveRDS(
    phi_df,
    file.path(ddir, sprintf("genewise-overdispersions_%s.rds", prefix))
  )

  return(phi_df)
}
```

```{r}
#| output: false
get_genewise_rates_and_dispersions_per_sample()
```

### Estimate genewise expression rates within cell cycle phases per sample.

```{r}
#' Calculate gene rates within each cell cycle phase for each sample.
#'
#' These rates will be used for ranking genes and gene set testing across samples,
#' but subsetted to just within-the same cell cycle phases.
#'
#' @param infile_seurat_rds Path to Seurat object rds.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of rate (lambda) terms per gene, per sample, per cell cycle phase.
get_genewise_rates_per_phase_within_sample <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds <- readRDS(infile_seurat_rds)
  cts <- SeuratObject::GetAssayData(dds, layer = "counts")
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))

  lam_df <- c(
    "Mut_AbemaR",
    "WT_AbemaR",
    "Mut_PalboR",
    "WT_PalboR",
    "WT_PalboR_Clone",
    "Mut_Vehicle",
    "WT_Vehicle",
    "Parental"
  ) |>
    purrr::map_dfr(\(b) {
      c("S", "G1", "G2M") |>
        purrr::map_dfr(\(ph) {
          cts <- SeuratObject::GetAssayData(
            dds[, (dds$orig.ident == b & dds$Phase == ph)],
            layer = "counts"
          ) # gene x cells orientation
          n <- Matrix::colSums(cts)
          N <- sum(n)
          lam <- Matrix::rowSums(cts) / N # gene-specific rates

          lam |>
            as.data.frame() |>
            tibble::rownames_to_column("gene") |>
            dplyr::mutate(N_reads = N) |>
            dplyr::mutate(N_cells = ncol(cts)) |>
            dplyr::mutate(orig.ident = b) |>
            dplyr::mutate(Phase = ph)
        })
    })

  readr::write_csv(
    lam_df,
    file.path(
      ddir,
      sprintf("genewise-rates-within-phase-per-sample_%s.csv", prefix)
    )
  )
  saveRDS(
    lam_df,
    file.path(
      ddir,
      sprintf("genewise-rates-within-phase-per-sample_%s.rds", prefix)
    )
  )
  return(lam_df)
}
```

```{r}
#| output: false
get_genewise_rates_per_phase_within_sample()
```

### Estimate genewise expression rates within clusters per sample.

```{r}
#' Calculate genewise expression rates per each kmeans cluster for the given
#' sample.
#'
#' These rates get used in calculating the Wald statistic of the difference
#' in average expression between groups when doing differential expression and
#' gene set testing.
#'
#' @param infile_seurat_rds Path to Seurat object rds of a single sample.
#' @param cluster_var Integer of which kmeans clustering (2 to 10) to use.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of gene expression rates per gene, per sample.
get_genewise_rates_per_cluster_within_sample <- function(
  infile_seurat_rds,
  cluster_var,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds <- readRDS(infile_seurat_rds)
  cts <- SeuratObject::GetAssayData(dds, layer = "counts")
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))

  orig.ident <- unique(dds$orig.ident)

  lam_df <- 1:cluster_var |>
    purrr::map_dfr(\(x) {
      cts <- SeuratObject::GetAssayData(
        dds[,
          dds@meta.data |> pull(!!sym(sprintf("kmeans%i", cluster_var))) == x
        ],
        layer = "counts"
      ) # gene x cells orientation
      n <- Matrix::colSums(cts)
      N <- sum(n)
      lam <- Matrix::rowSums(cts) / N # gene-specific rates

      lam |>
        as.data.frame() |>
        tibble::rownames_to_column("gene") |>
        dplyr::mutate(N_reads = N) |>
        dplyr::mutate(N_cells = ncol(cts)) |>
        dplyr::mutate(orig.ident = orig.ident) |>
        dplyr::mutate(kmclust = sprintf("kmeans%i", cluster_var)) |>
        dplyr::mutate(cluster = x)
    })

  readr::write_csv(
    lam_df,
    file.path(
      ddir,
      sprintf("genewise-rates-within-kmeans%i_%s.csv", cluster_var, prefix)
    )
  )
  saveRDS(
    lam_df,
    file.path(
      ddir,
      sprintf("genewise-rates-within_kmeans%i_%s.rds", cluster_var, prefix)
    )
  )
  return(lam_df)
}
```

#### Parental

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### WT PalboR

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### Mut PalboR

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### WT PalboR Clone

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### WT AbemaR

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### Mut AbemaR

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### WT Vehicle

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

#### Mut Vehicle

```{r}
#| output: false
purrr::map(c(2,3), \(x) {
  get_genewise_rates_per_cluster_within_sample(
    infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds",
    cluster_var = x)
})
```

### Estimate genewise expression rates within cell cycle phases within clusters per sample.

```{r}
#' Calculate genewise rates per each cell cycle phase within each kmeans 
#' cluster for the given sample.
#'
#' These rates get used in calculating the Wald statistic of the difference
#' in average expression between groups when doing differential expression and
#' gene set testing.
#'
#' @param infile_seurat_rds Path to Seurat object rds of a single sample.
#' @param cluster_var 
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of gene expression rates per gene, 
#' within cluster within cell cycle phase for a single sample.
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample <- function(
  infile_seurat_rds,
  cluster_var,
  min_cells = 50,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dds <- readRDS(infile_seurat_rds)
  cts <- SeuratObject::GetAssayData(dds, layer = "counts")
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))

  lam_df <- 1:cluster_var |>
    purrr::map(\(x) {
      purrr::map(c("G1", "G2M", "S"), \(phase) { # cell cycle phase within cluster
        print(sprintf(
          "%s:%s %i",
          x,
          phase,
          sum(
            ((dds@meta.data |> pull(!!sym(sprintf("kmeans%i", cluster_var))) ==
              x) &
              (dds$Phase == phase))
          )
        ))

        if (
          sum(
            (dds@meta.data |> pull(!!sym(sprintf("kmeans%i", cluster_var))) ==
              x) &
              (dds$Phase == phase)
          ) <
            min_cells
        ) {
          return(NULL) # too few cells to calculate rate confidently
        }

        cts <- SeuratObject::GetAssayData(
          dds[,
            ((dds@meta.data |> pull(!!sym(sprintf("kmeans%i", cluster_var))) ==
              x) &
              (dds$Phase == phase))
          ],
          layer = "counts"
        ) # gene x cells orientation
        n <- Matrix::colSums(cts)
        N <- sum(n)
        lam <- Matrix::rowSums(cts) / N # gene-specific rates

        lam |>
          as.data.frame() |>
          tibble::rownames_to_column("gene") |>
          dplyr::mutate(N_reads = N) |>
          dplyr::mutate(N_cells = ncol(cts)) |>
          dplyr::mutate(orig.ident = unique(dds$orig.ident)) |>
          dplyr::mutate(kmclust = sprintf("kmeans%i", cluster_var)) |>
          dplyr::mutate(cluster = x) |>
          dplyr::mutate(phase = phase) |>
          dplyr::mutate(cluster_phase = sprintf("%s:%s", cluster, phase))
      }) |>
        purrr::list_rbind()
    }) |>
    purrr::list_rbind() |>
    dplyr::filter(N_cells >= min_cells) # require at least min_cells cells to confidently call rate within a cluster

  readr::write_csv(
    lam_df,
    file.path(
      ddir,
      sprintf(
        "genewise-rates-within-kmeans%i-within-phase_geq%icells_%s.csv",
        cluster_var,
        min_cells,
        prefix
      )
    )
  )
  saveRDS(
    lam_df,
    file.path(
      ddir,
      sprintf(
        "genewise-rates-within_kmeans%i-within-phase_geq%icells_%s.rds",
        cluster_var,
        min_cells,
        prefix
      )
    )
  )
  return(lam_df)
}
```

#### Parental

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds",
  cluster_var = 3)
```

#### WT PalboR

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds",
  cluster_var = 3)
```

#### Mut PalboR

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds",
  cluster_var = 3)
```

#### WT PalboR Clone

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds",
  cluster_var = 3)
```

#### WT AbemaR

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds",
  cluster_var = 2)
```

#### Mut AbemaR

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds",
  cluster_var = 2)
```

#### WT Vehicle

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds",
  cluster_var = 2)
```

#### Mut Vehicle

```{r}
#| output: false
get_genewise_rates_per_cluster_per_cell_cycle_phase_within_sample(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds",
  cluster_var = 2)
```

### Calculate quasi-Poisson Wald statistics comparing two samples at a time.

```{r}
#' Calculate the Wald statistic in a quasiPoisson glm that compares two 
#' samples at a time per gene.
#'
#' Uses the sample-specific, per-gene estimated dispersions calculated above
#' as part of the standard error of the tested coefficient in the QP glm.
#'
#' @param infile_seurat_rds Path to rds of Seurat object.
#' @param infile_dispersions Path to rds of estimated genewise dispersion per 
#' sample.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of Wald statistics comparing samples and sample 
#' specific expression rates and dispersion estimates.
quasipoisson_glm_de_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  infile_dispersions = "data/processed/02-scrnaseq/genewise-overdispersions_simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  disps <- readRDS(infile_dispersions)

  disps_A = disps |> filter(orig.ident == "Mut_AbemaR")
  disps_B = disps |> filter(orig.ident == "WT_AbemaR")
  stopifnot(all(disps_A$gene == disps_B$gene))

  disps_wide <- disps |>
    pivot_wider(
      id_cols = gene,
      names_from = orig.ident,
      values_from = c("phi", "lam", "stat", "N_cells", "N_reads")
    )

  nms <- c(
    "Mut_AbemaR",
    "WT_AbemaR",
    "Mut_PalboR",
    "WT_PalboR",
    "WT_PalboR_Clone",
    "Mut_Vehicle",
    "WT_Vehicle",
    "Parental"
  )
  tidyr::expand_grid(x = nms, y = nms) |>
    dplyr::filter(x != y) |>
    purrr::pwalk(\(x, y) {
      disps_wide <<- disps_wide |> # global assignment
        dplyr::mutate(
          !!sym(sprintf("waldstat_%s_minus_%s", x, y)) := ((log(
            !!sym(sprintf("lam_%s", x))
          ) -
            log(!!sym(sprintf("lam_%s", y)))) /
            sqrt(
              (pmax(1, !!sym(sprintf("phi_%s", x))) /
                (!!sym(sprintf("lam_%s", x)) *
                  !!sym(sprintf("N_reads_%s", x)))) +
                (pmax(1, !!sym(sprintf("phi_%s", y))) /
                  (!!sym(sprintf("lam_%s", y)) *
                    !!sym(sprintf("N_reads_%s", y))))
            ))
        )
    })

  saveRDS(
    disps_wide,
    file.path(
      ddir,
      sprintf("waldstat-ranked-genes-across-samples_%s.rds", prefix)
    )
  )
  return(disps_wide)
}
```

```{r}
quasipoisson_glm_de_simpleaf()
```

### Calculate quasi-Poisson Wald statistics comparing two samples within a cell cycle phase at a time.

```{r}
#' Calculate the Wald statistic in a quasiPoisson glm that compares two 
#' samples at a time within a cell cycle phase per gene.
#'
#' Uses the sample-specific, per-gene estimated dispersions calculated above
#' as part of the standard error of the tested coefficient in the QP glm.
#'
#' @param infile_seurat_rds Path to rds of genewise expression rates within a 
#' cell cycle phase per all samples.
#' @param infile_dispersions Path to rds of estimated genewise dispersion per 
#' sample.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of Wald statistics comparing samples within a cell 
#' cycle phase and (sample and cell cycle phase) specific expression rates and 
#' sample specific dispersion estimates.
quasipoisson_glm_de_across_samples_within_phase_simpleaf <- function(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within-phase-per-sample_simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  infile_dispersions = "data/processed/02-scrnaseq/genewise-overdispersions_simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_rates))
  rates <- readRDS(infile_rates)
  disps <- readRDS(infile_dispersions)

  rates_wide <- rates |>
    dplyr::left_join(
      disps |> dplyr::select(gene, phi, orig.ident),
      by = c("gene", "orig.ident")
    ) |>
    tidyr::pivot_wider(
      id_cols = gene,
      names_from = c("orig.ident", "Phase"),
      values_from = c("lam", "N_reads", "N_cells", "phi")
    )

  nms_S <- c(
    "Mut_AbemaR_S",
    "WT_AbemaR_S",
    "Mut_PalboR_S",
    "WT_PalboR_S",
    "WT_PalboR_Clone_S",
    "Mut_Vehicle_S",
    "WT_Vehicle_S",
    "Parental_S"
  )
  nms_G1 <- c(
    "Mut_AbemaR_G1",
    "WT_AbemaR_G1",
    "Mut_PalboR_G1",
    "WT_PalboR_G1",
    "WT_PalboR_Clone_G1",
    "Mut_Vehicle_G1",
    "WT_Vehicle_G1",
    "Parental_G1"
  )
  nms_G2M <- c(
    "Mut_AbemaR_G2M",
    "WT_AbemaR_G2M",
    "Mut_PalboR_G2M",
    "WT_PalboR_G2M",
    "WT_PalboR_Clone_G2M",
    "Mut_Vehicle_G2M",
    "WT_Vehicle_G2M",
    "Parental_G2M"
  )
  rbind(
    tidyr::expand_grid(x = nms_S, y = nms_S) |> dplyr::filter(x != y),
    tidyr::expand_grid(x = nms_G1, y = nms_G1) |> dplyr::filter(x != y),
    tidyr::expand_grid(x = nms_G2M, y = nms_G2M) |> dplyr::filter(x != y)
  ) |>
    purrr::pwalk(\(x, y) {
      rates_wide <<- rates_wide |> # global assignment
        dplyr::mutate(
          !!sym(sprintf("waldstat_%s_minus_%s", x, y)) := ((log(
            !!sym(sprintf("lam_%s", x))
          ) -
            log(!!sym(sprintf("lam_%s", y)))) /
            sqrt(
              (pmax(1, !!sym(sprintf("phi_%s", x))) /
                (!!sym(sprintf("lam_%s", x)) *
                  !!sym(sprintf("N_reads_%s", x)))) +
                (pmax(1, !!sym(sprintf("phi_%s", y))) /
                  (!!sym(sprintf("lam_%s", y)) *
                    !!sym(sprintf("N_reads_%s", y))))
            ))
        )
    })

  saveRDS(
    rates_wide,
    file.path(
      ddir,
      sprintf("waldstat-ranked-genes-across-sample_within-phase_%s.rds", prefix)
    )
  )
  return(rates_wide)
}
```

```{r}
quasipoisson_glm_de_across_samples_within_phase_simpleaf()
```

### Calculate quasi-Poisson Wald statistics comparing clusters within a single sample.

```{r}
#' Calculate the Wald statistic in a quasiPoisson glm that compares two 
#' clusters within a single sample at a time.
#'
#' Uses the sample-specific, per-gene estimated dispersions calculated above
#' as part of the standard error of the tested coefficient in the QP glm.
#'
#' @param infile_rates Path to rds of estimated genewise rates per cluster 
#' for a single sample.
#' @param infile_dispersions Path to rds of estimated genewise dispersion per 
#' sample.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of Wald statistics comparing clusters, cluster 
#' specific expression rates for a single sample, and sample-specific 
#' dispersion estimates.
quasipoisson_glm_de_across_clusters_within_sample_simpleaf <- function(
  infile_rates,
  infile_dispersions = "data/processed/02-scrnaseq/genewise-overdispersions_simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_rates))
  rates <- readRDS(infile_rates)
  k = as.numeric(gsub(
    "^genewise-rates-within_kmeans([0-9]+)_.*",
    "\\1",
    prefix
  ))

  disps <- readRDS(infile_dispersions) |>
    dplyr::filter(orig.ident == unique(rates$orig.ident))

  rates_wide <- rates |>
    tidyr::pivot_wider(
      id_cols = gene,
      names_from = cluster,
      values_from = c("lam", "N_reads", "N_cells")
    ) |>
    dplyr::left_join(disps, by = "gene")

  tidyr::expand_grid(x = 1:k, y = 1:k) |>
    dplyr::filter(x != y) |>
    purrr::pwalk(\(x, y) {
      rates_wide <<- rates_wide |> # global assignment
        dplyr::mutate(
          !!sym(sprintf("waldstat_cluster%s_minus_cluster%s", x, y)) := ((log(
            !!sym(sprintf("lam_%s", x))
          ) -
            log(!!sym(sprintf("lam_%s", y)))) /
            sqrt(
              (pmax(1, phi) /
                (!!sym(sprintf("lam_%s", x)) *
                  !!sym(sprintf("N_reads_%s", x)))) +
                (pmax(1, phi) /
                  (!!sym(sprintf("lam_%s", y)) *
                    !!sym(sprintf("N_reads_%s", y))))
            ))
        )
    })

  saveRDS(
    rates_wide,
    file.path(
      ddir,
      sprintf(
        "waldstat-ranked-genes-across-cluster_within-sample_%s.rds",
        prefix
      )
    )
  )
  return(rates_wide)
}
```

#### Parental

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds"
)
```

#### WT PalboR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds"
)
```

#### Mut PalboR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3_simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds"
)
```

#### WT PalboR Clone

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds"
)
```

#### WT AbemaR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds"
)
```

#### Mut AbemaR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds"
)
```

#### WT Vehicle

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds"
)
```

#### Mut Vehicle

```{r}
quasipoisson_glm_de_across_clusters_within_sample_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds"
)
```

### Calculate quasi-Poisson Wald statistics comparing clusters within a single sample within a cell cycle phase.

```{r}
#' Calculate the Wald statistic in a quasiPoisson glm that compares two 
#' clusters of cells within the same cell cycle phase within a single sample.
#'
#' Uses the sample-specific, per-gene estimated dispersions calculated above
#'  as part of the standard error of the tested coefficient in the QP glm.
#'
#' @param infile_rates Path to rds of estimated genewise rates per cluster 
#' within cell cycle phase for a single sample.
#' @param infile_dispersions Path to rds of estimated genewise dispersion per 
#' sample.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns Data frame of Wald statistics comparing clusters within each cell 
#' cycle phase, (cluster x cell cycle phase) specific expression rates, and 
#' sample specific dispersion estimates.
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf <- function(
  infile_rates,
  infile_dispersions = "data/processed/02-scrnaseq/genewise-overdispersions_simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_rates))
  rates <- readRDS(infile_rates)
  k = as.numeric(gsub("^genewise-rates-within_kmeans([0-9]+).*", "\\1", prefix))

  disps <- readRDS(infile_dispersions) |>
    dplyr::filter(orig.ident == unique(rates$orig.ident))

  rates_wide <- rates |>
    tidyr::pivot_wider(
      id_cols = gene,
      names_from = cluster_phase,
      values_from = c("lam", "N_reads", "N_cells")
    ) |>
    dplyr::left_join(disps, by = "gene")

  purrr::map(c("G1", "G2M", "S"), \(phase) {
    tidyr::expand_grid(x = 1:k, y = 1:k) |>
      dplyr::filter(x != y) |>
      purrr::pwalk(\(x, y) {
        if (
          !(sprintf("lam_%s:%s", x, phase) %in% colnames(rates_wide)) |
            !(sprintf("lam_%s:%s", y, phase) %in% colnames(rates_wide))
        ) {
          # Not enough cells,
          print(sprintf(
            "Not enough cells for either %s:%s or %s:%s.",
            x,
            phase,
            y,
            phase
          ))
          return()
        } else {
          rates_wide <<- rates_wide |> # global assignment
            dplyr::mutate(
              !!sym(sprintf(
                "waldstat_cluster%s-%s_minus_cluster%s-%s",
                x,
                phase,
                y,
                phase
              )) := ((log(!!sym(sprintf("lam_%s:%s", x, phase))) -
                log(!!sym(sprintf("lam_%s:%s", y, phase)))) /
                sqrt(
                  (pmax(1, phi) /
                    (!!sym(sprintf("lam_%s:%s", x, phase)) *
                      !!sym(sprintf("N_reads_%s:%s", x, phase)))) +
                    (pmax(1, phi) /
                      (!!sym(sprintf("lam_%s:%s", y, phase)) *
                        !!sym(sprintf("N_reads_%s:%s", y, phase))))
                ))
            )
        }
      })
  })

  (plt_hist <- rates_wide |>
    dplyr::select(gene, starts_with("waldstat")) |>
    tidyr::pivot_longer(
      cols = !gene,
      names_to = "comparison",
      values_to = "waldstat"
    ) |>
    dplyr::mutate(
      comparison_title = gsub(
        "minus_",
        "minus\n",
        gsub("waldstat_", "", comparison)
      )
    ) |>
    ggplot(aes(x = pmax(pmin(waldstat, 10), -10))) +
    geom_histogram() +
    labs(title = sprintf("%s Wald statistics", unique(rates_wide$orig.ident))) +
    theme_bw() +
    facet_wrap(~comparison_title))
  ggplot2::ggsave(
    file.path(
      odir,
      sprintf(
        "waldstat-ranked-genes-across-cluster-within-sample-per-phase-histogram_%s.png",
        prefix
      )
    ),
    plt_hist,
    width = 8,
    height = 8
  )

  saveRDS(
    rates_wide,
    file.path(
      ddir,
      sprintf(
        "waldstat-ranked-genes-across-cluster_within-sample-per-phase_%s.rds",
        prefix
      )
    )
  )
  return(rates_wide)
}
```

#### Parental

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds"
)
```

#### WT PalboR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds"
)
```

#### Mut PalboR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds"
)
```

#### WT PalboR Clone

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds"
)
```

#### WT AbemaR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds"
)
```

#### Mut AbemaR

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds"
)
```

#### WT Vehicle

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds"
)
```

#### Mut Vehicle

```{r}
quasipoisson_glm_de_across_clusters_within_sample_per_phase_simpleaf(
  infile_rates = "data/processed/02-scrnaseq/genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds"
)
```


### Calculate cameraPR gene set testing across two samples within a cell cycle phase.

```{r}
#' Run limma::cameraPR gene set testing for Hallmark gene sets for contrasts 
#' of interest within each cell cycle phase on gene lists ranked by 
#' quasi-Poisson Wald statistics.
#'
#' @param infile_waldstats Data frame of Wald statistics comparing samples 
#' within a cell cycle phase and (sample and cell cycle phase) specific 
#' expression rates and sample specific dispersion estimates.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
run_cameraPR_across_sample_within_phase_1v1_contrasts_quasipoisson_glm_simpleaf <- function(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-sample_within-phase_genewise-rates-within-phase-per-sample_simpleaf-quant-dds-allsamples_filterqc_glmpca-umap-kmeans.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  set.seed(1039101)
  prefix <- gsub(".rds", "", basename(infile_waldstats))

  ws <- readRDS(infile_waldstats)
  hk <- msigdbr::msigdbr(species = "human", category = "H") # Hallmark gene collection
  phk = split(x = hk$human_gene_symbol, f = hk$gs_name)

  (contrastdf <- rbind(
    c("Mut_AbemaR", "Mut_Vehicle"),
    c("WT_AbemaR", "WT_Vehicle"),
    c("Mut_PalboR", "Mut_Vehicle"),
    c("WT_PalboR", "WT_Vehicle"),

    c("Mut_Vehicle", "WT_Vehicle"),
    c("Mut_PalboR", "WT_PalboR"),
    c("Mut_AbemaR", "WT_AbemaR")
  ) |>
    as_tibble())
  names(contrastdf) <- c("x", "y")

  rbind(
    contrastdf |> dplyr::mutate(x = str_c(x, "_S"), y = str_c(y, "_S")),
    contrastdf |> dplyr::mutate(x = str_c(x, "_G1"), y = str_c(y, "_G1")),
    contrastdf |> dplyr::mutate(x = str_c(x, "_G2M"), y = str_c(y, "_G2M"))
  ) |>
    purrr::pmap(\(x, y) {
      res <- ws |>
        dplyr::filter(!is.na(!!sym(sprintf("waldstat_%s_minus_%s", x, y)))) |>
        dplyr::arrange(dplyr::desc(
          !!sym(sprintf("waldstat_%s_minus_%s", x, y))
        ))
      indexbp = limma::ids2indices(phk, res$gene) # order same as res
      res_camerapr <- limma::cameraPR(
        res |> dplyr::pull(!!sym(sprintf("waldstat_%s_minus_%s", x, y))),
        indexbp
      )

      res_camerapr <- res_camerapr |>
        as.data.frame() |>
        tibble::rownames_to_column("gene_set") |>
        tibble::column_to_rownames("gene_set")

      saveRDS(
        res_camerapr,
        file.path(
          ddir,
          sprintf(
            "cameraPR-by-quasipoisson-glm-waldstat-%s-minus-%s_%s.rds",
            x,
            y,
            prefix
          )
        )
      )
      return(res_camerapr)
    })
}
```

```{r}
#| output: false
run_cameraPR_across_sample_within_phase_1v1_contrasts_quasipoisson_glm_simpleaf()
```

### Plot cameraPR gene set testing results across samples within cell cycle phase.

```{r}
#' Plot a bubble plot of the gene set testing results for the across sample
#' (one vs. one) comparisons within each separate cell cycle phase.
#'
#' @param infiles Vector of filepaths of gene set testing output RDS files from
#' [run_cameraPR_across_sample_within_phase_1v1_contrasts_quasipoisson_glm_simpleaf()].
#' @param thresh FDR threshold per columns of plot.
#' @param title Plot title.
#' @param suffix Plot suffix to go in output filepath.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns ggplot2 object of plot.
plot_cameraPR_heatmap_quasipoisson_glm_across_sample_within_phase_one_over_one_simpleaf <- function(
  infiles,
  thresh = 0.10,
  title,
  suffix,
  width = 9,
  height = 9,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dat <- infiles |>
    purrr::map_dfr(\(x) {
      readRDS(x) |>
        dplyr::mutate(
          sid = gsub(
            "data/processed/02-scrnaseq/cameraPR-by-quasipoisson-glm-waldstat-(.*)_waldstat.*rds",
            "\\1",
            x
          )
        ) |> # extract sample names
        tibble::rownames_to_column("gene_set") |>
        dplyr::mutate(gs_name = gsub("HALLMARK_", "", gene_set)) |>
        dplyr::mutate(Phase = gsub(".*_([^_]*)", "\\1", sid))
    }) |>
    dplyr::relocate(sid) |>
    dplyr::mutate(sid = gsub("-minus-", "_over_", sid)) |> # trims the between-cluster _ over rest names
    dplyr::mutate(
      sid = forcats::fct_relevel(
        sid,

        "Mut_AbemaR_S_over_Mut_PalboR_S",
        "Mut_AbemaR_S_over_Mut_Vehicle_S",
        "Mut_AbemaR_S_over_WT_AbemaR_S",
        "WT_AbemaR_S_over_WT_PalboR_S",
        "WT_AbemaR_S_over_WT_Vehicle_S",
        "Mut_PalboR_S_over_Mut_Vehicle_S",
        "Mut_PalboR_S_over_WT_PalboR_S",
        "WT_PalboR_Clone_S_over_WT_PalboR_S",
        "WT_PalboR_S_over_WT_Vehicle_S",
        "Mut_Vehicle_S_over_WT_Vehicle_S",
        "WT_Vehicle_S_over_Parental_S",

        "Mut_AbemaR_S_over_Parental_S",
        "WT_AbemaR_S_over_Parental_S",
        "Mut_PalboR_S_over_Parental_S",
        "WT_PalboR_S_over_Parental_S",

        "Mut_AbemaR_G1_over_Mut_PalboR_G1",
        "Mut_AbemaR_G1_over_Mut_Vehicle_G1",
        "Mut_AbemaR_G1_over_WT_AbemaR_G1",
        "WT_AbemaR_G1_over_WT_PalboR_G1",
        "WT_AbemaR_G1_over_WT_Vehicle_G1",
        "Mut_PalboR_G1_over_Mut_Vehicle_G1",
        "Mut_PalboR_G1_over_WT_PalboR_G1",
        "WT_PalboR_Clone_G1_over_WT_PalboR_G1",
        "WT_PalboR_G1_over_WT_Vehicle_G1",
        "Mut_Vehicle_G1_over_WT_Vehicle_G1",
        "WT_Vehicle_G1_over_Parental_G1",

        "Mut_AbemaR_G1_over_Parental_G1",
        "WT_AbemaR_G1_over_Parental_G1",
        "Mut_PalboR_G1_over_Parental_G1",
        "WT_PalboR_G1_over_Parental_G1",

        "Mut_AbemaR_G2M_over_Mut_PalboR_G2M",
        "Mut_AbemaR_G2M_over_Mut_Vehicle_G2M",
        "Mut_AbemaR_G2M_over_WT_AbemaR_G2M",
        "WT_AbemaR_G2M_over_WT_PalboR_G2M",
        "WT_AbemaR_G2M_over_WT_Vehicle_G2M",
        "Mut_PalboR_G2M_over_Mut_Vehicle_G2M",
        "Mut_PalboR_G2M_over_WT_PalboR_G2M",
        "WT_PalboR_Clone_G2M_over_WT_PalboR_G2M",
        "WT_PalboR_G2M_over_WT_Vehicle_G2M",
        "Mut_Vehicle_G2M_over_WT_Vehicle_G2M",
        "WT_Vehicle_G2M_over_Parental_G2M",

        "Mut_AbemaR_G2M_over_Parental_G2M",
        "WT_AbemaR_G2M_over_Parental_G2M",
        "Mut_PalboR_G2M_over_Parental_G2M",
        "WT_PalboR_G2M_over_Parental_G2M"
      )
    ) |>
    dplyr::filter(FDR < thresh) |>
    dplyr::mutate(Direction = factor(Direction, levels = c("Up", "Down")))
  if (nrow(dat) == 0) {
    print("No gene sets passing FDR threshold.")
  }

  gs_rnk <- dat |>
    dplyr::ungroup() |>
    dplyr::group_by(gs_name) |>
    dplyr::mutate(rnk = sum(Direction == "Up") - sum(Direction == "Down")) |>
    dplyr::arrange(rnk, gs_name) |>
    dplyr::distinct(gs_name) |>
    dplyr::pull()

  (plt <- dat |>
    dplyr::mutate(Phase = factor(Phase, levels = c("S", "G1", "G2M"))) |>
    dplyr::mutate(gs_name = forcats::fct_relevel(factor(gs_name), gs_rnk)) |>
    ggplot2::ggplot(aes(
      x = sid,
      y = gs_name,
      color = Direction,
      label = NGenes,
      size = FDR
    )) +
    ggplot2::geom_point() +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
    ) +
    ggplot2::scale_color_manual(values = c(col_high, col_low), drop = FALSE) +
    ggplot2::scale_size(trans = "reverse", range = c(.1, 3)) +
    ggplot2::guides(
      colour = ggplot2::guide_legend(order = 1),
      size = ggplot2::guide_legend(order = 2)
    ) +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      title = title,
      subtitle = sprintf("%s, %s%% FDR per column", "condition", thresh * 100)
    ) +
    ggplot2::facet_wrap(~Phase, nrow = 1, scales = "free_x"))

  ggsave(
    file.path(
      odir,
      sprintf("cameraPR-by-QPglm-heatmap_%.2f_%s.pdf", thresh, suffix)
    ),
    plt,
    width = width,
    height = height
  )
  ggsave(
    file.path(
      odir,
      sprintf("cameraPR-by-QPglm-heatmap_%.2f_%s.png", thresh, suffix)
    ),
    plt,
    width = width,
    height = height
  )

  return(plt)
}
```

#### Treatment over Vehicle (Figure 6D).
```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_heatmap_quasipoisson_glm_across_sample_within_phase_one_over_one_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR-by-quasipoisson.*(AbemaR|PalboR).*minus.*Vehicle.*phase",
    full.names = TRUE
  ),
  title = "AbemaR PalboR over Vehicle, by phase",
  width = 6,
  suffix = "QPglm-minmaxzero-delta-waldstat-across-sample-within-phase-AbemaR-PalboR-over-Vehicle"
)
```

#### Mutant over WT (Supplementary Figure S7C).

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_heatmap_quasipoisson_glm_across_sample_within_phase_one_over_one_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR-by-quasipoisson.*Mut_(AbemaR|PalboR|Vehicle).*minus.*WT.*phase",
    full.names = TRUE
  ),
  title = "Mut over WT (AbemaR, PalboR, Vehicle), by phase",
  width = 6,
  suffix = "QPglm-minmaxzero-delta-waldstat-across-sample-within-phase-Mut-over-WT-AbemaR-PalboR-Vehicle"
)
```

### Calculate cameraPR gene set testing across all clusters (1 vs. rest) per sample.

```{r}
#' Run limma::cameraPR gene set testing for Hallmark gene sets for contrasts of
#' (one vs. rest) on gene lists ranked using quasipoisson Wald statistic between kmeans
#' clusters within a sample using the following criteria:
#'
#' for a set of (one vs. rest) comparisons, if the QPglm Wald statistics:
#' are all positive, use the minimum delta as a marker ranking,
#' are all negative, use the maximum delta as a marker ranking,
#' are mixed positive and negative, use zero as a marker ranking (noninformative), and
#' is NA (phi, rates were zero), we set this to zero, as this indicates no 
#' difference between the clusters (zero is called genewide in this case)
#'
#' @param infile_rds Data frame of Wald statistics comparing clusters, cluster 
#' specific expression rates for a single sample, and sample-specific 
#' dispersion estimates.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns List of cameraPR outputs for each tested cluster.
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf <- function(
  infile_waldstats,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_waldstats))
  k = as.numeric(gsub(
    "^waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans([0-9]+).*",
    "\\1",
    prefix
  ))
  prefix <- gsub(
    "waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_",
    "",
    prefix
  ) # shorten the filename

  ws <- readRDS(infile_waldstats)
  hk <- msigdbr::msigdbr(species = "human", category = "H") # Hallmark gene collection
  phk = split(x = hk$human_gene_symbol, f = hk$gs_name)

  res_list <- 1:k |>
    purrr::map(\(i) {
      res <- ws |>
        dplyr::mutate(
          !!sym(sprintf(
            "minmaxzero_delta_waldstat_cluster%s_minus_other",
            i
          )) := dplyr::case_when(
            MatrixGenerics::rowMins(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
              "waldstat_cluster%s_minus",
              i
            ))))) >
              0 ~ MatrixGenerics::rowMins(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
              "waldstat_cluster%s_minus",
              i
            ))))),
            MatrixGenerics::rowMaxs(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
              "waldstat_cluster%s_minus",
              i
            ))))) <
              0 ~ MatrixGenerics::rowMaxs(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
              "waldstat_cluster%s_minus",
              i
            ))))),
            .default = 0
          )
        ) |>
        dplyr::filter(
          !is.na(
            !!sym(sprintf("minmaxzero_delta_waldstat_cluster%i_minus_other", i))
          )
        ) |>
        dplyr::arrange(dplyr::desc(
          !!sym(sprintf("minmaxzero_delta_waldstat_cluster%i_minus_other", i))
        ))

      # Check that we calculated this delta correctly
      stopifnot(all(
        res |>
          dplyr::select(
            dplyr::starts_with(sprintf("waldstat_cluster%s_minus", i)),
            !!sym(sprintf("minmaxzero_delta_waldstat_cluster%s_minus_other", i))
          ) |>
          dplyr::filter(
            !!sym(sprintf(
              "minmaxzero_delta_waldstat_cluster%s_minus_other",
              i
            )) <
              0
          ) <
          0
      ))
      stopifnot(all(
        res |>
          dplyr::select(
            dplyr::starts_with(sprintf("waldstat_cluster%s_minus", i)),
            !!sym(sprintf("minmaxzero_delta_waldstat_cluster%s_minus_other", i))
          ) |>
          dplyr::filter(
            !!sym(sprintf(
              "minmaxzero_delta_waldstat_cluster%s_minus_other",
              i
            )) >
              0
          ) >
          0
      ))

      indexbp = limma::ids2indices(phk, res$gene) # order same as res
      res_camerapr <- limma::cameraPR(
        res |>
          dplyr::pull(
            !!sym(sprintf("minmaxzero_delta_waldstat_cluster%s_minus_other", i))
          ),
        indexbp
      )
      res_camerapr$N_cells <- res |>
        dplyr::pull(!!sym(sprintf("N_cells_%i", i))) |>
        unique() # For nice labels when plotting in heatmap

      res_camerapr <- res_camerapr |>
        as.data.frame() |>
        tibble::rownames_to_column("gene_set")

      saveRDS(
        res_camerapr,
        file.path(
          ddir,
          sprintf(
            "cameraPR-by-QPglm-minmaxzero-delta-waldstat-cluster%i-minus-other_%s.rds",
            i,
            prefix
          )
        )
      )
      saveRDS(
        res,
        file.path(
          ddir,
          sprintf(
            "QPglm-minmaxzero-delta-waldstat-cluster%i-minus-other_%s.rds",
            i,
            prefix
          )
        )
      )
      return(res_camerapr)
    })
  return(res_list)
}
```

### Plot cameraPR gene set testing across clusters per sample.

```{r}
#' Plot a bubble plot of the gene set testing results for the across clusters
#' within a sample (one vs. rest) comparisons.
#'
#' @param infiles Vector of filepaths of gene set testing output RDS files from
#' [run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf()].
#' @param thresh FDR threshold per columns of plot.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns ggplot2 object of plot.
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf <- function(
  infiles,
  thresh = 0.10,
  width = 6,
  height = 8,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dat <- infiles |>
    purrr::map_dfr(\(x) {
      readRDS(x) |>
        dplyr::mutate(
          sid = gsub(
            "data/processed/02-scrnaseq/cameraPR-by-QPglm-minmaxzero-delta-waldstat-(.*)-minus.*rds",
            "\\1",
            x
          )
        ) |> # extract sample names
        dplyr::mutate(gs_name = gsub("HALLMARK_", "", gene_set))
    }) |>
    dplyr::relocate(sid) |>
    dplyr::mutate(sid = gsub("-minus-", "_over_", sid)) |> # trims the between-cluster _ over rest names
    dplyr::filter(FDR < thresh) |>
    dplyr::mutate(Direction = factor(Direction, levels = c("Up", "Down")))
  if (nrow(dat) == 0) {
    print("No gene sets passing FDR threshold.")
  }

  gs_rnk <- dat |>
    dplyr::ungroup() |>
    dplyr::group_by(gs_name) |>
    dplyr::mutate(rnk = sum(Direction == "Up") - sum(Direction == "Down")) |>
    dplyr::arrange(rnk, gs_name) |>
    dplyr::distinct(gs_name) |>
    dplyr::pull()

  clustername <- gsub(
    ".*other_(.*)_simpleaf-quant-dds-allsamples_filterqc_.*",
    "\\1",
    basename(infiles[[1]])
  )
  samplename <- gsub(
    ".*other_(.*)_simpleaf-quant-dds-allsamples_filterqc_(.*)_glmpca-umap-kmeans.rds",
    "\\2",
    basename(infiles[[1]])
  )
  title = sprintf(
    "%s\n%s\ncluster-over-rest gene set testing",
    samplename,
    clustername
  )

  (plt <- dat |>
    dplyr::mutate(
      sid_name = sprintf("Cluster %s (N=%i)", gsub("cluster", "", sid), N_cells)
    ) |>
    ggplot2::ggplot(aes(
      x = sid_name,
      y = gs_name,
      color = Direction,
      label = NGenes,
      size = FDR
    )) +
    ggplot2::geom_point() +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
    ) +
    ggplot2::scale_y_discrete(limits = rev) +
    ggplot2::scale_color_manual(values = c(col_high, col_low), drop = FALSE) +
    ggplot2::scale_size(trans = "reverse", range = c(.1, 3)) +
    ggplot2::guides(
      colour = ggplot2::guide_legend(order = 1),
      size = ggplot2::guide_legend(order = 2)
    ) +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      title = title,
      subtitle = sprintf("%s, %s%% FDR per column", "condition", thresh * 100)
    ))

  ggsave(
    file.path(
      odir,
      sprintf(
        "cameraPR-by-QPglm-heatmap_%.2f_%s_%s.pdf",
        thresh,
        samplename,
        clustername
      )
    ),
    plt,
    width = width,
    height = height
  )
  ggsave(
    file.path(
      odir,
      sprintf(
        "cameraPR-by-QPglm-heatmap_%.2f_%s_%s.png",
        thresh,
        samplename,
        clustername
      )
    ),
    plt,
    width = width,
    height = height
  )

  return(plt)
}
```

#### Parental (Figure 6E).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2_simpleaf.*Parental_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```


#### WT PalboR (Figure 6F).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans3_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3_simpleaf.*WT_PalboR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### Mut PalboR (Figure 6G).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans3_simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3_simpleaf.*Mut_PalboR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT PalboR Clone (Figure 6H).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans3_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3_simpleaf.*WT_PalboR_Clone_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT AbemaR (Figure 6I).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2_simpleaf.*WT_AbemaR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### Mut AbemaR (Figure 6J).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2_simpleaf.*Mut_AbemaR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT Vehicle (Supplementary Figure S7D).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2_simpleaf.*WT_Vehicle_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### Mut Vehicle (Supplementary Figure S7E).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample_genewise-rates-within_kmeans2_simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2_simpleaf.*Mut_Vehicle_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

### Calculate cameraPR gene set testing across clusters within cell cycle phase per sample

```{r}
#' Run limma::cameraPR gene set testing for Hallmark gene sets for contrasts of
#' interest on gene lists ranked by quasipoisson Wald statistic between kmeans
#' clusters within a sample within separate called cell cycle phases using the
#' following criteria:
#'
#' for a certain set of comparisons, if the QPglm Wald statistics:
#' are all positive, uses the minimum delta as a marker ranking,
#' are all negative, uses the maximum delta as a marker ranking,
#' are mixed positive and negative, use zero as a marker ranking (noninformative),
#' is NA (phi, rates were zero), we set this to zero, as this indicates no 
#' difference between the clusters (zero is called genewide in this case)
#'
#' @param infile_rds Path of rds with dataframe of Seurat::FoldChanges.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns List of cameraPR outputs for each tested cluster.
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf <- function(
  infile_waldstats,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_waldstats))
  k = as.numeric(gsub(
    "^waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans([0-9]+).*",
    "\\1",
    prefix
  ))
  prefix <- gsub(
    "waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_",
    "",
    prefix
  ) # just to shorten the filename

  ws <- readRDS(infile_waldstats)
  hk <- msigdbr::msigdbr(species = "human", category = "H") # Hallmark gene collection
  phk = split(x = hk$human_gene_symbol, f = hk$gs_name)

  res_list <- purrr::map(c("G1", "G2M", "S"), \(phase) {
    res_list <- 1:k |>
      purrr::map(\(i) {
        if (
          !any(grepl(
            sprintf("waldstat_cluster%s-%s_minus", i, phase),
            colnames(ws)
          ))
        ) {
          # Not enough cells to do that test
          print(sprintf(
            "Not able to test waldstat_cluster%s-%s_minus",
            i,
            phase
          ))
          return()
        } else {
          res <- ws |>
            dplyr::mutate(
              !!sym(sprintf(
                "minmaxzero_delta_waldstat_cluster%s-%s_minus_other",
                i,
                phase
              )) := dplyr::case_when(
                MatrixGenerics::rowMins(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
                  "waldstat_cluster%s-%s_minus",
                  i,
                  phase
                ))))) >
                  0 ~ MatrixGenerics::rowMins(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
                  "waldstat_cluster%s-%s_minus",
                  i,
                  phase
                ))))),
                MatrixGenerics::rowMaxs(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
                  "waldstat_cluster%s-%s_minus",
                  i,
                  phase
                ))))) <
                  0 ~ MatrixGenerics::rowMaxs(as.matrix(dplyr::pick(dplyr::starts_with(sprintf(
                  "waldstat_cluster%s-%s_minus",
                  i,
                  phase
                ))))),
                .default = 0
              )
            ) |>
            dplyr::filter(
              !is.na(
                !!sym(sprintf(
                  "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                  i,
                  phase
                ))
              )
            ) |>
            dplyr::arrange(dplyr::desc(
              !!sym(sprintf(
                "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                i,
                phase
              ))
            ))

          # Check that we calculated this delta correctly
          stopifnot(all(
            res |>
              dplyr::select(
                dplyr::starts_with(sprintf(
                  "waldstat_cluster%s-%s_minus",
                  i,
                  phase
                )),
                !!sym(sprintf(
                  "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                  i,
                  phase
                ))
              ) |>
              dplyr::filter(
                !!sym(sprintf(
                  "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                  i,
                  phase
                )) <
                  0
              ) <
              0
          ))
          stopifnot(all(
            res |>
              dplyr::select(
                dplyr::starts_with(sprintf(
                  "waldstat_cluster%s-%s_minus",
                  i,
                  phase
                )),
                !!sym(sprintf(
                  "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                  i,
                  phase
                ))
              ) |>
              dplyr::filter(
                !!sym(sprintf(
                  "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                  i,
                  phase
                )) >
                  0
              ) >
              0
          ))

          indexbp = limma::ids2indices(phk, res$gene) # order same as res
          res_camerapr <- limma::cameraPR(
            res |>
              dplyr::pull(
                !!sym(sprintf(
                  "minmaxzero_delta_waldstat_cluster%i-%s_minus_other",
                  i,
                  phase
                ))
              ),
            indexbp
          )
          res_camerapr$N_cells <- res |>
            dplyr::pull(!!sym(sprintf("N_cells_%i:%s", i, phase))) |>
            unique() # For nice labels when plotting in heatmap

          res_camerapr <- res_camerapr |>
            as.data.frame() |>
            tibble::rownames_to_column("gene_set")

          saveRDS(
            res_camerapr,
            file.path(
              ddir,
              sprintf(
                "cameraPR-by-QPglm-minmaxzero-delta-waldstat-cluster%i-%s-minus-other_%s.rds",
                i,
                phase,
                prefix
              )
            )
          )
          saveRDS(
            res,
            file.path(
              ddir,
              sprintf(
                "QPglm-minmaxzero-delta-waldstat-cluster%i-%s-minus-other_%s.rds",
                i,
                phase,
                prefix
              )
            )
          )
          return(res_camerapr)
        }
      })
  })

  return(res_list)
}
```

### Plot cameraPR gene set testing across clusters within cell cycle phase per sample

```{r}
#' Plot a bubble plot of the gene set testing results for the across clusters
#' (one vs. rest) within a cell cycle phase within sample comparisons.
#'
#' @param infiles Vector of filepaths of gene set testing output RDS files from
#' [run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf()].
#' @param thresh FDR threshold per columns of plot.
#' @param title Plot title.
#' @param suffix Plot suffix to go in output filepath.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
#'
#' @returns ggplot2 object of bubble plot.
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf <- function(
  infiles,
  thresh = 0.10,
  facet_by = NULL,
  width = 6,
  height = 8,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  dat <- infiles |>
    purrr::map_dfr(\(x) {
      readRDS(x) |>
        dplyr::mutate(
          sid = gsub(
            "data/processed/02-scrnaseq/cameraPR-by-QPglm-minmaxzero-delta-waldstat-(.*)-minus.*rds",
            "\\1",
            x
          )
        ) |> # extract sample names
        dplyr::mutate(gs_name = gsub("HALLMARK_", "", gene_set)) |>
        tidyr::separate_wider_delim(
          sid,
          delim = "-",
          names = c("Cluster", "Phase"),
          cols_remove = FALSE
        )
    }) |>
    dplyr::relocate(sid) |>
    dplyr::mutate(sid = gsub("-minus-", "_over_", sid)) |> # trims the between-cluster _ over rest names
    dplyr::filter(FDR < thresh) |>
    dplyr::mutate(Direction = factor(Direction, levels = c("Up", "Down")))
  if (nrow(dat) == 0) {
    print("No gene sets passing FDR threshold.")
  }

  gs_rnk <- dat |>
    dplyr::ungroup() |>
    dplyr::group_by(gs_name) |>
    dplyr::mutate(rnk = sum(Direction == "Up") - sum(Direction == "Down")) |>
    dplyr::arrange(rnk, gs_name) |>
    dplyr::distinct(gs_name) |>
    dplyr::pull()

  clustername <- gsub(
    ".*other_(.*)_simpleaf-quant-dds-allsamples_filterqc_.*",
    "\\1",
    basename(infiles[[1]])
  )
  samplename <- gsub(
    ".*other_(.*)_simpleaf-quant-dds-allsamples_filterqc_(.*)_glmpca-umap-kmeans.rds",
    "\\2",
    basename(infiles[[1]])
  )
  title = sprintf(
    "%s\n%s\ncluster-over-rest gene set testing",
    samplename,
    clustername
  )

  (plt <- dat |>
    dplyr::mutate(Phase = factor(Phase, levels = c("S", "G1", "G2M"))) |>
    dplyr::mutate(
      sid_name = sprintf("Cluster %s (N=%i)", gsub("cluster", "", sid), N_cells)
    ) |>
    ggplot2::ggplot(aes(
      x = sid_name,
      y = gs_name,
      color = Direction,
      label = NGenes,
      size = FDR
    )) +
    ggplot2::geom_point() +
    ggplot2::theme_bw() +
    ggplot2::theme(
      axis.text.x = ggplot2::element_text(angle = 90, vjust = 0.5, hjust = 1)
    ) +
    ggplot2::scale_y_discrete(limits = rev) +
    ggplot2::scale_color_manual(values = c(col_high, col_low), drop = FALSE) +
    ggplot2::scale_size(trans = "reverse", range = c(.1, 3)) +
    ggplot2::guides(
      colour = ggplot2::guide_legend(order = 1),
      size = ggplot2::guide_legend(order = 2)
    ) +
    ggplot2::labs(
      x = NULL,
      y = NULL,
      title = title,
      subtitle = sprintf("%s, %s%% FDR per column", "condition", thresh * 100)
    ) +
    ggplot2::facet_wrap(~Phase, nrow = 1, scales = "free_x"))

  ggsave(
    file.path(
      odir,
      sprintf(
        "cameraPR-by-QPglm-heatmap_%.2f_%s_%s.pdf",
        thresh,
        samplename,
        clustername
      )
    ),
    plt,
    width = width,
    height = height
  )
  ggsave(
    file.path(
      odir,
      sprintf(
        "cameraPR-by-QPglm-heatmap_%.2f_%s_%s.png",
        thresh,
        samplename,
        clustername
      )
    ),
    plt,
    width = width,
    height = height
  )

  return(plt)
}
```

#### Parental (Supplementary Figure S8A).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Parental_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3.*within-phase.*Parental_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT Vehicle (Supplementary Figure S8B).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_Vehicle_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2.*within-phase.*WT_Vehicle_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### Mut Vehicle (Supplementary Figure S8C).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Mut_Vehicle_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2.*within-phase.*Mut_Vehicle_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT PalboR (Supplementary Figure S8D).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3.*within-phase.*WT_PalboR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### Mut PalboR (Supplementary Figure S8E).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Mut_PalboR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3.*within-phase.*Mut_PalboR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT PalboR Clone (Supplementary Figure S8F).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans3-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_PalboR_Clone_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans3.*within-phase.*WT_PalboR_Clone_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### WT AbemaR (Supplementary Figure S8G).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_WT_AbemaR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2.*within-phase.*WT_AbemaR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

#### Mut AbemaR (Supplementary Figure S8H).

```{r}
#| output: false
run_cameraPR_with_delta_1vrest_contrasts_across_clusters_within_sample_within_phase_quasipoisson_glm_simpleaf(
  infile_waldstats = "data/processed/02-scrnaseq/waldstat-ranked-genes-across-cluster_within-sample-per-phase_genewise-rates-within_kmeans2-within-phase_geq50cells_simpleaf-quant-dds-allsamples_filterqc_Mut_AbemaR_glmpca-umap-kmeans.rds"
)
```

```{r}
#| fig-width: 7
#| fig-height: 9
plot_cameraPR_with_delta_heatmap_quasipoisson_glm_across_clusters_within_sample_within_phase_simpleaf(
  infiles = list.files(
    "data/processed/02-scrnaseq",
    pattern = "cameraPR.*other_kmeans2.*within-phase.*Mut_AbemaR_glmpca-umap-kmeans\\.rds",
    full.names = TRUE
  )
)
```

## Inferred copy number analysis

Note, due to large memory requirements, we recommend running this step on a high-performance computing environment and include sample SLURM submission code and R code and show pre-computed output plots below.

### Create gene order table from GTF used for alignment

```{r}
#' Convert a gtf to a format compatible with Broad's infercnv.
#'
#' Takes the min and max coordinates of each gene in the gtf that was used to
#' run nf-core scrnaseq simpleaf.
#'
#' Removes genes that occur in GTF across multiple chromosomes.
#'
#' @param infile_gtf Path to GTF file used to align scRNAseq samples.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
make_gene_order_file_simpleaf <- function(
  infile_gtf = "/rafalab/drusso/genomes/refgenie_0.12.1_refgenconf_0.12.2_hg38_tmp/gencode_gtf/hg38.gtf.gz",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  d <- data.table::fread(
    infile_gtf,
    skip = 5,
    col.names = c(
      "seqname",
      "source",
      "feature",
      "start",
      "end ",
      "score",
      "strand",
      "frame",
      "attribute"
    )
  )
  out <- d |>
    dplyr::mutate(
      gid = gsub("\"", "", gsub(".*gene_name\ ([^;]*).*", "\\1", attribute))
    ) |> # get gene name
    dplyr::rename(end = `end `) |>
    dplyr::select(gid, seqname, start, end) |>
    dplyr::group_by(gid, seqname) |>
    dplyr::mutate(start = min(start), end = max(end)) |> # span entire gene start to end
    dplyr::ungroup() |>
    dplyr::select(gid, seqname, start, end) |>
    dplyr::distinct()

  # A few genes appear on multiple chromosomes -- filter these out
  multi_genes <- out |>
    dplyr::count(gid) |>
    dplyr::filter(n > 1) |>
    dplyr::pull(gid)
  out <- out |>
    dplyr::filter(!(gid %in% multi_genes))

  readr::write_tsv(out, file.path(ddir, sprintf("hg38_genes_simpleaf.txt")))
  return(out)
}
```

```{r}
make_gene_order_file_simpleaf()
```

### Run inverCNV

#### Sample R script to run invercnf: R/10-sc-inferCNV.R
```{r}
#| eval: false

library(tidyverse)
library(Seurat)
library(infercnv)

#' Run Broad's infercnv on scRNAseq samples using the Parental sample as the
#' reference.
#'
#' May need to set options(bitmapType='cairo') to get plots to render,
#' plotting moved to plot_infercnv_simpleaf() function below.
#'
#' @param infile_seurat_rds Path to rds of Seurat object of all samples.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
run_infercnv_simpleaf <- function(
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc.rds",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  dds <- readRDS(infile_seurat_rds)
  counts <- LayerData(dds, assay = "RNA", layer = "counts")

  meta <- dds@meta.data

  # Gene order txt file created above.
  d <- readr::read_tsv("data/processed/02-scrnaseq/hg38_genes_simpleaf.txt")
  gcts <- d$seqname |> table() |> sort()
  names(gcts[gcts < 10]) # scaffold transcripts

  d <- d |>
    filter(!(seqname %in% c(names(gcts[gcts < 10])))) |>
    mutate(
      seqname = factor(seqname, levels = str_c("chr", c(1:23, "X", "Y", "M")))
    ) |>
    arrange(seqname, start)

  infercnv_obj = CreateInfercnvObject(
    raw_counts_matrix = counts,
    annotations_file = (meta |> select(orig.ident)),
    delim = "\t",
    ref_group_names = c("Parental"),
    gene_order_file = (d |> tibble::column_to_rownames("gid")),
    chr_exclude = c(names(gcts[gcts < 10]))
  )

  out_dir = tempfile()
  infercnv_obj_default = infercnv::run(
    infercnv_obj,
    cutoff = 0.1,
    out_dir = ddir,
    cluster_by_groups = TRUE,
    plot_steps = FALSE,
    denoise = TRUE,
    HMM = FALSE,
    no_plot = TRUE, # can be set to FALSE to generate plot in this step
    no_prelim_plot = TRUE,
    png_res = 60
  )
}

#' Plot inferCNV heatmap.
#'
#' @param infile_infercnv Path to final output file from infercnv.
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_infercnv_simpleaf <- function(
  infile_infercnv = "data/processed/02-scrnaseq/run.final.infercnv_obj",
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  d <- readRDS(infile_infercnv)
  infercnv::plot_cnv(
    d,
    out_dir = odir,
    obs_title = "Observations (Cells)",
    ref_title = "References (Cells)",
    cluster_by_groups = TRUE,
    x.center = 1,
    x.range = "auto",
    hclust_method = 'ward.D',
    color_safe_pal = FALSE,
    output_filename = "infercnv-simpleaf",
    output_format = "pdf",
    png_res = 300,
    dynamic_resize = 0
  )
}

#' Plot a ridgeplot of inferred copynumber for each gene in a given vector on
#' the inferCNV run on the simpleaf processed scRNAseq samples.
#'
#' @param infile_infercnv Path to rds of infercnv final object results.
#' @param infile_seurat_rds Path to rds of Seurat object used to run infercnv.
#' @param glist Vector of genes to plot.
#' @param width Plot width (in).
#' @param height Plot height (in).
#' @param ddir Processed data directory.
#' @param odir Processed output directory.
plot_ridgeplot_gene_infercnv_simpleaf <- function(
  infile_infercnv = "data/processed/02-scrnaseq/run.final.infercnv_obj",
  infile_seurat_rds = "data/processed/02-scrnaseq/simpleaf-quant-dds-allsamples_filterqc.rds",
  glist = c("RB1"),
  width = 6,
  height = 5,
  ddir = "data/processed/02-scrnaseq",
  odir = "data/processed/02-scrnaseq"
) {
  d <- readRDS(infile_infercnv)

  cnv_data <- d@expr.data |>
    t() |>
    as.data.frame() |>
    tibble::rownames_to_column("bid") |>
    as_tibble()

  prefix <- gsub(".rds", "", basename(infile_seurat_rds))
  dds <- readRDS(infile_seurat_rds)
  meta <- dds@meta.data |>
    tibble::rownames_to_column("bid")

  for (gid in glist) {
    plt <- cnv_data |>
      select(bid, !!sym(gid)) |>
      inner_join(meta, by = "bid") |>
      ggplot(aes(y = orig.ident, x = !!sym(gid), fill = orig.ident)) +
      ggridges::geom_density_ridges(scale = .9, alpha = 1) +
      labs(title = gid, y = NULL, x = "Inferred copy number") +
      theme_bw() +
      scale_y_discrete(limits = rev) +
      scale_fill_manual(name = "Condition", values = pal_orig.ident)
    ggsave(
      file.path(odir, sprintf("infercnv-simpleaf-%s_%s.pdf", gid, prefix)),
      plt,
      width = width,
      height = height
    )
    ggsave(
      file.path(odir, sprintf("infercnv-simpleaf-%s_%s.png", gid, prefix)),
      plt,
      width = width,
      height = height
    )
    print(plt)
  }
}

run_infercnv_simpleaf()
plot_infercnv_simpleaf()
plot_ridgeplot_gene_infercnv_simpleaf()
```

#### Sample SLURM submission script to run inferCNV.

Note, JAGS needs to be installed to run the inferCNV R package.

```{bash}
#| eval: false

#!/bin/bash
#SBATCH --job-name=10-sc-inferCNV
#SBATCH --time=12:00:00
#SBATCH --mem=124G
#SBATCH --ntasks=1
#SBATCH --ntasks-per-node=1
#SBATCH --error=logs/10-sc-inferCNV_%A.err
#SBATCH --output=logs/10-sc-inferCNV_%A.out
#SBATCH --open-mode=append

module load jags/4.3.1
module load R/4.3.2

Rscript R/10-sc-inferCNV.R
```

### Plot heatmap of inferCNV results (Supplementary Figure S5B).

![](data/raw/02-scrnaseq/infercnv-simpleaf.pdf)

### Plot inferCNV density plot of RB1 (Figure 4D).

![](data/raw/02-scrnaseq/infercnv-simpleaf-RB1_simpleaf-quant-dds-allsamples_filterqc.pdf)

## Print session info.

```{r}
sessionInfo()
```
